### 2019-06-14 rre_manuscript/paper.tex
initial commit of MS

\documentclass[oupdraft]{bio}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
\usepackage{url}

% Add history information for the article if required
% \history{Received August 1, 2010;
% revised October 1, 2010;
% accepted for publication November 1, 2010}

\usepackage{amsmath}
\usepackage{xparse}
\usepackage{mathtools}
\usepackage[table]{xcolor}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ev}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\var}[1]{\text{Var}\left[#1\right]}

% \newcommand{\kmax}{k_{\text{max}}}
% \newcommand{\kmax}[1]{k_{\text{max}(#1)}}
% \newcommand{\kmax}[jj]{k_{\text{max}#1}}

\NewDocumentCommand{\kmax}{o}{%
    {%
        \IfNoValueTF {#1}{
          k_{\text{max}}
        }{
          k_{\text{max}(#1)}
        }%
    }%
}
\newcommand{\lambdagrid}{\lambda^{\text{grid}}}

\begin{document}

% Title of paper
\title{Tuning parameter selection for a penalized maximum likelihood estimator of species richness}

% List of authors, with corresponding author marked by asterisk
\author{ALEX PAYNTER$^\ast$, AMY WILLIS\\[4pt]
% Author addresses
\textit{Department of Biostatistics,
University of Washington,
Health Sciences Building, Box 357232,
1705 NE Pacific St., Seattle, WA 98195}
\\[2pt]
% E-mail address for correspondence
{payntera@uw.edu adwillis@uw.edu}}

% Running headers of paper:
\markboth%
% First field is the short list of authors
{A. Paynter and others}
% Second field is the short title of the paper
{Tuning parameter selection in species richness estimation}

\maketitle

% Add a footnote for the corresponding author if one has been
% identified in the author list
\footnotetext{To whom correspondence should be addressed.}

\begin{abstract}
{Our goal is estimating the true number of classes in a population.   We focus on the scenario where multiple frequency count tables have been collected from the same population.  In this setting we demonstrate the efficacy of a previously published penalized maximum likelihood method.  Four novel methods to tune the requisite penalization parameter are proposed.  The performance of all proposed tuning methods is compared in simulations.}
\end{abstract}

\section{Introduction}
\label{sec:introduction}

In the \textit{species problem} the goal is estimating the number of classes $C$ which truly exist in a population.  Typically exhaustive sampling is not possible, so our estimation is based on a sample of $c < C$ classes.  The problem is named for its origins in biological ecosystems, where $C$ is \textit{species richness}, the number of species.  However, methods developed for the species problem can be readily applied to applications far removed from biology.  For example, \citet{efron_1976} estimated the number of words $C$ Shakespeare truly knew by considering the words he used in his works to be a sample of $c$ classes.  \citet{fegatelli_2018} estimated the number of cars $C$ covered by an insurer based on accident data where $c$ cars had at least one accident.

In ecology, species richness is worthy of estimation because it is one quantifiable aspect of ecosystem diversity.  We focus our attention on microbial ecology where high species richness is associated with ecosystem stability, including in the human microbiome (see \citet{sekirov2010gut}). Microbial data often contain many species observed only a few times (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).  In data with both rare and abundant species, richness estimation methods tend to have high variance.

Let $f_k$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.  Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent collected data.  Using this notation, the species problem can then be rephrased as predicting $f_0$ (always unobserved) because $C = f_0 + f_1 + f_2 + \dots$ and $c = C - f_0$.  In this paper we focus on the analysis of multiple frequency count tables.  By demonstrating that the penalization proposal due to \citet{wang_2005} is effective in analyzing data with replicates, we extend their results.  Further, we show that replicate data is useful in tuning the required penalization parameter.

This paper will be organized into the following sections:  We will review the literature most pertinent to our investigations in Section \ref{sec:literature_review}.  In Section \ref{sec:fixed_lambda} we specify the  extension to replicate data (more than one frequency count table).  In Section \ref{sec:efficacy_sims} the utility of penalization in replicate analysis is established using simulations.  In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter.  In Section \ref{sec:tuning_simulations} we will evaluate our tuning proposals using simulations.  Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work.

\section{Literature Review}
\label{sec:literature_review}

Poisson mixture models were used in the earliest examinations of species abundance.  In a Poisson model the number of times we observe an arbitrary species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.  The literature is filled with proposals for mixing distributions including the log normal in \citet{bulmer_1974}, inverse Gaussian in \citet{ord_1986} and a nonparametric distribution composed of point masses proposed by \citet{norris_1998}.  In this paper we focus on the simple, easily interpretable gamma distribution, proposed to study species abundance by \citet{fisher_1943}.

Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$.  We use $\eta$ to reference the abundance parameters, so $\eta = (\alpha, \delta)$ in the gamma-Poisson model.  \citet{greenwood_1920} showed that the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
\[ p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x \]
which makes it an especially convenient model to work with.

There exist Bayesian (\citet{efron_1976}), objective Bayesian (\citet{barger_2010}) and other approaches to parameter estimation in addition to the maximum likelihood approach that Fisher used.  However, we will focus on focus on likelihood maximization in this paper.  The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.  A complication in this model is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, so it is not immediately clear that derivative-based optimization methods are appropriate.  This issue has been studied by \citet{lindsay_1987} and they provide a discrete analog to the score function for this model and others.  However, they point out that the asymptotics required use $C \to \infty$, which may not be appropriate for all problems.

There are two more major approaches to likelihood maximization, the \textit{conditional} and \textit{profile} approaches.  In order to understand these we review the work of \citet{sanathanan_1977}, who showed we can decompose the likelihood for the gamma-Poisson model into two parts.
\begin{align}
 L(C, \eta) &= L_b(C, \eta)L_c(\eta) \label{eq:likelihood}\\
 L_b(C, \eta) &= \frac{C!}{(C-c)! \ c!} \left[1 - p_{\eta}(0) \right]^c \left[ p_{\eta}(0) \right]^{C-c}  & \text{Binomial likelihood.} \label{eq:binomial_likelihood}\\
 L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k} \label{eq:conditional_likelihood}  & \text{Conditional likelihood.}
\end{align}
In this decomposition only $L_b$ depends on $C$, while $L_c$ is a function of $\eta$ alone.  $L_b$ shows the probability of observing $c$ species follows a binomial distribution, i.e. $c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$.  Letting $\kmax$ be the largest $k$ such that $f_k > 0$, $L_c$ shows that observing the frequency counts given by $f_1, f_2, \dots , f_{\kmax}$ follows a multinomial distribution given $c$.  The multinomial cell probabilities are proportional to $p_{\eta}(k)$.

\citet{sanathanan_1977} uses this decomposition to propose the conditional approach to likelihood maximization, which is done in two steps (see \citet{chao_2002} for a detailed presentation).  First we maximize $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$.  Then $L_b(C, \widehat{\eta}_c)$ is maximized over $C$.  With $\eta$ fixed there is a closed form expression for this second step,
\begin{equation}
\widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor \label{eq:conditional_c}
\end{equation}
where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.  In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).  The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.

A perennial issue in the species problem is the instability of estimators, especially maximum likelihood estimators.  The estimates often have unacceptably high variance and can become dramatically inflated.  Fisher noted that if there are many low abundance species then $\alpha$ will be near 0.  The existence of highly abundant species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.  The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.

One way of improving stability is to abandon the Poisson mixture paradigm entirely.  Though we will focus on the gamma-Poisson model in this paper, we would be remiss if we did not briefly mention other popular alternatives.  Coverage-based estimators, which use the proportion of species observed $k$ times rather than the actual counts were originally proposed by \citet{good_1953} and remain popular.  Capture-recapture methods (see \citet{chao_1987}) have a deep literature in animal abundance studies but are not as applicable in microbial data where the sampling  destroys the organism.  See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.

A proposal which retains the gamma-Poisson model and encourages stability is suggested by \citet{wang_2005}.  Their proposal is to add a penalty term to the log likelihood, where the penalization is based on the probability of observing a species zero times (they give three penalty functions and we focus on only the first).  If we let $\ell(C,\eta)$ be the log likelihood, then they define a penalized log likelihood $\ell_{\lambda}(C, \eta)$ as
\begin{equation}
\ell_\lambda(C, \eta) = \ell(C,\eta) - \lambda \log p_{\eta}(0)
\label{eq:wang_lindsay}
\end{equation}
where $\lambda > 0$ is a penalization parameter.  This encourages the maximum likelihood solution for $\eta$ to fit so that $p_{\widehat{\eta}}(0)$ is smaller.  Therefore $\alpha$ and $\delta$ are further from the boundary, which alleviates some of the instability. The penalization term also causes a monotonic decrease in $C$ estimates as greater values of $\lambda$ are used.  Consequently, if a large enough $\lambda$ value is applied the $C$ estimates will shrink to the observed richness $c$.

In their paper they show a trade-off exists: greater values of $\lambda$ correspond to a more stable estimator, but at the potential cost of negative bias.  This is a realization of the bias-variance tradeoff.
The choice of penalization parameter implies a preference for lower variance or lower bias, and therefore adds a degree of subjectivity to the estimation procedure (comparable to a Bayesian prior as discussed in \citet{wang_2005}).  We would also expect different data sets to require different $\lambda$ values for optimal estimation, which makes direct application of this method to real data challenging.  Our primary interest in studying the analysis of replicate frequency count tables is to leverage the replicates to tune the penalization parameter in Equation \ref{eq:wang_lindsay}.  The goal of this exploration is to obtain an estimator which is more stable than the unpenalized MLE but does not have the disadvantage of requiring a fixed, user-specified penalization parameter.

\section{Extension to Replicate Data}
\label{sec:fixed_lambda}

Throughout we will focus on the gamma-Poisson model with penalization as proposed by \citet{wang_2005}.  In this section we extend the penalized likelihood for a single frequency count table to the case where multiple frequency count tables have been collected.

Let $r$ be the number of frequency count tables.  We will assume that these $r$ frequency count tables are biological replicates drawn independently from the same population, i.e. they have a common structure specified by the parameters $C$ and $\eta$.  A set of $r$ frequency count tables from the same population will be called a \textit{sample}, and a single frequency count table will be referred to as a \textit{replicate}.  Let the number of species observed $k$ times in replicate $j$ be $f_{kj}$, $j \in \{1, \dots , r\}$.  Let $c_j$ be the observed richness in replicate $j$.

Microbial data is often available as labeled abundance information $\{ X_{ij} \}$, the number of times species $i$ was observed in replicate $j$.  In this paper we study data represented in frequency counts $\{f_{kj}\}$ and it is worth pointing out that we lose some information in this representation.  For instance, in sample of frequency count tables it cannot be determined whether species $i$ was observed in two different replicates.  By using a mixed Poisson model we assume a common abundance structure between the species in order to estimate $C$.  While there may be other methods which leverage the additional information in an abundance table representation, the frequency counts are a minimal sufficient statistic for mixed Poisson models.

In Section \ref{sec:tuning_proposals} we will require an objective function defined in terms of a subset of indices of the frequency count tables, so we define it that way now.  Let $J \subseteq \{1, \dots, r\}$.  $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.  Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.  Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.  We will refer to this as the \textit{penalized log likelihood} and it is given by
\begin{align}
\mathcal{O}_{\lambda}(C, \eta; J) =& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
 =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
 & \qquad  - \sum_{k \geq 1} \log{} f_{kj}! + \sum_{k \geq 1} f_{kj} \log p_\eta(k) \biggr] - \lambda \log p_\eta(0)    \label{eq:objective}
\end{align}
By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.  We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate using $\lambda$, that is
\begin{equation}
\left(\widehat{C}_{\lambda},  \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{\lambda} \left(C, \eta ; \{1, \dots , r\} \right) \label{eq:ccc_hat_lambda}
\end{equation}
where $\widehat{\eta}_{\lambda}$ functions as a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.

We make extensive use of root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.  While this is ubiquitously used we define it here to avoid confusion.  Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.  Then we define
\begin{equation}
 \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
\end{equation}

To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.  While it may be faster and asymptotically equivalent to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.  In this paper our end goal is to explore tuning, so the stability and accuracy of our underlying optimization algorithm have priority over speed.  Crucially, will use exactly the same optimization for the likelihood throughout all simulations in this paper.

To implement a direct optimization approach we search over candidate $C$ values in a grid, respecting the fact that $C$ is a discrete parameter.  A gradient-based search over $\eta$ values is used at each $C$ in the grid.  The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.  While this paper does not focus on optimization, we display our code in full detail for the sake of reproducibility (see section \ref{sec:code} for links).

\section{Efficacy of Penalization}
\label{sec:efficacy_sims}
\subsection{Simulation setup}

Establishing that penalization can improve richness estimates is a crucial foundation before we attempt to evaluate tuning methods.  \citet{wang_2005} established that it can improve estimates for $r = 1$ and in their particular application area.  We need to show this for our models and $r > 1$.  We propose to do so by simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.  If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) over all simulations is less than RMSE($\widehat{C}_0)$ we conclude that penalization is effective under those simulation parameters.

We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates (e.g. $r = 1000$ is unrealistic).  The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.   We used two microbial datasets as motivating examples.  \citet{walsh2014restricted} provide a dataset from soil in a Swiss apple orchard and we found $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$ produced a frequency count table with a similar structure to their data.  The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.  We note that is possible that the gamma-Poisson model is not the best fit for these examples, our goal is only to choose $\eta$ in an appropriate range given the model.  Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.  These add variety to our simulations in this section and context for our simulations in Section \ref{sec:tuning_simulation_2}.  Under each combination of $\eta$ and $r$ 100 simulations were completed.

In Table \ref{tab:eta_intuition} we provide a few summary measures for a single frequency count table generated under each choice of $\eta$.  This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.  We make this information available here for the interested reader but delay our discussion of the differences until Section \ref{sec:tuning_simulations} where they strongly influence our results.

% latex table generated in R 3.5.1 by xtable 1.8-2 package
% Thu May 23 22:11:47 2019
\begin{table}[t]
\tblcaption{Summaries of the proportion of species which are unobserved, observed once (singletons), rarely observed and abundantly observed are calculated for each choice of $\eta$.    %These allow intuition on the probability scale and frequency count scale.  For instance, the proportion of species unobserved is $p_{\eta}(0)$ also equivalent to $\mathbb{E}(f_k)/C$ (see Section \ref{sec:tuning_proposals}).
We also estimate the maximum abundance $k_{\text{max}}$, which is a random variable indicating the number of times the most abundant species was observed.  In this table only $\mathbb{E}(k_{\text{max}})$ depends on $C$, and  $C = 1000$ is used here.
\label{tab:eta_intuition}}
\centering
\footnotesize
\begin{tabular}{rcccc}
  \hline
 $\eta = $ & ($10^{-1}$, $10^{-1}$) & ($10^{-2}$, $10^{-5}$) & ($10^{-1}$, $10^{-3}$) & ($10^{-1}$, $10^{-5}$) \\
  \hline
  Proportion unobserved ($p_{\eta}(0)$) & 0.787 & 0.891 & 0.501 & 0.316 \\
  Proportion singletons ($p_{\eta}(1)$) & 0.072 & 0.009 & 0.050 & 0.032 \\
  Proportion rare ($\sum_{k=1}^3 p_{\eta}(k)$) & 0.130 & 0.016 & 0.097 & 0.061 \\
  Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
  Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1*10^{1}$ & $2.0*10^{5}$ & $4.0*10^{3}$ & $3.9*10^{5}$ \\
   \hline
\end{tabular}
\normalsize
\end{table}
One final setting is necessary for these simulations: a set of $\lambda$ values to test over.  Our primary concern is to ensure the set covers the optimal value, i.e. $\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.  While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to reassure ourselves that the set was sufficiently large after the simulation is complete.  As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.  If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ it will only stray farther from the true value, and the RMSE will increase.  Then we can rest assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.  This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}. Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.
\vspace{30pt}

\begin{table}[h]
\tblcaption{Simulation results for various abundance ($\eta$) and replicate ($r$) choices.  For each row in the table 100 simulations were run, e.g. for the first row we generated a group of 6 frequency count tables under $C = 1000, \eta = (10^{-1},10^{-1})$ one hundred times.  $\lambda_{\text{opt}}$ is the optimal value of $\lambda$, i.e. $\lambda$ which produced the lowest RMSE over all simulations.  $\widehat{C}_0$ is the $C$ estimate with $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the $C$ estimate using $\lambda = \lambda_{\text{opt}}$.
\label{tab:fixed_lambda_results}}
\centering
\begin{tabular}{ccccc}
  \hline
$\eta = (\alpha, \delta)$ & $r$ & RMSE ($\widehat{C}_{0}$) & $\lambda_{opt}$ & RMSE ($\widehat{C}_{\lambda_{ \text{opt}}}$) \\
  \hline
$\left(10^{-1}, 10^{-1}\right)$ &   6 & 796.90 &  10 & 326.50 \\
$\left(10^{-1}, 10^{-1}\right)$ &  10 & 527.53 &  15 & 286.27 \\
  $\left(10^{-1}, 10^{-1}\right)$ &  14 & 337.31 &  10 & 235.51 \\
  $\left(10^{-2}, 10^{-5}\right)$ &   6 & 735.95 &  15 & 516.37 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  10 & 700.21 &  20 & 456.04 \\
  $\left(10^{-2}, 10^{-5}\right)$ &  14 & 666.55 &  20 & 470.35 \\
  $\left(10^{-1}, 10^{-3}\right)$ &   6 & 200.09 &  20 & 156.29 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  10 & 213.22 &  55 & 142.64 \\
  $\left(10^{-1}, 10^{-3}\right)$ &  14 & 243.42 &  55 & 148.26 \\
  $\left(10^{-1}, 10^{-5}\right)$ &   6 & 401.17 &  55 & 147.13 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  10 & 283.91 &  25 & 137.04 \\
  $\left(10^{-1}, 10^{-5}\right)$ &  14 & 415.19 &  70 & 126.72 \\
   \hline
\end{tabular}
\end{table}

\subsection{Simulation results}

The simulation parameters and RMSE results are shown in Table \ref{tab:fixed_lambda_results}.  For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is attained for some $\lambda > 0$.  Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.  These results show that for a variety of $(\eta, r)$ choices penalization is efficacious in improving species richness estimates.  In other words if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.  This extends the results of \citet{wang_2005} to the analysis of replicate frequency count tables and our particular application.  We also note that depending on the simulation settings the optimal $\lambda$ value varies considerably, ranging from 10 to 70.  This simulation therefore provides little evidence that there is a universally appropriate $\lambda$ choice.  Even in our narrowly targeted simulation scheme we have evidence that tuning the $\lambda$ value to the sample is highly desirable.

In order to demonstrate the effect on $\widehat{C}_{\lambda}$ as we increase $\lambda$, we plot the results for two $\eta$ values in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.  For small values of $\lambda$ we acquire lower variance in the estimates while incurring some negative bias, and the RMSE improves as a result.  Eventually the bias difference dominates and the RMSE increases as $\lambda \to \infty$.

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-1}, 10^{-1})$, over 100 simulations. The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta1.pdf}}
\end{figure}

\begin{figure}[p]
\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-2}, 10^{-5})$, over 100 simulations.  The black line indicates the true value $C = 1000$.
\label{fig:fixed_lambda_2}}
\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta2.pdf}}
\end{figure}

%
%
%
\section{Methods for Tuning $\lambda$}
\label{sec:tuning_proposals}

We have established penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.  Therefore a method which can tune the parameter based on a sample is warranted.  We propose several novel and intuitive methods in this section.  Each is evaluated in simulations in Section \ref{sec:tuning_simulations}.

\setcounter{subsection}{-1}
\subsection{Method 0: MLE without penalization}
We begin by proposing a reasonable baseline to compare our tuning methods with.
 The easiest extension to the replicate case is to maximize the unpenalized likelihood using $J = \{1, \dots r \}$.  Recall that fixing $\lambda = 0$ gives us the unpenalized likelihood in Equation \ref{eq:objective}, so if we denote the $C$ estimate obtained using Method 0 as $\widehat{C}_{[0]}$, then
\begin{equation}
\left(\widehat{C}_{[0]},  \widehat{\eta}_{[0]} \right) = \argmax_{C, \, \eta}  \mathcal{O}_{0}\left(C, \eta ; \{1, \dots , r\} \right) \label{eq:c_hat_0}
\end{equation}
This method is faster and makes fewer assumptions than any other method, so it serves as the standard to beat.

\ \\

For all of the remaining methods (1-4) we generate estimates $\widehat{C}_\lambda$ over $\lambda \in \lambda^{\text{grid}}$, where $\lambdagrid$ is a user specified constant required to run each method (e.g. $\lambdagrid = \{0, 5, 10, \dots 60\}$ in Section \ref{sec:tuning_simulation_1}).  The methods differ from one another in how they subset the data to create these estimates, and in the steps they take to arrive at a final estimate $\widehat{C}_{\text{[method]}}$.

%
%
\subsection{Method 1: Minimum subset variance}
Method 1 is motivated by the belief that if we resample from the same population, an ideal $C$ estimator should have low variance.  Exploiting the fact that we have replicate data, the idea is to repeatedly partition the replicates into two subsets and come up with two estimates.  We select the $\lambda$ which yields the lowest between-subset variance.  This partitioning is repeated $p$ times to average out the arbitrary choice of subsets.

Let $T_1(l)$ be the first subset of the $l^{\text{th}}$ partition and $T_2(l)$ be the second subset of the $l^{\text{th}}$ partition, for $l \in \{1, \dots , p\}$ ($T$ is for training).  We select the subsets randomly and we are using the word partition in the mathematical sense.  So $T_1(l), T_2(l) \subseteq \{1, \dots , r\}$, $T_1(l) \cap T_2(l) = \emptyset$ and $T_1(l) \cup T_2(l) = \{1, \dots ,r\}$.  For each $\lambda \in \lambda^{\text{grid}}$, $l \in \{1, \dots, p\}$ let
\begin{align}
\left(\widehat{C}_{\lambda}^{ \ T_1(l)}, \widehat{\eta}_{\lambda}^{ \ T_1(l)} \right) &= \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T_1(l) \right) \\
\left(\widehat{C}_{\lambda}^{ \ T_2(l)}, \widehat{\eta}_{\lambda}^{ \ T_2(l)} \right) &= \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T_2(l) \right)
\end{align}
We now have a $\widehat{C}$ corresponding to each $\lambda \in \lambdagrid$ in each subset of each partition.  Our goal in Method 1 is to use these estimates to select the $\lambda$ value which gave us the lowest average variance over all partitions, denoted by $\widetilde{\lambda}_{[1]}$.  The overall estimate for Method 1, $\widehat{C}_{[1]}$, is a simple average of the estimates produced under $\widetilde{\lambda}_{[1]}$.
\begin{align}
\widetilde{\lambda}_{[1]} &= \argmin_{\lambda} \frac{1}{p} \sum_{l=1}^p \text{Var}\left[ \widehat{C}_{\lambda}^{ \ T_1(l)}, \widehat{C}_{\lambda}^{ \ T_2(l)} \right] \\
\widehat{C}_{[1]} &=  \frac{1}{p} \sum_{l=1}^p \left[ \frac{\widehat{C}_{\widetilde{\lambda}_{[1]}}^{ \ T_1(l)} + \widehat{C}_{\widetilde{\lambda}_{[1]}}^{ \ T_2(l)}}{2} \right]
\end{align}

Practically speaking, in our simulations we use a 50/50 split of the indices for each partition.  The subsets of each partition are selected at random, that is we could select the same subset split more than once.  We partition a total 10 times ($p = 10$).  For example if $r = 4$ then $T_1(1) = \{1,4\}$ and $T_2(1) = \{2, 3\}$ would be a valid subset selection for partition 1.  This holds in methods 2 and 4 as well, which also use partitioning.

%
%
\subsection{Method 2: Cross validated likelihood}
A well known method which takes advantage of replicate data is cross validation.  Cross validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.  The model is trained on one subset of the data and then evaluated using another.  However, in the species problem we never observe true outcomes, i.e. $C$ is always estimated using the frequency counts.  Methods 2 and 4 are motivated by the desire to do something similar to cross validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.

In Method 2 we use one subset of the data to generate estimates.  Then we evaluate those estimates using the likelihood with the data from the other subset.  As before, we partition the data into two subsets $p$ times, calling them $T(l)$ for the training subset of the $l$-th partition, and $E(l)$ for the evaluation subset of the $l$-th partition.  $T(l), E(l)$ are created exactly as in Method 1, we only relabel them to indicate the change in purpose.  Then for each $\lambda \in \lambdagrid$, $l \in \{1, \dots , p\}$ let
\begin{equation}
\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right) \label{eq:part_c_hats_method_2}
\end{equation}
so that we have an estimate using only the training subset for each $\lambda$ value and each partition.  A $\lambda$ value is selected by plugging those estimates into the unpenalized likelihood with $J = E(l)$.
\begin{align}
\widetilde{\lambda}_{[2]} &= \argmax_\lambda \sum_{l=1}^p \mathcal{O}_0\left( \widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} ; \  E(l) \right) \label{eq:selected_lambda_2} \\
\widehat{C}_{[2]} &= \frac{1}{p} \sum_{l=1}^p \widehat{C}_{\widetilde{\lambda}_{[2]}}^{ \ T(l)}
\end{align}
 In the evaluation step given in Equation \ref{eq:selected_lambda_2} we must use the unpenalized likelihood, as opposed to the penalized version.  The penalty term is in fact always positive, and has the potential dominate the likelihood if a very large $\lambda$ exists in $\lambdagrid$. For example, suppose we used the penalized likelihood and $10^7 \in \lambdagrid$.  Then no matter what data was passed in, Method 2 would select $\widetilde{\lambda} = 10^7$, and $\widehat{C}_{[2]}$ would shrink to the observed richness as a consequence.

%
\subsection{Method 3: Goodness of fit}

We depart from our partitioning scheme to explore another desirable property of an estimator.  An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.  Let us momentarily consider a single frequency count table.  Using Equation \ref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
\begin{equation}
 \mathbb{E}\left(f_k\right) = \mathbb{E}\left(f_k|c \right) = \mathbb{E}\left( c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right) = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k)
\end{equation}
Then given estimates $\widehat{C}$ and $\widehat{\eta}$ we use a plug-in estimate for the expected counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.  If we consider each $f_k$ a category, the usual $\chi^2$ goodness of fit statistic is
\begin{align}
&\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} \\
&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
& = \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
\end{align}
Equation \ref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.  Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.  Equation \ref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.  Nonetheless we use Equation \ref{eq:chi_sq_compact} below in extending to replicates for more concise notation.

In method 3 we make use of this goodness of fit metric by first generating estimates using all replicates.  For each $\lambda \in \lambdagrid$ let
\begin{equation}
\left(\widehat{C}_{\lambda}, \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; \left\{1, \dots, r \right\} \right) \label{eq:c_hat_lambdas_method_3}
\end{equation}
$\widehat{C}_{[3]}$ is determined by evaluating the goodness of fit metric shown above for each choice of $\lambda$.  $\widetilde{\lambda}_{[3]}$ is the $\lambda$ value which provides $\widehat{C}_{\lambda}$ estimates which fit the data best.
\begin{align}
\widetilde{\lambda}_{[3]} &= \argmin_{\lambda} \sum_{j=1}^r \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)^2}{\widehat{C}_{\lambda}p_{\widehat{\eta}_{\lambda}}(k)} \label{eq:selected_lambda_3} \\
\widehat{C}_{[3]} &= \widehat{C}_{\widetilde{\lambda}_{[3]}}
\end{align}
In Equation \ref{eq:selected_lambda_3} note the estimated counts $\left( \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)$ are the same over all replicates ($j$).  This follows because we do not have independent $C, \eta$ estimates from each replicate, but a shared estimate for the sample.

\subsection{Method 4: Cross validated goodness of fit}

In Method 4 we return to the partitioning scheme of Method 2.  Rather than using the likelihood in the evaluation step, we hypothesize that the goodness of fit metric may be a better choice.  Partition the data $p$ times into $T(l)$, the training set in partition $l$, and $E(l)$, the evaluation set in partition $l$.  For all $\lambda \in \lambdagrid$ and $l \in \{1, \dots , p \}$, we generate estimates exactly as in Method 2,
\begin{equation}
\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right) \label{eq:c_hat_lambda_method_4}
\end{equation}
To select $\lambda$ we evaluate the goodness of fit metric using only the evaluation subset data, $j \in E(l)$.  The $\lambda$ which produces the best fitting estimates is $\widetilde{\lambda}_{[4]}$,
\begin{align}
\widetilde{\lambda}_{[4]} &= \argmin_{\lambda} \sum_{l = 1}^p \sum_{j \in E(l)} \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda}^{T(l)} p_{\widehat{\eta}_{\lambda}^{T(l)}}(k) \right)^2}{\widehat{C}_{\lambda}^{T(l)}p_{\widehat{\eta}_{\lambda}^{T(l)}}(k)} \\
\widehat{C}_{[4]} &= \frac{1}{p} \sum_{l=1}^p \widehat{C}_{\widetilde{\lambda}_{[4]}}^{T(l)}
\end{align}
Compared to Method 2, this method generates estimates $\widehat{C}^{T(l)}_{\lambda}$ in the same fashion but evaluates in a completely different way.   Compared with Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all the replicates are used in both steps.

\section{Tuning Method Simulations}
\label{sec:tuning_simulations}

We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  However, we have not provided any theoretical guarantees that they are effective, let alone worth the additional computation time compared to Method 0.  In this section we propose to discover whether this is the case by comparing the RMSE of estimates using each method in simulations.

To evaluate which method has the best performance we will apply them to gamma-Poisson simulated data with the same parameters used in Section \ref{sec:efficacy_sims}.  Based on the results in Section \ref{sec:efficacy_sims} we know that estimation can be improved on this set of parameters through the use of a fixed penalization parameter.  The purpose of this section is to determine if any method can reliably select a $\lambda$ that improves the RMSE.

We present two simulation below, and in each it is important to select an appropriate $\lambdagrid$.  In particular our grid needs to be large enough to reveal the negative aspects of methods which over-penalize.  For instance, if $\lambdagrid = \{0, \dots , 50\}$ and the optimal value is $\lambda = 45$, a method which deterministically selects the maximum value in $\lambdagrid$ (an extreme over-penalizer) would perform well.  This would lead us to spuriously conclude this hypothetical over-penalizer is effective.  One of the important foundations laid in Section \ref{sec:efficacy_sims} is that we now know approximately what the optimal values are for each simulation setting (Table \ref{tab:fixed_lambda_results}).  We select a large enough grid by extending to approximately double the optimal value in both simulations presented.

\vspace{50pt}

\subsection{Primary simulation}
\label{sec:tuning_simulation_1}

\subsubsection{Setup}

In this simulation we let $C = 1000$, and simulate 100 times over all combinations of $\eta \in \{(10^{-1}, 10^{-1}), (10^{-2}, 10^{-5}) \}$, $r \in \{6, 10, 14\}$.  Recall these $\eta$ are the values chosen based on our motivating examples.  We know that the optimal $\lambda$ values are between 10 and 20 under these $\eta$ parameters for $r$ frequency count tables, so we chose $\lambdagrid = \{0, 5, 10, \dots 60\}$ for this simulation.  Methods 0-4 are all evaluated over the same random draws for each parameter choice.

\subsubsection{Results}

In Table \ref{tab:tuning_sim_1} we show the RMSE over all simulations for each method and each $\eta, r$ combination.  Over all parameter choices Method 3 performs at least as well as Method 0, with a RMSE which was 0-23\% lower depending on $\eta$ and $r$.  Method 3 is the only method which is at least as good as Method 0 for all parameter choices.

Universally superior RMSE is the most desirable property for a method, but if we recall Table \ref{tab:eta_intuition} we can characterize the performance of the methods based on the simulation parameters as well.  For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species.  The RMSE for Method 2 under this $\eta$ is 14-24\% smaller than Method 0, and for Method 4 the RMSE is slightly better for 2 of the 3 $r$ values.  Conversely, $\eta = (10^{-1}, 10^{-1})$ has more rare species and fewer abundant species and it is only Method 3 which has consistently improved performance.

\begin{table}[ht]
\tblcaption{RMSE$(\widehat{C}_{[\text{method}]})$ over all simulations.  Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.  Under each $\eta, r$ combination 100 simulations were run.  Methods with RMSE better than Method 0 have a \textcolor{blue!50}{blue} background, and the best method is \textbf{bolded}.
\label{tab:tuning_sim_1}}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|}
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (10^{-1},10^{-1})$} & \multicolumn{3}{|c|}{$\eta = (10^{-2},10^{-5})$} \\
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{1}{|c|}{$r = 6$} & $r = 10$ & $r = 14$ & $r = 6$ & $r = 10$ & $r = 14$ \\
\hline
Method 0: MLE (no penalization)& 709 & \textbf{326} & \textbf{339} & 787 & 775 & 716 \\
\hline
Method 1: Minimum subset variance & \cellcolor{blue!25}{\textbf{689}} & 630 & 578 & \cellcolor{blue!25}{763} & 821 & 796 \\
\hline
Method 2: Cross validated likelihood & 797 & 521 & 492 & \cellcolor{blue!25}{\textbf{602}} & \cellcolor{blue!25}{\textbf{658}} & \cellcolor{blue!25}{617} \\
\hline
Method 3: Goodness of fit & \cellcolor{blue!25}{707} & \cellcolor{blue!25}{\textbf{326}} & \cellcolor{blue!25}{\textbf{339}} & \cellcolor{blue!25}{781} & \cellcolor{blue!25}{663} & \cellcolor{blue!25}{\textbf{554}} \\
\hline
Method 4: Cross validated g.o.f. & 812 & 571 & 533 & \cellcolor{blue!25}{738} & 787 & \cellcolor{blue!25}{679}  \\
\hline
\end{tabular}
\end{table}

Method 1 has better performance only for $r = 6$ on both $\eta$ choices.  In general, we would expect that any truly effective method would improve relative to Method 0 as $r$ increases, as this means more data was collected.  The fact that Method 1 loses its edge over Method 0 as we collect more data is a discouraging finding.  Method 3 shows the opposite trend, which is especially clear in Figure \ref{fig:tuning_sim_1} on $\eta = (10^{-2},10^{-5})$.  As we increase $r$ the simulations appear closer to the truth in Method 3, and Method 3 has the best RMSE for $r = 14$ under both $\eta$ values.

In Figure \ref{fig:tuning_sim_1} we display the $C$ estimates for each method.  Recall that as we increase $\lambda$, $\widehat{C}_\lambda$ monotonically decreases.  Therefore we can discuss whether a method is over-penalizing (selecting $\widetilde{\lambda}$ which is larger than optimal) or under-penalizing (selecting $\widetilde{\lambda}$ which is smaller than optimal) without additional information.  If $\widehat{C}$ is too high then the method has under-penalized and if $\widehat{C}$ is too low then the method has over-penalized.

\begin{figure}[t]
\caption{Simulation results for all proposed tuning methods over $\eta = (10^{-1}, 10^{-1})$ and $\eta = (10^{-2}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_1}}
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_1.pdf}}
\end{figure}

It is clear in Figure \ref{fig:tuning_sim_1} that Method 1 over-penalizes for all parameter choices, as all of the estimates are far below the truth.  We can make sense of this observation based on what we observed in Figure \ref{fig:fixed_lambda}: for very high values of $\lambda$ the estimates have low variance because they are driven down to the observed richness $\max_j c_j$.  However the observed richness has a large negative bias.  Method 1 inherits this large bias as it minimizes variance, so RMSE is not effectively minimized.

Method 2 has the opposite behavior: the estimates tend to be too high, so Method 2 underpenalizes.  This is especially noticeable in $\eta = (10^{-2}, 10^{-5})$, where the Method 0 bias is substantial.  Under this $\eta$ value Method 2 does better than Method 0, but the estimates are almost all too high.

Even for $\eta = (10^{-1}, 10^{-1})$, where the Method 0 bias is smaller, we see that Method 2 has poor performance (Table \ref{tab:tuning_sim_1}).  Recall that in each partition of Method 2 we use only half the data to generate the estimates $\widehat{C}^{T(l)}_{\lambda}$.  As a consequence Method 2 has slightly higher variance when compared with Method 0, as evidenced by the interquartile range in Figure \ref{fig:tuning_sim_1} for $\eta = (10^{-1}, 10^{-1})$.  In conclusion, Method 2 does not seem to be effectively tuning $\lambda$, making the cost of using only half the data to generate estimates especially problematic.

Methods 3 and 4 have different penalization tendencies depending on the $\eta$ and $r$ values of the simulation.  This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.  In methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.  Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.

While we find the indications that Method 4 is adaptable to the dataset encouraging, under $\eta = (10^{-1}, 10^{-1})$ the RMSE for Method 4 is 15-75\% greater than for Method 0, while Method 3 is at least as good as Method 0.  Method 3 is adaptive while being simpler, faster and having better performance than Method 4.  Given deficits noted for Method 1 and 2, Method 3 is by far the most promising method based on this simulation.  We propose to conduct a secondary simulation to further challenge Method 3, testing whether it remains superior to Method 0 over a wider range of $\eta$ values.

\subsection{Secondary simulation}
\label{sec:tuning_simulation_2}

\subsubsection{Setup}

In this simulation we will fix $C = 1000$, $r \in \{6, 10, 14\}$ and do 100 simulations for each combination of $\eta, r$ as before.  We let $\eta \in \{ (10^{-1}, 10^{-3}), (10^{-1}, 10^{-5}) \}$ and only consider methods 0 and 3.  To account for the fact that a larger $\lambda$ was optimal for these $\eta$ values (see Table \ref{tab:fixed_lambda_results}) we use $\lambdagrid = \{0,10,20, \dots 140 \}$.

\begin{table}[t]
\tblcaption{RMSE in $\widehat{C}_{[\text{method}]}$.  Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.  Under each $\eta, r$ combination 100 simulations were run.  If Method 3 has better performance than Method 0 is has a \textcolor{blue!50}{blue} background and it is \textbf{bolded}.  This is redundant but consistent with Table \ref{tab:tuning_sim_1}.
\label{tab:tuning_sim_2}}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|}
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (0.1,0.001)$} & \multicolumn{3}{|c|}{$\eta = (0.1,0.00001)$} \\
\cline{2-7}
\multicolumn{1}{c}{} & \multicolumn{1}{|c|}{$r = 6$} & $r = 10$ & $r = 14$ & $r = 6$ & $r = 10$ & $r = 14$ \\
\hline
Method 0: MLE (no penalization) & \textbf{228} & \textbf{236} & \textbf{188} & \textbf{163} & 426 & 681 \\
%\hline
%Method 1: Min. variance & - & - & - & - & - & - \\
%\hline
%Method 2: CV likelihood & - & - & - & - & - & - \\
\hline
Method 3: Goodness of fit & 260 & 274 & 256 & 183 & \cellcolor{blue!25}{\textbf{175}} & \cellcolor{blue!25}{\textbf{191}} \\
%\hline
%Method 4: CV gof & - & - & - & - & - & - \\
\hline
\end{tabular}
\end{table}

\subsubsection{Results}

In Table \ref{tab:tuning_sim_2} we display the RMSE for both methods and each $\eta, r$ combination.  Method 3 has slightly worse performance on all $r$ values for $\eta = (10^{-1}, 10^{-3})$, with a RMSE which was 14-36\% greater than Method 0.  This $\eta$ generates more rare species so the relatively poor performance of Method 3 is consistent with the results our previous simulation.  In Section \ref{sec:tuning_simulation_1}, Method 3 had greatly improved estimation in data with many abundant species, but the improvements over Method 0 were weaker in datasets that had many rare species.

We also note that Method 3 has 12\% higher RMSE on $r=6, \eta = (10^{-1}, 10^{-5})$, but as $r$ increases Method 3 improves relative to Method 0.  This finding is also consistent with our previous simulation where Method 3 tends to perform better as more data is introduced.

In Figure \ref{fig:tuning_sim_2} we display boxplots of the $C$ estimates over all simulations.  For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is noticeably inflated on a few simulations.  Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species (while still having some singletons and unobserved species).  Therefore this is a excellent example of a dataset which will cause the instability problem in maximum likelihood estimation we reviewed in Section \ref{sec:literature_review}.  Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.
%In this figure we can see that Method 3 improves the RMSE on $\eta = (10^{-1}, 10^{-5})$ primarily by lowering the estimates in a few simulations where $\widehat{C}_{[0]}$ is inflated (4000-6500).

\begin{figure}[t]
\caption{Simulation results for methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
\label{fig:tuning_sim_2}}
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_2.pdf}}
\end{figure}

\vspace{-8pt}
\subsection{Simulation Conclusions}
\vspace{-8pt}
As a result of these simulations, we conclude that none of the methods (including method 0) are best for all simulation settings.  Method 1 resulted in a strong negative bias no matter which data are being analyzed.  Method 3 had better performance than Method 0 when there are highly abundant species present in the data, which makes Method 0 highly unstable.  When there are many rare species Method 3 is better than methods 1,2 and 4 but not necessarily Method 0.  Method 3 also improved relative to Method 0 with greater $r$ and shows qualitative signs of adapting to the data.  Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but we cannot recommend them over Method 3 due to their inferior performance on data dominated by rare species.

\section{Discussion}
\label{sec:discussion}

In this paper we have proposed an extension of the penalization in \citet{wang_2005} to data with biological replicates and proposed several methods for tuning the penalization parameter.  This extends the existing body of literature on species richness by demonstrating that penalization is useful in analyzing replicate data.  Our goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.  In this section we note promising directions for future work.

Our results from Section \ref{sec:fixed_lambda} hint at a positive correlation between $r$ and the optimal choice of $\lambda$.  We hypothesized that this may be the case and considered other ways of extending the objective function given in Equation \ref{eq:objective}.  For instance, using $-h(J)\lambda \log p_{\eta}(0)$ where $h(J)$ is $|J|$ or $\sqrt{|J|}$ seemed like plausible extensions, whereas the definition we give uses $h(J) = 1$.  Our investigations revealed that the trend is unfortunately not this simple as $|J|$ becomes large.  A parsimonious extension which accounts for this observed trend would be a valuable modification to our work.

In \citet{wang_2005} the gamma-Poisson model is explicitly considered but their results apply to many other Poisson mixture models.  Our most promising method, Method 3, is also amenable to other modeling choices provided that we have some way to calculate the estimated counts $\widehat{f_k}$.  It would be interesting to explore whether our results hold for a more flexible mixed Poisson distribution, such as the proposal by \citet{norris_1998}.  Flexible models often have better performance under model misspecification, and we simulated under correctly specified models.  Therefore, this variation on  Method 3 would be especially interesting to study under model misspecification.

We intentionally chose a likelihood optimization algorithm which was stable and exhaustive but admittedly slow.  Confirming that our results hold when a faster approach is used is another valuable extension to our work. We did not derive theoretical standard errors in this paper, and the long computation times precluded the consideration of a bootstrapping approach.  However, if our results hold when a faster likelihood maximization algorithm is used then a standard error procedure could become feasible.

We have been careful to try to create methods which only depend on $\lambdagrid$ being large enough to contain the optimal value, with no change in results if it is too large.  This makes $\lambdagrid$ a more forgiving parameter than fixed $\lambda$ values, but we still need to make a choice to analyze data.  Faster computation times would aid in this problem because we could set a default grid which is generously large, thus alleviating concerns that our results are dependent on our choice.  Additionally, it may be possible to select $\lambdagrid$ for the methods adaptively.  For instance, we could expand the grid until the fixed estimates (e.g. $\widehat{C}_{\lambda}$ in Equation \ref{eq:c_hat_lambdas_method_3}) have been suppressed to the observed richness.  Choosing $\lambdagrid$ remains the most difficult part of applying our methods to real data.

Method 3 was our top performer and it is the only tuning method we proposed which can be run without replicates ($r = 1$).  The ever-decreasing cost of high-throughput sequencing suggests biological replicates might be more available in the future.  However, there are currently more projects with a single frequency count table available.  Exploring the use of Method 3 for single frequency count tables is a worthwhile extension of our work for this reason.

% meh:
%\textcolor{red}{Find a home for me:
%However, we also point out that methods which deal with replicate frequency count tables have an inherent use.  While it is possible to create a single frequency count table for multiple samples with labeled abundance information $\{X_i\}$, many data sets are published with only the frequency counts $\{f_k\}$.  Analyzing two data sets from the same population which are stored as frequency counts is not possible with current methods, as there is no way to aggregate those into a single frequency count table.}

\section{Code References}
\label{sec:code}
Our method implementations including our core objective optimization algorithm can be found at \texttt{github.com/statdivlab/rre}.  The code to reproduce our figures and simulations can be found at \texttt{github.com/statdivlab/rre\_sims}.  All code is written in \texttt{R} (\citet{r_project}).

\section*{Acknowledgments}
List of acknowledgements for formalize:

{\it Conflict of Interest}:

\bibliographystyle{biorefs}
\bibliography{refs}

\end{document}

### 2019-06-14 rre_manuscript/paper.tex
thesis-as-paper

@@
 % List of authors, with corresponding author marked by asterisk
-\author{ALEX PAYNTER$^\ast$, AMY WILLIS\\[4pt]
+\author{ALEX PAYNTER$^\ast$, AMY WILLIS$^\ast$\\[4pt]
 % Author addresses
@@
 \end{equation}
-where $\widehat{\eta}_{\lambda}$ functions as a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.
+where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.
 
-We make extensive use of root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.
-While this is ubiquitously used we define it here to avoid confusion.
+We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.
 Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.
@@
 The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.
-While this paper does not focus on optimization, we display our code in full detail for the sake of reproducibility (see section \ref{sec:code} for links).
+While this paper does not focus on optimization, our code is made available to the reader (see Section \ref{sec:code} for links).
 
@@
 
-We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates (e.g.
-$r = 1000$ is unrealistic).
+We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates.
 The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.
@@
 The second example from \citet{tromas_2017} is a sample of cyanobacteria from Lake Champlain, and we found $\eta = (10^{-2}, 10^{-5})$ mimicked the structure of their data well.
-We note that is possible that the gamma-Poisson model is not the best fit for these examples, our goal is only to choose $\eta$ in an appropriate range given the model.
+We note that is possible that the gamma-Poisson model is not the best fit for these examples, but our goal is only to choose $\eta$ in an appropriate range.
 Two nearby $\eta$ values are included, $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
@@
 
-In Table \ref{tab:eta_intuition} we provide a few summary measures for a single frequency count table generated under each choice of $\eta$.
+In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.
 This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.
@@
 \tblcaption{Summaries of the proportion of species which are unobserved, observed once (singletons), rarely observed and abundantly observed are calculated for each choice of $\eta$.
-%These allow intuition on the probability scale and frequency count scale.
-For instance, the proportion of species unobserved is $p_{\eta}(0)$ also equivalent to $\mathbb{E}(f_k)/C$ (see Section \ref{sec:tuning_proposals}).
-We also estimate the maximum abundance $k_{\text{max}}$, which is a random variable indicating the number of times the most abundant species was observed.
+We give the expected maximum abundance $k_{\text{max}}$, the number of times the most abundant species was observed.
 In this table only $\mathbb{E}(k_{\text{max}})$ depends on $C$, and  $C = 1000$ is used here.
@@
   Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\ 
-  Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1*10^{1}$ & $2.0*10^{5}$ & $4.0*10^{3}$ & $3.9*10^{5}$ \\
+Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\ 
    \hline
@@
 $\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.
-While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to reassure ourselves that the set was sufficiently large after the simulation is complete.
+While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to confirm that the set was sufficiently large after the simulation is complete.
 As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.
-If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ it will only stray farther from the true value, and the RMSE will increase.
-Then we can rest assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.
+If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ $\widehat{C}_{\lambda}$ will be further from $C$, and the RMSE will increase.
+Then we can be assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.
 This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.
 Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.
-\vspace{30pt}
 
@@
 Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.
-These results show that for a variety of $(\eta, r)$ choices penalization is efficacious in improving species richness estimates.
-In other words if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.
+These results show that for a variety of $(\eta, r)$ choices penalization improves species richness estimates.
+In other words, if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.
 This extends the results of \citet{wang_2005} to the analysis of replicate frequency count tables and our particular application.
 We also note that depending on the simulation settings the optimal $\lambda$ value varies considerably, ranging from 10 to 70.
-This simulation therefore provides little evidence that there is a universally appropriate $\lambda$ choice.
-Even in our narrowly targeted simulation scheme we have evidence that tuning the $\lambda$ value to the sample is highly desirable.
+This simulation suggests that there is not a universally appropriate $\lambda$ choice.
+Therefore, we have evidence that tuning the $\lambda$ value to the sample is highly desirable.
 
@@
 \label{fig:fixed_lambda}}
-\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta1.pdf}}
+\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{./images/fixed_lambda_eta1.pdf}}
 \end{figure}
@@
 \label{fig:fixed_lambda_2}}
-\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{images/fixed_lambda_eta2.pdf}}
+\centering\makebox[\textwidth]{\includegraphics[width=1\textwidth]{./images/fixed_lambda_eta2.pdf}}
 \end{figure}
@@
 
-We have established penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.
+We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.
 Therefore a method which can tune the parameter based on a sample is warranted.
@@
 %
+%
 \subsection{Method 2: Cross validated likelihood}
@@
 %
+%
+%
 \subsection{Method 3: Goodness of fit}
@@
 An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.
-Let us momentarily consider a single frequency count table.
+Consider a single frequency count table.
 Using Equation \ref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
@@
 
+%
+%
+%
 \subsection{Method 4: Cross validated goodness of fit}
@@
 
-We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.
-However, we have not provided any theoretical guarantees that they are effective, let alone worth the additional computation time compared to Method 0.
-In this section we propose to discover whether this is the case by comparing the RMSE of estimates using each method in simulations.
+We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the RMSE of estimates from each method in simulations.
 
@@
 We select a large enough grid by extending to approximately double the optimal value in both simulations presented.
-
-\vspace{50pt}
 
@@
 
-Universally superior RMSE is the most desirable property for a method, but if we recall Table \ref{tab:eta_intuition} we can characterize the performance of the methods based on the simulation parameters as well.
-For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species.
+Universally superior RMSE is the most desirable property for a method, but we can characterize the performance of the methods based on the simulation parameters as well.
+For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
 The RMSE for Method 2 under this $\eta$ is 14-24\% smaller than Method 0, and for Method 4 the RMSE is slightly better for 2 of the 3 $r$ values.
@@
 \label{fig:tuning_sim_1}}  
-\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_1.pdf}}
+\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/method_sim_1.pdf}}
 \end{figure}
@@
 This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.
-In methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.
+In Methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.
 Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.
@@
 Method 0: MLE (no penalization) & \textbf{228} & \textbf{236} & \textbf{188} & \textbf{163} & 426 & 681 \\
-%\hline
-%Method 1: Min. variance & - & - & - & - & - & - \\
-%\hline
-%Method 2: CV likelihood & - & - & - & - & - & - \\
 \hline
 Method 3: Goodness of fit & 260 & 274 & 256 & 183 & \cellcolor{blue!25}{\textbf{175}} & \cellcolor{blue!25}{\textbf{191}} \\
-%\hline
-%Method 4: CV gof & - & - & - & - & - & - \\
 \hline
@@
 Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.
-%In this figure we can see that Method 3 improves the RMSE on $\eta = (10^{-1}, 10^{-5})$ primarily by lowering the estimates in a few simulations where $\widehat{C}_{[0]}$ is inflated (4000-6500).
 
@@
 \label{fig:tuning_sim_2}}  
-\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{images/method_sim_2.pdf}}
+\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/method_sim_2.pdf}}
 \end{figure}
 
-\vspace{-8pt}
 \subsection{Simulation Conclusions}
-\vspace{-8pt}
+
 As a result of these simulations, we conclude that none of the methods (including method 0) are best for all simulation settings.
@@
 Method 3 had better performance than Method 0 when there are highly abundant species present in the data, which makes Method 0 highly unstable.
-When there are many rare species Method 3 is better than methods 1,2 and 4 but not necessarily Method 0.
-Method 3 also improved relative to Method 0 with greater $r$ and shows qualitative signs of adapting to the data.
+When there are many rare species Method 3 is better than methods 1, 2 and 4 but not necessarily Method 0.
+Method 3 also improved relative to Method 0 with greater $r$ and appears to better adapt to the data.
 Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but we cannot recommend them over Method 3 due to their inferior performance on data dominated by rare species.
@@
 In \citet{wang_2005} the gamma-Poisson model is explicitly considered but their results apply to many other Poisson mixture models.
-Our most promising method, Method 3, is also amenable to other modeling choices provided that we have some way to calculate the estimated counts $\widehat{f_k}$.
+Our most promising method, Method 3, is also amenable to other modeling choices provided that we have some way to calculate the fitted counts $\widehat{f_k}$.
 It would be interesting to explore whether our results hold for a more flexible mixed Poisson distribution, such as the proposal by \citet{norris_1998}.
@@
 We intentionally chose a likelihood optimization algorithm which was stable and exhaustive but admittedly slow.
-Confirming that our results hold when a faster approach is used is another valuable extension to our work.
+Confirming that our results hold when a faster approach is used would be another valuable extension to our work.
 We did not derive theoretical standard errors in this paper, and the long computation times precluded the consideration of a bootstrapping approach.
@@
 $\widehat{C}_{\lambda}$ in Equation \ref{eq:c_hat_lambdas_method_3}) have been suppressed to the observed richness.
-Choosing $\lambdagrid$ remains the most difficult part of applying our methods to real data.
+Choosing $\lambdagrid$ remains a consideration when applying our methods.
 
@@
 Exploring the use of Method 3 for single frequency count tables is a worthwhile extension of our work for this reason.
-
-% meh:
-%\textcolor{red}{Find a home for me:
-%However, we also point out that methods which deal with replicate frequency count tables have an inherent use.
-While it is possible to create a single frequency count table for multiple samples with labeled abundance information $\{X_i\}$, many data sets are published with only the frequency counts $\{f_k\}$.
-Analyzing two data sets from the same population which are stored as frequency counts is not possible with current methods, as there is no way to aggregate those into a single frequency count table.}
 

### 2019-06-14 rre_manuscript/paper.tex
removed some extaneous macros
minor updates to title section

--- 2019-06-14 rre_manuscript/paper.tex
@@
 \usepackage{mathtools}
-\usepackage[table]{xcolor}
+\usepackage[table,dvipsnames]{xcolor}
+
+\usepackage{ifthen}
+\usepackage{pstricks}
+
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 
+\newboolean{DEBUG}
+\setboolean{DEBUG}{true}
+
+\newrgbcolor{amycolor}{.7 .1 .1}
+\ifthenelse {\boolean{DEBUG}}
+{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
+{\newcommand{\amy}[1]{}}
+
+\newrgbcolor{alexcolor}{.1 .7 .7}
+\ifthenelse {\boolean{DEBUG}}
+{\newcommand{\alex}[1]{{\bf \alexcolor{[\@alex: #1]}}}}
+{\newcommand{\brian}[1]{}}
+
 \begin{document}
@@
 % Title of paper
-\title{Tuning parameter selection for a penalized maximum likelihood estimator of species richness}
+\title{Tuning parameter selection for a penalized estimator of species richness}
 
 % List of authors, with corresponding author marked by asterisk
-\author{ALEX PAYNTER$^\ast$, AMY WILLIS\\[4pt]
+\author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
 % Author addresses
@@
 1705 NE Pacific St., Seattle, WA 98195}
-\\[2pt]
-% E-mail address for correspondence
-{payntera@uw.edu adwillis@uw.edu}}
+% \\[2pt]
+% % E-mail address for correspondence
+% {payntera@uw.edu, adwillis@uw.edu}
+}
 
@@
 % First field is the short list of authors
-{A. Paynter and others}
+{A. Paynter and A.D. Willis}
 % Second field is the short title of the paper
-{Tuning parameter selection in species richness estimation}
+{Tuning parameter selection for species richness}
 
@@
 % identified in the author list
-\footnotetext{To whom correspondence should be addressed.}
+\footnotetext{adwillis@uw.edu}
 
@@
 \end{abstract}
+
+\amy{Happy to discuss, but I think it's probably easiest for me to be corresponding author, because you won't have your UW e-mail forever}
 

### 2019-06-14 rre_manuscript/paper.tex

--- 2019-06-14 rre_manuscript/paper.tex
@@
 \documentclass[oupdraft]{bio}
-% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
 \usepackage{url}
-
-% Add history information for the article if required
-% \history{Received August 1, 2010;
-% revised October 1, 2010;
-% accepted for publication November 1, 2010}
 
@@
 \usepackage{mathtools}
-\usepackage[table]{xcolor}
+\usepackage[table,dvipsnames]{xcolor}
+
+\usepackage{ifthen}
+\usepackage{pstricks}
+
 \DeclareMathOperator*{\argmin}{arg\,min}
 \DeclareMathOperator*{\argmax}{arg\,max}
-\newcommand{\ev}[1]{\mathbb{E}\left[#1\right]}
-\newcommand{\var}[1]{\text{Var}\left[#1\right]}
 
-% \newcommand{\kmax}{k_{\text{max}}}
-% \newcommand{\kmax}[1]{k_{\text{max}(#1)}}
-% \newcommand{\kmax}[jj]{k_{\text{max}#1}}
+\newcommand{\kmax}{k_{\text{max}}}
+\newcommand{\lambdagrid}{\lambda^{\text{grid}}}
 
-\NewDocumentCommand{\kmax}{o}{%
-    {%
-        \IfNoValueTF {#1}{
-          k_{\text{max}} 
-        }{ 
-          k_{\text{max}(#1)} 
-        }%
-    }%
-}
-\newcommand{\lambdagrid}{\lambda^{\text{grid}}}
+\newboolean{DEBUG}
+\setboolean{DEBUG}{true}
+
+\newrgbcolor{amycolor}{.7 .1 .1}
+\ifthenelse {\boolean{DEBUG}}
+{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
+{\newcommand{\amy}[1]{}}
+
+\newrgbcolor{alexcolor}{.1 .7 .7}
+\ifthenelse {\boolean{DEBUG}}
+{\newcommand{\alex}[1]{{\bf \alexcolor{[\@alex: #1]}}}}
+{\newcommand{\brian}[1]{}}
 
@@
 % Title of paper
-\title{Tuning parameter selection for a penalized maximum likelihood estimator of species richness}
+\title{Tuning parameter selection for a penalized estimator of species richness}
 
 % List of authors, with corresponding author marked by asterisk
-\author{ALEX PAYNTER$^\ast$, AMY WILLIS$^\ast$\\[4pt]
+\author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
 % Author addresses

### 2019-06-14 rre_manuscript/paper.tex
edit S1

@@
 {\newcommand{\alex}[1]{{\bf \alexcolor{[\@alex: #1]}}}}
-{\newcommand{\brian}[1]{}}
+{\newcommand{\alex}[1]{}}
+
+\newrgbcolor{editcolor}{.1 .5 .7}
+\ifthenelse {\boolean{DEBUG}}
+{\newcommand{\edit}[1]{{\bf \editcolor{[\@edit: #1]}}}}
+{\newcommand{\edit}[1]{}}
 
@@
 We focus on the scenario where multiple frequency count tables have been collected from the same population.
-In this setting we demonstrate the efficacy of a previously published penalized maximum likelihood method.
+In this setting we demonstrate the \edit{efficacy of a previously published penalized maximum likelihood method}.
 Four novel methods to tune the requisite penalization parameter are proposed.
@@
 
-In the \textit{species problem} the goal is estimating the number of classes $C$ which truly exist in a population.
-Typically exhaustive sampling is not possible, so our estimation is based on a sample of $c < C$ classes.
-The problem is named for its origins in biological ecosystems, where $C$ is \textit{species richness}, the number of species.
-However, methods developed for the species problem can be readily applied to applications far removed from biology.
-For example, \citet{efron_1976} estimated the number of words $C$ Shakespeare truly knew by considering the words he used in his works to be a sample of $c$ classes.
-\citet{fegatelli_2018} estimated the number of cars $C$ covered by an insurer based on accident data where $c$ cars had at least one accident.
+The \textit{species problem} concerns estimating the number of classes $C$ that are present in a population.
+$n$ individuals from the population can be sampled to find which classes they belong to, but since the  number of individuals in the sample is less than the total population size (which may be infinite), only $c \leq C$ classes are observed.
+The problem is named for its origins in biological ecosystems, where $C$ is \textit{species richness}, or the total number of species.
+However, methods developed for the species problem can be applied to applications far removed from biology.
+For example, \citet{efron_1976} estimated the number of words Shakespeare truly knew by modeling the frequencies of words in his published work, and  \citet{fegatelli_2018} estimated the number of cars covered by an insurer based on accident data where $c$ cars had at least one accident.
 
-In ecology, species richness is worthy of estimation because it is one quantifiable aspect of ecosystem diversity.
-We focus our attention on microbial ecology where high species richness is associated with ecosystem stability, including in the human microbiome (see \citet{sekirov2010gut}).
-Microbial data often contain many species observed only a few times (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).
-In data with both rare and abundant species, richness estimation methods tend to have high variance.
+In ecology, species richness is a quantitative measurement of ecosystem diversity.
+We focus on the specific application of estimating microbial diversity, that is, the number of strains of bacteria present in a population (e.g., a lake microenvironment or in an individual's oral cavity).
+Microbial diversity is often linked with ecosystem health, such as in the vaginal microbiome (where high diversity is associated with infection \cite{placeholder}) and in the gut microbiome (where high diversity is associated with healthy metabolism \cite{placeholder}).
+Microbial abundance data typically contain many species observed infrequently (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).
+In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{willis_2015,bunge_2014} \amy{confirm most appropriate reference}, motivating a regularised estimation approach.
 
-Let $f_k$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.
-Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent collected data.
-Using this notation, the species problem can then be rephrased as predicting $f_0$ (always unobserved) because $C = f_0 + f_1 + f_2 + \dots$ and $c = C - f_0$.
-In this paper we focus on the analysis of multiple frequency count tables.
-By demonstrating that the penalization proposal due to \citet{wang_2005} is effective in analyzing data with replicates, we extend their results.
-Further, we show that replicate data is useful in tuning the required penalization parameter.
+In this paper we consider the penalised maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter.
+Because the true species richness $C$ is never observed for any sample, standard approaches for tuning parameter selection such as cross validation cannot be readily applied to the species problem.
+We propose analyzing biological replicates for tuning, thereby extending the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
+We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.
 
-This paper will be organized into the following sections:  We will review the literature most pertinent to our investigations in Section \ref{sec:literature_review}.
-In Section \ref{sec:fixed_lambda} we specify the  extension to replicate data (more than one frequency count table).
-In Section \ref{sec:efficacy_sims} the utility of penalization in replicate analysis is established using simulations.
-In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter.
-In Section \ref{sec:tuning_simulations} we will evaluate our tuning proposals using simulations.
+This paper is organized as follows: We review species richness estimation in Section \ref{sec:literature_review}.
+In Section \ref{sec:fixed_lambda} we describe our extension to biological replicates.
+In Section \ref{sec:efficacy_sims} we establish the utility of penalization via simulations.
+In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter, and in Section \ref{sec:tuning_simulations}
+we evaluate our proposals.
 Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work.
+Software implementing the methods is available in the \url{R} package \url{rre}, available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.
 
 \section{Literature Review}
+
 \label{sec:literature_review}
 
-Poisson mixture models were used in the earliest examinations of species abundance.
-In a Poisson model the number of times we observe an arbitrary species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
-The literature is filled with proposals for mixing distributions including the log normal in \citet{bulmer_1974}, inverse Gaussian in \citet{ord_1986} and a nonparametric distribution composed of point masses proposed by \citet{norris_1998}.
-In this paper we focus on the simple, easily interpretable gamma distribution, proposed to study species abundance by \citet{fisher_1943}.
+The classical model for estimating species abundance is a Poisson mixture model.
+Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
+Many mixing distributions have been proposed (see, e.g., \cite{bulmer_1974,ord_1986,norris_1998}, and \cite{bunge_1993} for a comprehensive review).
+While our proposal may be easily generalized to other mixing distributions, in this paper we consider the simple, interpretable gamma distribution, which was first proposed to model species abundances by \citet{fisher_1943}.
+Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.
+Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data.
+The species problem is to predicte $f_0$, because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$.
 

### 2019-06-14 rre_manuscript/paper.tex
edit label

@@
 \label{sec:literature_review}
-
+\amy{up to here with editing}
 The classical model for estimating species abundance is a Poisson mixture model.
@@
 Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data.
-The species problem is to predicte $f_0$, because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$.
+The species problem is predicting $f_0$, because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$.
 

### 2019-07-08 rre_manuscript/paper.tex
updates to intro and lit review

--- 2019-06-14 rre_manuscript/paper.tex
@@
 \begin{abstract}
-{Our goal is estimating the true number of classes in a population.
-We focus on the scenario where multiple frequency count tables have been collected from the same population.
-In this setting we demonstrate the \edit{efficacy of a previously published penalized maximum likelihood method}.
-Four novel methods to tune the requisite penalization parameter are proposed.
-The performance of all proposed tuning methods is compared in simulations.}
+{Our goal is to estimate the true number of classes in a population, called the species richness.
+We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
+Our method penalises the probability of a class being unobserved, since high probabilities of unobserved classes increase the variance of species richness estimates.
+Tuning the penalisation parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods to tune the penalization parameter.
+We illustrate the performance of the method by estimating the microbial diversity of Lake Champlain over ...}
 \end{abstract}
@@
 The \textit{species problem} concerns estimating the number of classes $C$ that are present in a population.
-$n$ individuals from the population can be sampled to find which classes they belong to, but since the  number of individuals in the sample is less than the total population size (which may be infinite), only $c \leq C$ classes are observed.
+$n$ individuals from the population can be sampled to find which classes they belong to, but since the  number of individuals in the sample is less than the total population size (which may be infinite), only $c$ classes are observed ($c \leq C$).
 The problem is named for its origins in biological ecosystems, where $C$ is \textit{species richness}, or the total number of species.
@@
 We focus on the specific application of estimating microbial diversity, that is, the number of strains of bacteria present in a population (e.g., a lake microenvironment or in an individual's oral cavity).
-Microbial diversity is often linked with ecosystem health, such as in the vaginal microbiome (where high diversity is associated with infection \cite{placeholder}) and in the gut microbiome (where high diversity is associated with healthy metabolism \cite{placeholder}).
+Microbial diversity is often linked with ecosystem health, such as in the vaginal microbiome (where high diversity is associated with infection \citep{Oakley:2008uo}) and in the gut microbiome (where high diversity is associated with healthy metabolism \citep{Morgan:2012bm,Minot:2019hc}).
 Microbial abundance data typically contain many species observed infrequently (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).
-In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{willis_2015,bunge_2014} \amy{confirm most appropriate reference}, motivating a regularised estimation approach.
+In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{wang_2005,willis_2015}, motivating our regularised approach to estimation.
 
 In this paper we consider the penalised maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter.
-Because the true species richness $C$ is never observed for any sample, standard approaches for tuning parameter selection such as cross validation cannot be readily applied to the species problem.
-We propose analyzing biological replicates for tuning, thereby extending the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
+Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross validation) cannot be readily applied in our case.
+We propose utilizing biological replicates to aid tuning, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
 We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.
@@
 This paper is organized as follows: We review species richness estimation in Section \ref{sec:literature_review}.
-In Section \ref{sec:fixed_lambda} we describe our extension to biological replicates.
-In Section \ref{sec:efficacy_sims} we establish the utility of penalization via simulations.
+In Section \ref{sec:fixed_lambda} we describe our extension to biological replicates, and in Section \ref{sec:efficacy_sims} we establish the utility of penalization via simulations.
 In Section \ref{sec:tuning_proposals} we propose several novel methods for tuning the penalization parameter, and in Section \ref{sec:tuning_simulations}
 we evaluate our proposals.
-Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work.
+\amy{Section \ref{sec:data_analysis} applies our methods to microbiome data on Lake Champlain...} Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work.
 Software implementing the methods is available in the \url{R} package \url{rre}, available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.
@@
 \label{sec:literature_review}
+\amy{up to here with editing}
 
@@
 Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
-Many mixing distributions have been proposed (see, e.g., \cite{bulmer_1974,ord_1986,norris_1998}, and \cite{bunge_1993} for a comprehensive review).
-While our proposal may be easily generalized to other mixing distributions, in this paper we consider the simple, interpretable gamma distribution, which was first proposed to model species abundances by \citet{fisher_1943}.
+Many mixing distributions have been proposed, including  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
+While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, a proposal first considered by \citet{fisher_1943}.
+
 Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.
 Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data.
-The species problem is to predicte $f_0$, because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$.
+Because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$, the species problem can be framed as predicting $f_0$ to obtain the species richness estimate $\hat{C} = \hat{f}_0 + c$.
 
-Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$.
-We use $\eta$ to reference the abundance parameters, so $\eta = (\alpha, \delta)$ in the gamma-Poisson model.
-\citet{greenwood_1920} showed that the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
-\[ p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x \]
-which makes it an especially convenient model to work with.
-
-There exist Bayesian (\citet{efron_1976}), objective Bayesian (\citet{barger_2010}) and other approaches to parameter estimation in addition to the maximum likelihood approach that Fisher used.
-However, we will focus on focus on likelihood maximization in this paper.
+Let $\Lambda \sim \text{Gamma}(\alpha, \delta)$ with distribution function $f(\lambda) = \delta^\alpha \Gamma (\alpha)^{-1} \lambda^{\alpha - 1} e^{-\delta \lambda}$, and write $\eta = (\alpha, \delta)$.
+% Then the distribution of $X_i$ under a gamma-Poisson model can written as a negative-binomial distribution,
+% \begin{align}
+% p_{\eta}(x) = \frac{\Gamma (x + \alpha)}{\Gamma (\alpha) x!} \left(\frac{\delta}{1+\delta} \right)^\alpha \left( \frac{1}{1+ \delta} \right)^x.
+% \end{align}
+% Note that we are interested in estimating the parameters of a negative binomial
+% a zero-truncated negative binomial
+%
+There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model.
+We will focus on focus on frequentist maximum likelihood in this paper.
 The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.
-A complication in this model is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, so it is not immediately clear that derivative-based optimization methods are appropriate.
-This issue has been studied by \citet{lindsay_1987} and they provide a discrete analog to the score function for this model and others.
-However, they point out that the asymptotics required use $C \to \infty$, which may not be appropriate for all problems.
-
-There are two more major approaches to likelihood maximization, the \textit{conditional} and \textit{profile} approaches.
-In order to understand these we review the work of \citet{sanathanan_1977}, who showed we can decompose the likelihood for the gamma-Poisson model into two parts.
+A complication in this model is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, and so derivative-based optimization methods are not appropriate.
+This issue has been studied by \citet{lindsay_1987}, who provide a discrete analog to the score function for this model.
+%However, they point out that the asymptotics required use $C \to \infty$, which may not be appropriate for all problems.
+Two other approaches to likelihood maximization are the \textit{conditional} and \textit{profile} approaches.
+The conditional approach \citep{sanathanan_1977} involves writing the likelihood as the product
 \begin{align}
- L(C, \eta) &= L_b(C, \eta)L_c(\eta) \label{eq:likelihood}\\
- L_b(C, \eta) &= \frac{C!}{(C-c)! \ c!} \left[1 - p_{\eta}(0) \right]^c \left[ p_{\eta}(0) \right]^{C-c}  & \text{Binomial likelihood.} \label{eq:binomial_likelihood}\\
- L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k} \label{eq:conditional_likelihood}  & \text{Conditional likelihood.}
+ L(C, \eta) &= L_b(C, \eta)L_c(\eta) \label{eq:likelihood}
 \end{align}
-In this decomposition only $L_b$ depends on $C$, while $L_c$ is a function of $\eta$ alone.
-$L_b$ shows the probability of observing $c$ species follows a binomial distribution, i.e.
+where
+ \begin{align}
+ L_b(C, \eta) &= \frac{C!}{(C-c)! \ c!} \left[1 - p_{\eta}(0) \right]^c \left[ p_{\eta}(0) \right]^{C-c}  %& \text{(``Binomial'' component)}
+  \label{eq:binomial_likelihood}\\
+ L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k}, \label{eq:conditional_likelihood} % & \text{(Multinomial component/``conditional'')}
+\end{align}
+i.e., the likelihood is the product of a binomial probability mass function ($c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$), and a multinomial probability mass function ($(f_1, f_2, \ldots)|c \sim \text{Multinomial}\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$).
+This decomposition is convenient because $L_c$ is a function of $\eta$ alone.
+% $L_b$ shows the probability of observing $c$ species follows a binomial distribution, i.e.
 $c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$.
@@
 The multinomial cell probabilities are proportional to $p_{\eta}(k)$.
+\citet{sanathanan_1977}'s conditional approach involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
 
@@
 \bibliographystyle{biorefs}
-\bibliography{refs}
+\bibliography{refs,amys-papers-library}
 

### 2019-07-08 rre_manuscript/paper.tex
edit Sec 2

@@
 
+% TODO(Amy)
+% - go through all references and confirm used correctly
+
 \section{Introduction}
@@
 
-In this paper we consider the penalised maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter.
-Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross validation) cannot be readily applied in our case.
-We propose utilizing biological replicates to aid tuning, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
+In this paper we consider the penalized maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter.
+Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross validation) cannot be employed.
+We propose using biological replicates to aid tuning.%, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
 We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.
@@
 \label{sec:literature_review}
-\amy{up to here with editing}
+
+\subsection{Poisson mixture models}
 
@@
 Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
-Many mixing distributions have been proposed, including  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
+Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
 While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, a proposal first considered by \citet{fisher_1943}.
@@
 We will focus on focus on frequentist maximum likelihood in this paper.
+
+\subsection{Estimation and computation}
+
 The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.
@@
   \label{eq:binomial_likelihood}\\
- L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k}, \label{eq:conditional_likelihood} % & \text{(Multinomial component/``conditional'')}
+ L_c(\eta) &= \frac{c!}{\displaystyle  \prod_{k \geq 1} f_k!} \prod_{k \geq 1} \left[ \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right]^{f_k}.
+\label{eq:conditional_likelihood} % & \text{(Multinomial component/``conditional'')}
 \end{align}
-i.e., the likelihood is the product of a binomial probability mass function ($c \sim \text{Binomial}\left(C, 1- p_{\eta}(0)\right)$), and a multinomial probability mass function ($(f_1, f_2, \ldots)|c \sim \text{Multinomial}\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$).
+We see that the likelihood is the product of a binomial probability mass function ($c \sim$ Binomial$\left(C, 1- p_{\eta}(0)\right)$), and a multinomial probability mass function ($(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$).
 This decomposition is convenient because $L_c$ is a function of $\eta$ alone.
@@
 \citet{sanathanan_1977}'s conditional approach involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
-
-\citet{sanathanan_1977} uses this decomposition to propose the conditional approach to likelihood maximization, which is done in two steps (see \citet{chao_2002} for a detailed presentation).
-First we maximize $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$.
-Then $L_b(C, \widehat{\eta}_c)$ is maximized over $C$.
-With $\eta$ fixed there is a closed form expression for this second step,
+The conditional estimate of $C$ is then
 \begin{equation}
@@
 The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
+\amy{Need to justify why we are bringing this up}
 
-A perennial issue in the species problem is the instability of estimators, especially maximum likelihood estimators.
-The estimates often have unacceptably high variance and can become dramatically inflated.
-Fisher noted that if there are many low abundance species then $\alpha$ will be near 0.
-The existence of highly abundant species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.
-The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.
+% \citet{sanathanan_1977} uses this decomposition to propose the conditional approach to likelihood maximization, which is done in two steps (see \citet{chao_2002} for a detailed presentation).
+First we maximize $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$.
+Then $L_b(C, \widehat{\eta}_c)$ is maximized over $C$.
+With $\eta$ fixed there is a closed form expression for this second step,
+% \begin{equation}
+% \widehat{C}_c = \Biggl\lfloor \frac{c}{1-p_{\widehat{\eta}_c}(0)} \Biggr\rfloor %\label{eq:conditional_c}
+% \end{equation}
+% where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.
+In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).
+The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
 
-One way of improving stability is to abandon the Poisson mixture paradigm entirely.
+\subsection{Challenges with Poisson mixture models}
+
+A perennial issue in the species problem, especially for microbial datasets, is the instability of species richness estimators \citep{Rocchetti:2011tv,willis_2015}.
+%
+% Under the gamma-Poisson model,
+%
+% To illustrate, consider that a first-order approximation gives $$Var(\hat{C}_c | c) \approx \frac{c^2}{(1-\hat{p}(0))^4}Var(\hat{p}(0))$$ \amy{check this}
+% Therefore,
+% , especially maximum likelihood estimators.  The estimates often have unacceptably high variance and can become dramatically inflated.
+%
+\textcolor{blue}{Since $Pr(X = 0) = \left( \frac{\delta}{1 + \delta}\right)^\alpha$, if there are many low abundance species then $\alpha$ will be near 0.
+Many high abundance species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.
+The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.}
+\amy{Clean up! Find unregularised estimates of alpha and delta for apples}
+
+% One way of improving stability is to abandon the Poisson mixture paradigm entirely.
 Though we will focus on the gamma-Poisson model in this paper, we would be remiss if we did not briefly mention other popular alternatives.
@@
 See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.
+\amy{up to here with editing}
 
-A proposal which retains the gamma-Poisson model and encourages stability is suggested by \citet{wang_2005}.
+A proposal which encourages stability of estimates under the gamma-Poisson model is due to \citet{wang_2005}.
 Their proposal is to add a penalty term to the log likelihood, where the penalization is based on the probability of observing a species zero times (they give three penalty functions and we focus on only the first).

### 2019-07-09 rre_manuscript/paper.tex
edit S2.3

@@
 \usepackage{pstricks}
+\usepackage[normalem]{ulem}
 
@@
 Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
-While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, a proposal first considered by \citet{fisher_1943}.
+While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, sal first considered by \citet{fisher_1943}.
 
@@
 There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model.
-We will focus on focus on frequentist maximum likelihood in this paper.
+We will focus on frequentist maximum likelihood in this paper.
 
@@
 The most straightforward maximum likelihood approach is to simultaneously maximize $C$ and $\eta$ using the full likelihood, which is known as the \textit{direct} approach.
-A complication in this model is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, and so derivative-based optimization methods are not appropriate.
+A complication is that we have continuous parameters $\eta$ as well as one discrete parameter $C$, and so derivative-based optimization methods are not appropriate.
 This issue has been studied by \citet{lindsay_1987}, who provide a discrete analog to the score function for this model.
@@
 \end{align}
-We see that the likelihood is the product of a binomial probability mass function ($c \sim$ Binomial$\left(C, 1- p_{\eta}(0)\right)$), and a multinomial probability mass function ($(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$).
+We see that the likelihood is the product of a binomial probability mass function $c \sim$ Binomial $\left(C, 1- p_{\eta}(0)\right)$, and a multinomial probability mass function $(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$.
 This decomposition is convenient because $L_c$ is a function of $\eta$ alone.
@@
 The multinomial cell probabilities are proportional to $p_{\eta}(k)$.
-\citet{sanathanan_1977}'s conditional approach involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
+\citet{sanathanan_1977}'s conditional approach to optimization involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
 The conditional estimate of $C$ is then
@@
 %
-\textcolor{blue}{Since $Pr(X = 0) = \left( \frac{\delta}{1 + \delta}\right)^\alpha$, if there are many low abundance species then $\alpha$ will be near 0.
+%\sout{\textcolor{blue}{Since $Pr(X = 0) = \left( \frac{\delta}{1 + \delta}\right)^\alpha$, if there are many low abundance species then $\alpha$ will be near 0.
 Many high abundance species, as we see in microbial data, tends to cause $\delta$ to be near 0 as well.
-The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.}
-\amy{Clean up! Find unregularised estimates of alpha and delta for apples}
-
-% One way of improving stability is to abandon the Poisson mixture paradigm entirely.
+The tendency of both abundance parameters to be near their boundary is one way to explain why this instability issue has plagued the species problem over a variety of approaches, and is especially prominent in microbial data.}}
+%% One way of improving stability is to abandon the Poisson mixture paradigm entirely.
 Though we will focus on the gamma-Poisson model in this paper, we would be remiss if we did not briefly mention other popular alternatives.
@@
 See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.
-\amy{up to here with editing}
-
 A proposal which encourages stability of estimates under the gamma-Poisson model is due to \citet{wang_2005}.
-Their proposal is to add a penalty term to the log likelihood, where the penalization is based on the probability of observing a species zero times (they give three penalty functions and we focus on only the first).
-If we let $\ell(C,\eta)$ be the log likelihood, then they define a penalized log likelihood $\ell_{\lambda}(C, \eta)$ as
+Their proposal is to add a penalty term to the log likelihood that penalizes the probability of observing a species zero times.
+They consider the
+penalized log likelihood
 \begin{equation}
@@
 \end{equation}
-where $\lambda > 0$ is a penalization parameter.
-This encourages the maximum likelihood solution for $\eta$ to fit so that $p_{\widehat{\eta}}(0)$ is smaller.
-Therefore $\alpha$ and $\delta$ are further from the boundary, which alleviates some of the instability.
-The penalization term also causes a monotonic decrease in $C$ estimates as greater values of $\lambda$ are used.
-Consequently, if a large enough $\lambda$ value is applied the $C$ estimates will shrink to the observed richness $c$.
+where $\ell(C,\eta) = \log L(C, \eta)$ is the log likelihood and $\lambda > 0$ is a penalization parameter.
+For $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda < \hat{C}_{MLE}$
+\amy{up to here with editing}
+for $\eta$ to fit so that $p_{\widehat{\eta}}(0)$ is smaller.
+\amy{Check this: Therefore $\alpha$ and $\delta$ are further from the boundary, which alleviates some of the instability.} The penalization term also causes a monotonic decrease in $C$ estimates as greater values of $\lambda$ are used.
+For a large enough $\lambda$, $\hat{C}$ will shrink to the observed richness $c$.
+
+\citet{wang_2005} also consider two other penalty functions, which we do not discuss here.
 

### 2019-07-10 rre_manuscript/paper.tex
data analysis results table added

@@
 \ifthenelse {\boolean{DEBUG}}
-{\newcommand{\alex}[1]{{\bf \alexcolor{[\@alex: #1]}}}}
+{\newcommand{\alex}[1]{{\bf \alexcolor{[\@Alex: #1]}}}}
 {\newcommand{\alex}[1]{}}
@@
 \amy{Happy to discuss, but I think it's probably easiest for me to be corresponding author, because you won't have your UW e-mail forever}
+\alex{Having you as the corresponding author sounds great, just didn't want to volunteer you for unwanted emails until we discussed!}
 
@@
 
+\section{Data Analysis}
+\label{sec:data_analysis}
+
+\alex{Data analysis writeup is forthcoming.  For now, here are the results with the same C grid we used in simulations.   I will give more thought into the best way to display these.}
+
+\begin{table}[h]
+\tblcaption{Estimates using a series of three samples drawn in consecutive years from the lakes data.
+Only replicates from the Summer months and the Littoral region of the lake are included in the samples.
+$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
+The number of replicates in each sample is given by $r$.
+\label{tab:data_analysis}}
+\centering
+\begin{tabular}{|r|r|r|r|}
+  \hline
+& 2009  & 2010 & 2011 \\ 
+  \hline
+$\widehat{C}_{[0]}$ & 11200 & 11440 & 14014 \\ 
+  $\widehat{C}_{[1]}$ & 9645 & 11139 & 12246 \\ 
+  $\widehat{C}_{[2]}$ & 10612 & 11120 & 11946 \\ 
+  $\widehat{C}_{[3]}$ & 11200 & 11440 & 14014 \\ 
+  $\widehat{C}_{[4]}$ & 10744 & 11165 & 10328 \\ 
+   \hline
+     $r$ & 8 & 6 & 6 \\
+  \hline
+\end{tabular}
+\end{table}
+
+\noindent \alex{The reason for running an expanded C grid is that many of the estimates in this table are near the boundary of our grid (11,200, 11,440 and 14,300 respectively).
+I'm not concerned that the estimates might be too large/bad, but I am concerned that this may be obscuring differences between the methods.}
+
 \section{Discussion}

### 2019-07-10 data_analysis/figures_tables.R

@@ -1,0 +1,63 @@
+library(magrittr)
+library(rre)
+
+get_n <- function(list_of_samples) {
+  get_n_one_sample <- function(samp) samp[[1]] %>% length
+  lapply(list_of_samples, get_n_one_sample) %>% unlist
+}
+
+get_n(lakes_results)
+
+get_estimate <- function(result) {
+  best <- result %>% '[['(.,"best") 
+  if ("ccc" %in% names(best)) {
+    best[["ccc"]]
+  } else if ("ccc_hat" %in% names(best)) {
+    best[["ccc_hat"]]
+  } else {
+    stop("Estimate not found for this method")
+  }
+}
+
+process_sample <- function(list_of_outputs) {
+  list_of_outputs %<>% '['(., 2:6) # the first list item is just the data, so 2-6 are the 5 methods
+  lapply(list_of_outputs, get_estimate) %>% unlist
+}
+
+process_sample_list <- function(list_of_sample_outs, 
+                                sample_names = NA,
+                                latex_format = T) {
+  table_output <- lapply(list_of_sample_outs, process_sample) %>%
+    do.call(cbind,.) 
+  if (any(!is.na(sample_names))) {
+    colnames(table_output) <- sample_names
+  }
+  
+  if (latex_format) {
+    r <- get_n(lakes_results)
+    table_output %<>% rbind(.,r)
+    rownames(table_output) <- c( paste0("$\\widehat{C}_{[", 0:4, "]}$"),  "$r$")
+    table_output %>% xtable::xtable(digits = 0) %>% print(
+      sanitize.text.function = identity,
+      sanitize.rownames.function = identity,
+      santitize.colnames.function = identity
+    )
+  } else { # console output
+    table_output
+  }
+}
+
+
+setwd("./data_analysis") 
+load("data_analysis_results.RData")
+process_sample_list(lakes_results, 
+                    sample_names = 2009:2011,
+                    latex_format = T)
+
+# Temporary while running C grid sensitivity analysis:
+get_cc <- function(list_of_results) {
+  one_sample <- function(samp) samp[[1]] %>% rre::get_cc_max(.)
+  lapply(list_of_results, one_sample) %>% unlist
+}
+
+get_cc(lakes_results) * 20 # extent of grid

### 2019-07-17 data_analysis/data_analysis_comparison.Rmd
data_analysis_update

---
title: "Data Analysis Comparison"
author: "Alex Paynter"
date: "7/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(magrittr)
source("./figures_tables.R")
```

## Introduction

In this document we show the data analysis results using two choices of $C$ grid.  We first used the small grid and noticed that the $\widehat{C}$ estimates were near the upper bound of the grid for many of the methods.  This motivated the use of a larger grid to see if the results were the same.

#

## Small $C$ grid.

Let $c$ be the maximum observed richness of a sample (maximum over the replciates).  In the small $C$ grid simulation we iterate our optimization over $C = \{c, 1.2*c, 1.4*c, \dots, 20*c\}$, or in other words we ignore the possibility that the coverage of our sample could be under 5\%.  This is the same grid we used in all simulations.  After running our algorithm with all methods we find

```{r small_out}
process_sample_list(lakes_results, 
                    sample_names = 2009:2011,
                    latex_format = F)

```

Given that the maximum observed richness in the samples is `r get_cc(lakes_results)` this caps our $C$ estimates at `r formatC(get_cc(lakes_results) * 20, format = "d")` respecitvely.  Therefore we have many estimates which are near the arbitrarily chosen upper bound of our $C$ grid.

## Large $C$ grid

In the large $C$ grid we iterate over $C = \{c, 1.5*c, 2*c, \dots, 50*c\}$.

# Common settings

The following settings are consistent in both analyses:

- The data is a series of three samples from the littoral region in the Summer: 2009, 2010 and 2011.  These have $r$ = 8,6,6 (replicates) respectively.
- The $\lambda^{\text{grid}}$ is $\{0, 10, 20, \dots , 140\}$ which is the largest grid we used in any simulation.
- Optimization starts for $\eta = (\alpha, \delta)$ include two fixed starts and one borrowed from the maximum likelihood solution in the previous $C$.

### 2019-07-17 data_analysis/data_analysis.R

@@ -1,13 +1,14 @@
 library(magrittr)
 library(rre)
-setwd("/Users/alex/Dropbox/blups/code/rre_sims/")
+setwd("/Users/alex/Dropbox/blups/code/rre_sims/data_analysis/")
+
 lakes <- readr::read_tsv("./lakes_data.txt") %>% as.data.frame()
 lakes_metadata <- readr::read_tsv("./lakes_metadata.txt") %>% as.data.frame()
 
 # Deciding which samples to extract:
 lakes_metadata %>% 
   dplyr::group_by(Site, Years, Period) %>%
-  dplyr::summarize(number = n()) %>%
+  dplyr::summarize(number = dplyr::n()) %>%
   dplyr::arrange(Site, Period, Years) %>%
   print(n = 500)
 
@@ -39,29 +40,46 @@ get_samples <- function(yr, zone, per = NULL, month = NULL) {
 # Note that one of the metadata files matches no lakes sample ID:  12.07.11.1
 list_of_samples <- mapply(get_samples, seq(2009, 2011), "Littoral", "Summer")
 
-run_methods <- function(fct_sample, lambda_grid = seq(0,140,by=10)) {
+run_methods <- function(fct_sample, lambda_grid = seq(0,140,by=10),
+                        multiplier = 20, # default for all methods.
+                        c_seq_len = 96) {
   fct_sample %>%
-    list(method_0 = rre::unregularized_mle(.),
+    list(method_0 = rre::unregularized_mle(., multiplier = multiplier,
+                                           c_seq_len = c_seq_len),
          method_1 = rre::minimum_subset_distance(.,
                                                  lambda_vec = lambda_grid,
-                                                 partitions = 10),
+                                                 partitions = 10,
+                                                 multiplier = multiplier,
+                                                 c_seq_len = c_seq_len),
          method_2 = rre::cv_replicates(.,  eval_function = "neg_unreg_like",
                                        lambda_vec = lambda_grid,
-                                       partitions = 10),
-         method_3 = rre::gof_criterion(., lambda_vec = lambda_grid),
+                                       partitions = 10,
+                                       multiplier = multiplier,
+                                       c_seq_len = c_seq_len),
+         method_3 = rre::gof_criterion(., lambda_vec = lambda_grid,
+                                       multiplier = multiplier,
+                                       c_seq_len = c_seq_len),
          method_4 = rre::cv_replicates(.,  eval_function = "gof_chi_sq",
[…]

### 2019-07-17 data_analysis/figures_tables.R

@@ -6,7 +6,11 @@ get_n <- function(list_of_samples) {
   lapply(list_of_samples, get_n_one_sample) %>% unlist
 }
 
-get_n(lakes_results)
+# Temporary while running C grid sensitivity analysis:
+get_cc <- function(list_of_results) {
+  one_sample <- function(samp) samp[[1]] %>% rre::get_cc_max(.)
+  lapply(list_of_results, one_sample) %>% unlist
+}
 
 get_estimate <- function(result) {
   best <- result %>% '[['(.,"best") 
@@ -48,16 +52,16 @@ process_sample_list <- function(list_of_sample_outs,
 }
 
 
-setwd("./data_analysis") 
 load("data_analysis_results.RData")
-process_sample_list(lakes_results, 
-                    sample_names = 2009:2011,
-                    latex_format = T)
+load("data_analysis_results_big_C.RData")
 
-# Temporary while running C grid sensitivity analysis:
-get_cc <- function(list_of_results) {
-  one_sample <- function(samp) samp[[1]] %>% rre::get_cc_max(.)
-  lapply(list_of_results, one_sample) %>% unlist
-}
+# latex table outputs:
 
-get_cc(lakes_results) * 20 # extent of grid
+
+#process_sample_list(lakes_results, 
+#                    sample_names = 2009:2011,
+#                    latex_format = T)
+
+#process_sample_list(lakes_results, 
+#                    sample_names = 2009:2011,
+#                    latex_format = T)

### 2019-07-17 data_analysis/data_analysis_comparison.Rmd
large C grid analysis complete

@@
 
-In this document we show the data analysis results using two choices of $C$ grid.
-We first used the small grid and noticed that the $\widehat{C}$ estimates were near the upper bound of the grid for many of the methods.
+In this document we show the data analysis results using two choices for $C$ grid.
+The small grid was our pre-specified plan, and we noticed that the $\widehat{C}$ estimates were near the upper bound of the grid for many of the methods.
 This motivated the use of a larger grid to see if the results were the same.
@@
 Let $c$ be the maximum observed richness of a sample (maximum over the replciates).
-In the small $C$ grid simulation we iterate our optimization over $C = \{c, 1.2*c, 1.4*c, \dots, 20*c\}$, or in other words we ignore the possibility that the coverage of our sample could be under 5\%.
+In the small $C$ grid simulation we iterate our optimization over $C = \{c, 1.2*c, 1.4*c, \dots, 20*c\}$, or in other words we set a lower bound for the coverage of our sample at 5\%.
 This is the same grid we used in all simulations.
@@
 ```{r small_out}
-process_sample_list(lakes_results, 
+small_out <- process_sample_list(lakes_results,
                                  sample_names = 2009:2011,
                                  latex_format = F)
+small_out
+```
+Here is a very preliminary plot, with the points slightly horizontally jittered:
+```{r small_plot}
+library(ggplot2)
+# Room for improvement here:
+plot_fn <- function(output_mat) {
+  df <- output_mat %>%
+    as.data.frame %>%
+    tibble::rownames_to_column(.) %>%
+    tidyr::gather(., key = "year", value = "estimate", -rowname)
 
+  # cleaning up names:
+  df["method"] <-df[["rowname"]] %>%
+    stringr::str_replace(., "_", " ") %>%
+    stringr::str_replace(., "m", "M") %>%
+    factor(., levels = unique(.))
+
+  pd <- position_dodge(0.1)
+  ggplot(df, aes(x = year, y = estimate, group = method, color = method)) +
+    geom_point(position = pd) + 
+    geom_line(position = pd) +
+    labs(y = "C estimate", x = "year")
+}
+plot_fn(small_out)
 ```
 
-Given that the maximum observed richness in the samples is `r get_cc(lakes_results)` this caps our $C$ estimates at `r formatC(get_cc(lakes_results) * 20, format = "d")` respecitvely.
+Given that the maximum observed richness in 2009, 2010, 2011 is `r get_cc(lakes_results)`,  this caps our $C$ estimates at `r formatC(get_cc(lakes_results) * 20, format = "d")` respecitvely.
 Therefore we have many estimates which are near the arbitrarily chosen upper bound of our $C$ grid.
@@
 
-In the large $C$ grid we iterate over $C = \{c, 1.5*c, 2*c, \dots, 50*c\}$.
+In the large $C$ grid we iterate over $C = \{c, 1.5*c, 2*c, \dots, 50*c\}$, and the results are
 
-# Common settings
+```{r big_out}
+big_out <- process_sample_list(lakes_results_bigger_C_grid, 
+                    sample_names = 2009:2011,
+                    latex_format = F)
+big_out
 
-The following settings are consistent in both analyses:
+```
 
-- The data is a series of three samples from the littoral region in the Summer: 2009, 2010 and 2011.  These have $r$ = 8,6,6 (replicates) respectively.
-- The $\lambda^{\text{grid}}$ is $\{0, 10, 20, \dots , 140\}$ which is the largest grid we used in any simulation.
-- Optimization starts for $\eta = (\alpha, \delta)$ include two fixed starts and one borrowed from the maximum likelihood solution in the previous $C$.
+Here is a very preliminary plot, with the points slightly horizontally jittered:
+```{r big_plot}
+library(ggplot2)
+# Room for improvement here:
+plot_fn <- function(output_mat) {
+  df <- output_mat %>%
+    as.data.frame %>%
+    tibble::rownames_to_column(.) %>%
+    tidyr::gather(., key = "year", value = "estimate", -rowname)
+
+  # cleaning up names:
+  df["method"] <-df[["rowname"]] %>%
+    stringr::str_replace(., "_", " ") %>%
+    stringr::str_replace(., "m", "M") %>%
+    factor(., levels = unique(.))
+
+  pd <- position_dodge(0.1)
+  ggplot(df, aes(x = year, y = estimate, group = method, color = method)) +
+    geom_point(position = pd) + 
+    geom_line(position = pd) +
+    labs(y = "C estimate", x = "year")
+}
+plot_fn(big_out)
+```
+
+Our results for 2009, 2010, 2011 under this analysis are capped at `r formatC(get_cc(lakes_results) * 50, format = "d")`.  Unfortunately many of the results are still very close to the upper bound.
+
+## Conclusions
+
+We don't learn anything new by expanding the $C$ grid.
+We might learn more by tinkering with the $\lambda^{\text{grid}}$, but I'm inclined to just report the results of the small grid which is what we did in our simulations.
+We can discuss the difficulty in selecting of $\lambda^{\text{grid}}$ as a limitation.

### 2019-07-17 data_analysis/data_analysis.R

@@ -71,7 +71,7 @@ save(lakes_results,
      file = "./data_analysis_results.RData")
 lakes_results_bigger_C_grid <- lapply(list_of_samples, run_methods,
                                       multiplier = 50,
-                                      c_seq_len = 246)
+                                      c_seq_len = 99)
 save(lakes_results_bigger_C_grid,
      file = "./data_analysis_results_big_C.RData")
 

### 2019-07-17 data_analysis/figures_tables.R

@@ -32,7 +32,7 @@ process_sample_list <- function(list_of_sample_outs,
                                 sample_names = NA,
                                 latex_format = T) {
   table_output <- lapply(list_of_sample_outs, process_sample) %>%
-    do.call(cbind,.) 
+    do.call(cbind,.)
   if (any(!is.na(sample_names))) {
     colnames(table_output) <- sample_names
   }
@@ -47,7 +47,7 @@ process_sample_list <- function(list_of_sample_outs,
       santitize.colnames.function = identity
     )
   } else { # console output
-    table_output
+    ceiling(table_output)
   }
 }
 

### 2019-07-17 data_analysis/data_analysis.R
cleanup old lines

@@ -74,14 +74,3 @@ lakes_results_bigger_C_grid <- lapply(list_of_samples, run_methods,
                                       c_seq_len = 99)
 save(lakes_results_bigger_C_grid,
      file = "./data_analysis_results_big_C.RData")
-
-
-# apples_results <- breakaway::apples %>%
-#  list(.) %>%
-#  list(method_0 = rre::unregularized_mle(.),
-#        method_3 = rre::gof_criterion(., lambda_vec = lambda_grid))
-
-lakes_results %>% lapply(. (function(y) lapply(y, function(x) x$best )))
-
-lapply(list_of_samples, rre::get_cc_max)
-756*20

### 2019-07-31 rre_manuscript/paper.tex
data analysis first draft

@@
 
-\alex{Data analysis writeup is forthcoming.  For now, here are the results with the same C grid we used in simulations.   I will give more thought into the best way to display these.}
+To illustrate the performance of these methods on real data we will use samples collected in consecutive years from Lake Champlain (\citet{tromas_2017}).
+We limit our samples to replicates from the summer months and the littoral zone of the lake  (near the shore).
+There were 8 replicates meeting these criteria from 2009, 6 replicates from 2010 and 6 from 2011.
+The maximum observed richness (most species observed in any replicate that year) was 560 in 2009, 572 in 2010 and 715 in 2011.
 
-\begin{table}[h]
-\tblcaption{Estimates using a series of three samples drawn in consecutive years from the lakes data.
-Only replicates from the Summer months and the Littoral region of the lake are included in the samples.
+The richness estimates from each method are displayed in Table \ref{tab:data_analysis}.
+Our estimates from each method are at least 10 times as high as the observed richness, and in some cases they reach the maximum value in our grid search for $C$ (which is 20 times the observed richness).
+These extreme estimates suggest that the Gamma-Poisson model which underlies all methods is a poor fit to these data.
+In simulations penalization had the intended effect of lowering inflated estimates.
+However, we always simulated under a Gamma-Poisson model in simulations.
+Using wild data our methods still produce estimates which are suspiciously high.
+
+\begin{table}[t]
+\tblcaption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
 $\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
@@
 
-\noindent \alex{The reason for running an expanded C grid is that many of the estimates in this table are near the boundary of our grid (11,200, 11,440 and 14,300 respectively).
-I'm not concerned that the estimates might be too large/bad, but I am concerned that this may be obscuring differences between the methods.}
+In Figure \ref{fig:data_analysis} we display the same results visually.
+The methods which use partitioning (Methods 1, 2 and 4) consistently produce lower estimates than those which do not (Methods 0 and 3).
+This trend was not found in simulations, where the highest estimates came from Method 2 and the lowest from Method 1.
+This may suggest that partitioning methods have an advantage on real data which could have greater heterogeneity between replicates than data generated from simple parametric distributions.
+
+\begin{figure}[t]
+\caption{A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.  
+\label{fig:data_analysis}}
+\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/data_analysis.pdf}}
+\end{figure}
+
+\noindent \alex{Next: Modify the discussion to harmonious with what you've written here.}
 
@@
 Our goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
-In this section we note promising directions for future work.
+In this section we note promising directions for future work.{}
 

### 2019-07-31 data_analysis/data_analysis.R

@@ -1,6 +1,6 @@
 library(magrittr)
 library(rre)
-setwd("/Users/alex/Dropbox/blups/code/rre_sims/data_analysis/")
+setwd("/Users/alex/Dropbox/classGrad/IS_Wil/rre_sims/data_analysis/")
 
 lakes <- readr::read_tsv("./lakes_data.txt") %>% as.data.frame()
 lakes_metadata <- readr::read_tsv("./lakes_metadata.txt") %>% as.data.frame()

### 2019-07-31 data_analysis/figures_tables.R

@@ -1,5 +1,6 @@
 library(magrittr)
 library(rre)
+library(ggplot2)
 
 get_n <- function(list_of_samples) {
   get_n_one_sample <- function(samp) samp[[1]] %>% length
@@ -51,17 +52,67 @@ process_sample_list <- function(list_of_sample_outs,
   }
 }
 
+plot_fn <- function(output_mat, 
+                    output_name = NA,
+                    output_height = 4,
+                    output_width = 7) {
+  # match method names to simulations:
+  rownames(output_mat) <- c("[0] MLE", # method 00
+                    "[1] Min Var", # method 01
+                    "[2] CV likelihood", # method 02
+                    "[3] GOF", # method 03
+                    "[4] CV GOF") # method 04
+  
+  df <- output_mat %>%
+    as.data.frame %>%
+    tibble::rownames_to_column(.) %>%
+    tidyr::gather(., key = "year", value = "estimate", -rowname) %>%
+    dplyr::rename(method = rowname)
+  
+  print(df)
+  
+  pd <- position_dodge(0.1)
+  gg <- ggplot(df, aes(x = year, y = estimate, group = method, color = method)) +
+    geom_point(position = pd) + 
+    geom_line(position = pd) +
+    labs(y = latex2exp::TeX("$\\hat{C}$"), x = "Year") + 
+    scale_color_viridis_d(option = "viridis", begin = 0.15, end = 0.7125) +
+    theme_bw() +
+    theme(axis.text.x = element_text(color = "grey20", size = 10, angle = 0, hjust = .5, vjust = .5, face = "plain"),
+          axis.text.y = element_text(color = "grey20", size = 12, angle = 0, face = "plain"),  
+          axis.title.x=element_blank(),
+          axis.title.y = element_text(color = "grey20", size = 12, angle = 90, hjust = .5, vjust = .5, face = "plain"),
+          #strip.placement = "outside",
+          legend.position="bottom") 
+
+  
+  if (!is.na(output_name)) {
+    ggsave(filename = output_name, plot = gg,
+           height = output_height, width = output_width)
+  } else {
+    gg
+  }
+}
+
+
+
+
+setwd("/Users/alex/Dropbox/classGrad/IS_Wil/rre_sims/data_analysis/")
 
 load("data_analysis_results.RData")
-load("data_analysis_results_big_C.RData")
+#load("data_analysis_results_big_C.RData")
 
 # latex table outputs:
-
-
-#process_sample_list(lakes_results, 
-#                    sample_names = 2009:2011,
[…]

### 2019-07-31 rre_manuscript/paper.tex
edit sec 3, 4, 5.1

@@
 
-\amy{Happy to discuss, but I think it's probably easiest for me to be corresponding author, because you won't have your UW e-mail forever}
-\alex{Having you as the corresponding author sounds great, just didn't want to volunteer you for unwanted emails until we discussed!}
-
 % TODO(Amy)
@@
 In this paper we consider the penalized maximum likelihood approach of \citet{wang_2005} and investigate the open problem of selecting the penalization parameter.
-Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross validation) cannot be employed.
-We propose using biological replicates to aid tuning.%, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
+Because the true species richness $C$ is never observed for any sample, common approaches to tuning parameter selection (e.g., cross-validation) cannot be employed.
+We propose using biological replicates to aid tuning.
+%, and so we extend the proposal of \citet{wang_2005} to the analysis of independent but homogeneous ecosystems.
 We find that replicate data is advantageous for tuning the required penalization parameter and provide a comparison of the performance of 4 proposed methods for tuning parameter selection.
@@
 Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
-While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, sal first considered by \citet{fisher_1943}.
+While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{fisher_1943}.
 
@@
 We will focus on frequentist maximum likelihood in this paper.
+\amy{introduce Gamma later?}
 
@@
 where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.
-In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).
+In the \textit{profile} likelihood approach the expression in Equation \eqref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).
 The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
@@
 % where $\lfloor a \rfloor$ is the largest integer less than or equal to $a$.
-In the \textit{profile} likelihood approach the expression in Equation \ref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).
+In the \textit{profile} likelihood approach the expression in Equation \eqref{eq:conditional_c} is substituted into the full likelihood, which gives us a function of $\eta$ alone (see \citet{wang_2005}).
 The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
@@
 See \citet{bunge_1993} for a comprehensive review and \citet{bunge_2014} for an updated review of species richness methods which are outside the scope of this paper.
-A proposal which encourages stability of estimates under the gamma-Poisson model is due to \citet{wang_2005}.
+A proposal which encourages stability of estimates under Poisson-mixture models is due to \citet{wang_2005}.
 Their proposal is to add a penalty term to the log likelihood that penalizes the probability of observing a species zero times.
@@
 where $\ell(C,\eta) = \log L(C, \eta)$ is the log likelihood and $\lambda > 0$ is a penalization parameter.
-For $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda < \hat{C}_{MLE}$
-\amy{up to here with editing}
-for $\eta$ to fit so that $p_{\widehat{\eta}}(0)$ is smaller.
-\amy{Check this: Therefore $\alpha$ and $\delta$ are further from the boundary, which alleviates some of the instability.} The penalization term also causes a monotonic decrease in $C$ estimates as greater values of $\lambda$ are used.
-For a large enough $\lambda$, $\hat{C}$ will shrink to the observed richness $c$.
+Let $\hat{C}_\lambda = \arg \max_{C} \ell_\lambda(C, \eta)$.
+Then, for $\lambda \geq \lambda',$ $\hat{C}_\lambda \leq \hat{C}_{\lambda'}$ \citep[Theorem 1]{wang_2005}.
+In particular, for $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda$ is less than or equal to $\hat{C}_0$, the maximum likelihood estimate.
+Furthermore, for a large enough $\lambda$, $\hat{C}_\lambda = c$; that is, the penalised maximum likelihood estimate shrinks to the observed richness $c$.
+\citet{wang_2005} also consider two other penalty functions of similar form, which we do not discuss here except to note that our tuning parameter selection methods would equally apply.
 
-\citet{wang_2005} also consider two other penalty functions, which we do not discuss here.
-
-In their paper they show a trade-off exists: greater values of $\lambda$ correspond to a more stable estimator, but at the potential cost of negative bias.
-This is a realization of the bias-variance tradeoff.
-The choice of penalization parameter implies a preference for lower variance or lower bias, and therefore adds a degree of subjectivity to the estimation procedure (comparable to a Bayesian prior as discussed in \citet{wang_2005}).
-We would also expect different data sets to require different $\lambda$ values for optimal estimation, which makes direct application of this method to real data challenging.
-Our primary interest in studying the analysis of replicate frequency count tables is to leverage the replicates to tune the penalization parameter in Equation \ref{eq:wang_lindsay}.
-The goal of this exploration is to obtain an estimator which is more stable than the unpenalized MLE but does not have the disadvantage of requiring a fixed, user-specified penalization parameter.
+\citet{wang_2005} show that a trade-off exists: greater values of $\lambda$ correspond to a more stable estimator, but at the potential cost of negative bias.
+The choice of penalization parameter implies a preference for lower variance or lower bias, adding subjectivity to the estimation procedure.
+\citet{wang_2005} note that a limitation of their proposal is that ``one must select a penalty function and a tuning parameter, and it is nigh impossible to make convincing statements about why one choice should be uniformly superior to another.''
+Furthermore, we expect different data sets to require different $\lambda$ values for optimal mean squared error.
+% Our primary interest in studying the analysis of replicate frequency count tables is to leverage the replicates to tune the penalization parameter in Equation \eqref{eq:wang_lindsay}.
+The goal of this paper is to propose and investigate data-adaptive methods to select $\lambda$.
 
@@
 
-Throughout we will focus on the gamma-Poisson model with penalization as proposed by \citet{wang_2005}.
+We focus on the gamma-Poisson model and penalised log-likelihood of the form \eqref{eq:wang_lindsay}, and consider the case where we observe $r$ independent frequency count tables from the population under study.
+% with penalization as proposed by \citet{wang_2005}.
 In this section we extend the penalized likelihood for a single frequency count table to the case where multiple frequency count tables have been collected.
-
-Let $r$ be the number of frequency count tables.
 We will assume that these $r$ frequency count tables are biological replicates drawn independently from the same population, i.e.
@@
 
-Microbial data is often available as labeled abundance information $\{ X_{ij} \}$, the number of times species $i$ was observed in replicate $j$.
+% Microbial data is often available as labeled abundance information $\{ X_{ij} \}$, the number of times species $i$ was observed in replicate $j$.
 In this paper we study data represented in frequency counts $\{f_{kj}\}$ and it is worth pointing out that we lose some information in this representation.
@@
 Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.
-We will refer to this as the \textit{penalized log likelihood} and it is given by
+We define the \textit{penalized log likelihood} for multiple samples as
 \begin{align}
-\mathcal{O}_{\lambda}(C, \eta; J) =& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
+\mathcal{O}_{\lambda}(C, \eta; J) :=& \sum_{j \in J} \ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right) - \lambda\log p_{\eta}(0) \\
  =& \sum_{j \in J} \biggl[ \log C! - \log (C-c)! + (C-c) \log p_\eta(0)  \nonumber \\
@@
 By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.
-We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate using $\lambda$, that is
+We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate for based on tuning parameter $\lambda$, that is,
 \begin{equation}
@@
 \end{equation}
-where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.  $\eta$ is critical to estimation but we are largely concerned with estimating $C$ in the species problem.
+where $\widehat{\eta}_{\lambda}$ is a nuisance parameter.
 
 We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.
-Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \ref{eq:ccc_hat_lambda} obtained in simulation $s$ with  $\lambda$.
+Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.
 Then we define
@@
 To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.
-While it may be faster and asymptotically equivalent to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.
-In this paper our end goal is to explore tuning, so the stability and accuracy of our underlying optimization algorithm have priority over speed.
-Crucially, will use exactly the same optimization for the likelihood throughout all simulations in this paper.
+While it may be faster to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.
+Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed.
 
-To implement a direct optimization approach we search over candidate $C$ values in a grid, respecting the fact that $C$ is a discrete parameter.
-A gradient-based search over $\eta$ values is used at each $C$ in the grid.
-The highest likelihood evaluation for any $(C, \eta)$ is our maximum likelihood estimate.
-While this paper does not focus on optimization, our code is made available to the reader (see Section \ref{sec:code} for links).
+To implement a direct optimization approach we search over candidate $C$ values in a grid.
+A gradient-based search over $\eta$ values is used at each $C$ in the grid to find the maximum penalised likelihood  $(C, \eta)$.
 
@@
 
-Establishing that penalization can improve richness estimates is a crucial foundation before we attempt to evaluate tuning methods.
-\citet{wang_2005} established that it can improve estimates for $r = 1$ and in their particular application area.
-We need to show this for our models and $r > 1$.
-We propose to do so by simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.
-If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) over all simulations is less than RMSE($\widehat{C}_0)$ we conclude that penalization is effective under those simulation parameters.
+Before evaluating potential tuning parameter selection methods, we first establish that penalization can improve richness estimation.
+\citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.
+We propose simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.
+If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective for the given data generating mechanism.
 
-We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$ to reflect the reality that few microbiome-focused labs are currently producing a large number of biological replicates.
-The abundance distribution of a population is characterized by $\eta$, so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.
+We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$.
+We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017}.
+For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
+% To select $\eta$, we investigated the unregularised
+%  two microbial datasets
+% so choosing appropriate simulation values of $\eta$ is crucial in targeting our approach to microbial data.
 We used two microbial datasets as motivating examples.
@@
 Under each combination of $\eta$ and $r$ 100 simulations were completed.
-
 In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.
 This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.
-We make this information available here for the interested reader but delay our discussion of the differences until Section \ref{sec:tuning_simulations} where they strongly influence our results.
+We return to a discussion of the different data structures in Section \ref{sec:tuning_simulations}.
+For each combination of $\eta$ and $r$, we performed 100 simulations.
+We investigated the grid $\lambda \in \{0, 5, \dots , 120\}$, and found that this grid was sufficiently expansive to ensure that the RMSE-minimising $\lambda$ was not on the boundary of the grid
+(see Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}).
+
+% One final setting is necessary for these simulations: a set of $\lambda$ values to test over.
+Our primary concern is to ensure the set covers the optimal value, i.e.
+$\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.
+While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to confirm that the set was sufficiently large after the simulation is complete.
+As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.
+If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ $\widehat{C}_{\lambda}$ will be further from $C$, and the RMSE will increase.
+Then we can be assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.
+% This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.
+Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.
 
@@
 \begin{table}[t]
-\tblcaption{Summaries of the proportion of species which are unobserved, observed once (singletons), rarely observed and abundantly observed are calculated for each choice of $\eta$.
-We give the expected maximum abundance $k_{\text{max}}$, the number of times the most abundant species was observed.
-In this table only $\mathbb{E}(k_{\text{max}})$ depends on $C$, and  $C = 1000$ is used here.
+\tblcaption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.
+We also give the expected frequency count of the most abundant species when $C = 1000$.
 \label{tab:eta_intuition}}
@@
 \end{table}
-One final setting is necessary for these simulations: a set of $\lambda$ values to test over.
-Our primary concern is to ensure the set covers the optimal value, i.e.
-$\lambda$ which produces the lowest RMSE$(\widehat{C}_{\lambda})$.
-While we do not have a theoretical way to determine the ideal $\lambda$ set, we can use the known behavior of the penalty function to confirm that the set was sufficiently large after the simulation is complete.
-As $\lambda$ increases, $\widehat{C}_{\lambda}$ decreases.
-If $\widehat{C}_{\lambda_0}$ is  below the true $C$, then for $\lambda > \lambda_0$ $\widehat{C}_{\lambda}$ will be further from $C$, and the RMSE will increase.
-Then we can be assured that the set is large enough if most or all simulations had a $\widehat{C}_{\lambda}$ below the truth for the largest $\lambda$ tested.
-This is clearly verifiable in the cases presented in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.
-Simulating using $\lambda \in \{0, 5, \dots , 120\}$ was sufficient for all simulations in this section.
 
 \begin{table}[h]
-\tblcaption{Simulation results for various abundance ($\eta$) and replicate ($r$) choices.
-For each row in the table 100 simulations were run, e.g.
-for the first row we generated a group of 6 frequency count tables under $C = 1000, \eta = (10^{-1},10^{-1})$ one hundred times.
-$\lambda_{\text{opt}}$ is the optimal value of $\lambda$, i.e.
-$\lambda$ which produced the lowest RMSE over all simulations.
+\tblcaption{The penalised maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
+$\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
 $\widehat{C}_0$ is the $C$ estimate with $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the $C$ estimate using $\lambda = \lambda_{\text{opt}}$.
@@
 
-\subsection{Simulation results}
+\subsection{Simulation results} \label{sec:first_sim}
 
-The simulation parameters and RMSE results are shown in Table \ref{tab:fixed_lambda_results}.
-For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is attained for some $\lambda > 0$.
+The results of the simulation are shown in Table \ref{tab:fixed_lambda_results}.
+For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is found for some $\lambda > 0$.
 Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.
-These results show that for a variety of $(\eta, r)$ choices penalization improves species richness estimates.
-In other words, if we knew the optimal $\lambda$ value we could improve our richness estimates using penalization with that $\lambda$ applied.
-This extends the results of \citet{wang_2005} to the analysis of replicate frequency count tables and our particular application.
-We also note that depending on the simulation settings the optimal $\lambda$ value varies considerably, ranging from 10 to 70.
-This simulation suggests that there is not a universally appropriate $\lambda$ choice.
-Therefore, we have evidence that tuning the $\lambda$ value to the sample is highly desirable.
+These results show that for a variety of $(\eta, r)$ choices, penalization improves species richness estimation.
+We also see that the optimal $\lambda$ value varies considerably over $(\eta, r)$, ranging from 10 to 70.
+This suggests that there is not a universally appropriate $\lambda$ choice.
+Therefore, tuning $\lambda$ to the sample appears desirable.
+We note that the optimal choice of $\lambda$ is not consistent either for fixed $r$ and variable $\eta$, nor for fixed $\eta$ and variable $r$.
 
@@
 Therefore a method which can tune the parameter based on a sample is warranted.
-We propose several novel and intuitive methods in this section.
-Each is evaluated in simulations in Section \ref{sec:tuning_simulations}.
+We propose several novel methods in this section.
+Each is evaluated via simulation in Section \ref{sec:tuning_simulations}.
 
@@
 \subsection{Method 0: MLE without penalization}
-We begin by proposing a reasonable baseline to compare our tuning methods with.
- The easiest extension to the replicate case is to maximize the unpenalized likelihood using $J = \{1, \dots r \}$.
-Recall that fixing $\lambda = 0$ gives us the unpenalized likelihood in Equation \ref{eq:objective}, so if we denote the $C$ estimate obtained using Method 0 as $\widehat{C}_{[0]}$, then
+
+We compare all proposed methods to the unpenalised MLE using $J = \{1, \dots r \}$:
+%Recall that fixing $\lambda = 0$ gives us the unpenalized likelihood in Equation \eqref{eq:objective}, and so by denoting the $C$ estimate obtained using Method 0 as $\widehat{C}_{[0]}$, then
 \begin{equation}
@@
 \end{equation}
-This method is faster and makes fewer assumptions than any other method, so it serves as the standard to beat.
+This method is fast and simple, making it an ideal baseline for comparison
 
-\ \\
-
-For all of the remaining methods (1-4) we generate estimates $\widehat{C}_\lambda$ over $\lambda \in \lambda^{\text{grid}}$, where $\lambdagrid$ is a user specified constant required to run each method (e.g.
-$\lambdagrid = \{0, 5, 10, \dots 60\}$ in Section \ref{sec:tuning_simulation_1}).
-The methods differ from one another in how they subset the data to create these estimates, and in the steps they take to arrive at a final estimate $\widehat{C}_{\text{[method]}}$.
+For all of the remaining methods (1-4) we generate estimates $\widehat{C}_\lambda$ over $\lambda \in \lambda^{\text{grid}}$, where $\lambdagrid$ is user-specified (e.g., $\lambdagrid = \{0, 5, 10, \dots 120\}$ in Section \ref{sec:first_sim}).
+%The methods differ in how they subset the data to create these estimates, and in the steps they take to arrive at a final estimate $\widehat{C}_{\text{[method]}}$.
 
@@
 \subsection{Method 1: Minimum subset variance}
-Method 1 is motivated by the belief that if we resample from the same population, an ideal $C$ estimator should have low variance.
-Exploiting the fact that we have replicate data, the idea is to repeatedly partition the replicates into two subsets and come up with two estimates.
-We select the $\lambda$ which yields the lowest between-subset variance.
+
+Ideally, our estimator for $C$ will have low variance, that is, there should be low variance in estimates from different resamples from the population.
+% Method 1 is motivated by the belief that if we resample from the same population, an ideal $C$ estimator should have low variance.
+Exploiting the fact that we have replicate data, we repeatedly partition the replicates into two subsets and calculate two estimates.
+We then select the $\lambda$ which yields the lowest between-subset variance.
 This partitioning is repeated $p$ times to average out the arbitrary choice of subsets.
 
-Let $T_1(l)$ be the first subset of the $l^{\text{th}}$ partition and $T_2(l)$ be the second subset of the $l^{\text{th}}$ partition, for $l \in \{1, \dots , p\}$ ($T$ is for training).
-We select the subsets randomly and we are using the word partition in the mathematical sense.
-So $T_1(l), T_2(l) \subseteq \{1, \dots , r\}$, $T_1(l) \cap T_2(l) = \emptyset$ and $T_1(l) \cup T_2(l) = \{1, \dots ,r\}$.
+Let $T_1(l)$ be the first subset of the $l^{\text{th}}$ partition and $T_2(l)$ be the second subset of the $l^{\text{th}}$ partition, for $l \in \{1, \dots , p\}$.
+That is, $T_i(l) \subseteq \{1, \dots , r\}$, $T_1(l) \cap T_2(l) = \emptyset$ and $T_1(l) \cup T_2(l) = \{1, \dots ,r\}$.
 For each $\lambda \in \lambda^{\text{grid}}$, $l \in \{1, \dots, p\}$ let
@@
 Our goal in Method 1 is to use these estimates to select the $\lambda$ value which gave us the lowest average variance over all partitions, denoted by $\widetilde{\lambda}_{[1]}$.
-The overall estimate for Method 1, $\widehat{C}_{[1]}$, is a simple average of the estimates produced under $\widetilde{\lambda}_{[1]}$.
+The overall estimate for Method 1, $\widehat{C}_{[1]}$, is a simple average of the estimates produced under $\widetilde{\lambda}_{[1]}$:
 \begin{align}
@@
 
-Practically speaking, in our simulations we use a 50/50 split of the indices for each partition.
-The subsets of each partition are selected at random, that is we could select the same subset split more than once.
+In our simulations we chose an equal split of the indices for each partition: $|T_1(l)| = |T_2(l)|$.
+The subsets of each partition are selected at random, so we could select the same subset split more than once.
 We partition a total 10 times ($p = 10$).
-For example if $r = 4$ then $T_1(1) = \{1,4\}$ and $T_2(1) = \{2, 3\}$ would be a valid subset selection for partition 1.
+For example, if $r = 4$ then $T_1(1) = \{1,4\}$ and $T_2(1) = \{2, 3\}$ would be a valid subset selection.
 This holds in methods 2 and 4 as well, which also use partitioning.
 
-%
-%
-%
-\subsection{Method 2: Cross validated likelihood}
-A well known method which takes advantage of replicate data is cross validation.
-Cross validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.
+\subsection{Method 2: Cross-validated likelihood}
+
+\amy{up to here with editing}
+
+A frequently employed method that takes advantage of replicate data is cross-validation.
+Cross-validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.
 The model is trained on one subset of the data and then evaluated using another.
@@
 $C$ is always estimated using the frequency counts.
-Methods 2 and 4 are motivated by the desire to do something similar to cross validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.
+Methods 2 and 4 are motivated by the desire to do something similar to cross-validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.
 
@@
 \end{align}
- In the evaluation step given in Equation \ref{eq:selected_lambda_2} we must use the unpenalized likelihood, as opposed to the penalized version.
+ In the evaluation step given in Equation \eqref{eq:selected_lambda_2} we must use the unpenalized likelihood, as opposed to the penalized version.
 The penalty term is in fact always positive, and has the potential dominate the likelihood if a very large $\lambda$ exists in $\lambdagrid$.
@@
 Consider a single frequency count table.
-Using Equation \ref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
+Using Equation \eqref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
 \begin{equation}
@@
 \end{align}
-Equation \ref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.
+Equation \eqref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.
 Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.
-Equation \ref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.
-Nonetheless we use Equation \ref{eq:chi_sq_compact} below in extending to replicates for more concise notation.
+Equation \eqref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.
+Nonetheless we use Equation \eqref{eq:chi_sq_compact} below in extending to replicates for more concise notation.
 
@@
 \end{align}
-In Equation \ref{eq:selected_lambda_3} note the estimated counts $\left( \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)$ are the same over all replicates ($j$).
+In Equation \eqref{eq:selected_lambda_3} note the estimated counts $\left( \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)$ are the same over all replicates ($j$).
 This follows because we do not have independent $C, \eta$ estimates from each replicate, but a shared estimate for the sample.
@@
 %
-\subsection{Method 4: Cross validated goodness of fit}
+\subsection{Method 4: Cross-validated goodness of fit}
 
@@
 \hline
-Method 2: Cross validated likelihood & 797 & 521 & 492 & \cellcolor{blue!25}{\textbf{602}} & \cellcolor{blue!25}{\textbf{658}} & \cellcolor{blue!25}{617} \\
+Method 2: Cross-validated likelihood & 797 & 521 & 492 & \cellcolor{blue!25}{\textbf{602}} & \cellcolor{blue!25}{\textbf{658}} & \cellcolor{blue!25}{617} \\
 \hline
@@
 \hline
-Method 4: Cross validated g.o.f. & 812 & 571 & 533 & \cellcolor{blue!25}{738} & 787 & \cellcolor{blue!25}{679}  \\
+Method 4: Cross-validated g.o.f. & 812 & 571 & 533 & \cellcolor{blue!25}{738} & 787 & \cellcolor{blue!25}{679}  \\
 \hline
@@
 Our results from Section \ref{sec:fixed_lambda} hint at a positive correlation between $r$ and the optimal choice of $\lambda$.
-We hypothesized that this may be the case and considered other ways of extending the objective function given in Equation \ref{eq:objective}.
+We hypothesized that this may be the case and considered other ways of extending the objective function given in Equation \eqref{eq:objective}.
 For instance, using $-h(J)\lambda \log p_{\eta}(0)$ where $h(J)$ is $|J|$ or $\sqrt{|J|}$ seemed like plausible extensions, whereas the definition we give uses $h(J) = 1$.
@@
 For instance, we could expand the grid until the fixed estimates (e.g.
-$\widehat{C}_{\lambda}$ in Equation \ref{eq:c_hat_lambdas_method_3}) have been suppressed to the observed richness.
+$\widehat{C}_{\lambda}$ in Equation \eqref{eq:c_hat_lambdas_method_3}) have been suppressed to the observed richness.
 Choosing $\lambdagrid$ remains a consideration when applying our methods.

### 2019-08-05 rre_manuscript/paper.tex
minor 5/2 and 5.3 edits

@@
 
-\amy{up to here with editing}
+% A frequently employed method that takes advantage of replicate data is cross-validation.
+Cross-validation is commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes \citep{statistical-machine-learning}.
+In this setting, the model is trained on one subset of the data and then evaluated using another.
+However, in the species problem we never observe true outcomes: $C$ is only ever estimated, and never observed.
+Methods 2 and 4 are motivated by the idea that subsets of the data should produce similar estimates,
+ % desire to do something similar to cross-validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.
 
-A frequently employed method that takes advantage of replicate data is cross-validation.
-Cross-validation is most commonly applied to evaluate the strength of a model in a dataset containing both features and outcomes.
-The model is trained on one subset of the data and then evaluated using another.
-However, in the species problem we never observe true outcomes, i.e.
-$C$ is always estimated using the frequency counts.
-Methods 2 and 4 are motivated by the desire to do something similar to cross-validation, but we have to find a function which takes the place of mean prediction error in the absence of any true $C$ observations.
-
-In Method 2 we use one subset of the data to generate estimates.
-Then we evaluate those estimates using the likelihood with the data from the other subset.
-As before, we partition the data into two subsets $p$ times, calling them $T(l)$ for the training subset of the $l$-th partition, and $E(l)$ for the evaluation subset of the $l$-th partition.
-$T(l), E(l)$ are created exactly as in Method 1, we only relabel them to indicate the change in purpose.
-Then for each $\lambda \in \lambdagrid$, $l \in \{1, \dots , p\}$ let
+In Method 2 we propose to partition the data into two subsets, and evaluate the estimates based on the ``training'' subset using the likelihood based on the ``evaluation'' subset.
+  % use one subset of the data to estimate $C$ and $\eta$.  Then we evaluate those estimates using the likelihood with the data from the other subset.  As before, w
+We partition the data into two subsets $p$ times, calling them $T(l)$ for the training subset of the $l$-th partition, and $E(l)$ for the evaluation subset of the $l$-th partition.
+For each $\lambda \in \lambdagrid$, $l \in \{1, \dots , p\}$ let
 \begin{equation}
@@
 so that we have an estimate using only the training subset for each $\lambda$ value and each partition.
-A $\lambda$ value is selected by plugging those estimates into the unpenalized likelihood with $J = E(l)$.
+A $\lambda$ value is selected by plugging those estimates into the unpenalized likelihood with $J = E(l)$:
 \begin{align}
@@
 \end{align}
- In the evaluation step given in Equation \eqref{eq:selected_lambda_2} we must use the unpenalized likelihood, as opposed to the penalized version.
+
+ % In the evaluation step given in Equation \eqref{eq:selected_lambda_2} we use the unpenalized likelihood, as opposed to the penalized version.
 The penalty term is in fact always positive, and has the potential dominate the likelihood if a very large $\lambda$ exists in $\lambdagrid$.
@@
 
-We depart from our partitioning scheme to explore another desirable property of an estimator.
-An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.
-Consider a single frequency count table.
-Using Equation \eqref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
+% We depart from our partitioning scheme to explore another desirable property of an estimator.
+Method 3 uses goodness of fit of the fitted frequency counts to select an optimal tuning parameter.
+% An ideal estimator would produce parameter estimates $\widehat{C}$, $\widehat{\eta}$ which fit well with the frequency counts in our sample.
+% Consider a single frequency count table.
+Given that $c \sim$ Binomial $\left(C, 1- p_{\eta}(0)\right)$ and $(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$, we have that
+% Using Equation \eqref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
 \begin{equation}
- \mathbb{E}\left(f_k\right) = \mathbb{E}\left(f_k|c \right) = \mathbb{E}\left( c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right) = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k)
+ \mathbb{E}\left[f_k\right] = \mathbb{E}\left[\mathbb{E}\left[f_k|c \right]\right] = \mathbb{E}\left[ c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right] = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k).
 \end{equation}
-Then given estimates $\widehat{C}$ and $\widehat{\eta}$ we use a plug-in estimate for the expected counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.
-If we consider each $f_k$ a category, the usual $\chi^2$ goodness of fit statistic is
+Then, given estimates $\widehat{C}$ and $\widehat{\eta}$, we consider a plug-in estimate for the expected frequency counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.
+The usual $\chi^2$ goodness of fit statistic is then
 \begin{align}
-&\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} \\
-&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
-& = \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
+\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} %&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
+= \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
 \end{align}
-Equation \eqref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.
-Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.
+where $\kmax$ is the largest $k$ such that $f_k > 0$.
+% Equation \eqref{eq:chi_sq_expanded} uses the fact that we have some $k_{max}$ in every frequency count table such that $f_k = 0$ for all $k > \kmax$.
+% Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.
 Equation \eqref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.
-Nonetheless we use Equation \eqref{eq:chi_sq_compact} below in extending to replicates for more concise notation.
+%Nonetheless we use Equation \eqref{eq:chi_sq_compact} below in extending to replicates for more concise notation.
 
-In method 3 we make use of this goodness of fit metric by first generating estimates using all replicates.  For each $\lambda \in \lambdagrid$ let
+\amy{up to here with editing}
+
+In Method 3 we make use of this goodness of fit metric by first generating estimates using all replicates.  For each $\lambda \in \lambdagrid$ let
 \begin{equation}

### 2019-08-06 rre_manuscript/paper.tex
finished first pass of edits

@@
 We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
-Our method penalises the probability of a class being unobserved, since high probabilities of unobserved classes increase the variance of species richness estimates.
-Tuning the penalisation parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods to tune the penalization parameter.
-We illustrate the performance of the method by estimating the microbial diversity of Lake Champlain over ...}
+Because high probabilities of unobserved classes increase the variance of species richness estimates,
+our method penalizes the probability of a class being unobserved, since high probabilities of unobserved classes increase the variance of species richness estimates.
+Tuning the penalization parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods for tuning the penalization parameter.
+We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.}
 \end{abstract}
@@
 
-The \textit{species problem} concerns estimating the number of classes $C$ that are present in a population.
+The \textit{species problem} concerns estimating the number of classes that are present in a population, which we call $C$.
 $n$ individuals from the population can be sampled to find which classes they belong to, but since the  number of individuals in the sample is less than the total population size (which may be infinite), only $c$ classes are observed ($c \leq C$).
@@
 The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
-\amy{Need to justify why we are bringing this up}
+%\amy{Need to justify why we are bringing this up}
 
@@
 In particular, for $\lambda > 0,$ the penalized maximum likelihood solution $\hat{C}_\lambda$ is less than or equal to $\hat{C}_0$, the maximum likelihood estimate.
-Furthermore, for a large enough $\lambda$, $\hat{C}_\lambda = c$; that is, the penalised maximum likelihood estimate shrinks to the observed richness $c$.
+Furthermore, for a large enough $\lambda$, $\hat{C}_\lambda = c$; that is, the penalized maximum likelihood estimate shrinks to the observed richness $c$.
 \citet{wang_2005} also consider two other penalty functions of similar form, which we do not discuss here except to note that our tuning parameter selection methods would equally apply.
@@
 
-We focus on the gamma-Poisson model and penalised log-likelihood of the form \eqref{eq:wang_lindsay}, and consider the case where we observe $r$ independent frequency count tables from the population under study.
+We focus on the gamma-Poisson model and penalized log-likelihood of the form \eqref{eq:wang_lindsay}, and consider the case where we observe $r$ independent frequency count tables from the population under study.
 % with penalization as proposed by \citet{wang_2005}.
@@
 To implement a direct optimization approach we search over candidate $C$ values in a grid.
-A gradient-based search over $\eta$ values is used at each $C$ in the grid to find the maximum penalised likelihood  $(C, \eta)$.
+A gradient-based search over $\eta$ values is used at each $C$ in the grid to find the maximum penalized likelihood  $(C, \eta)$.
 
@@
 \begin{table}[h]
-\tblcaption{The penalised maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
+\tblcaption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
 $\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
@@
 
-We compare all proposed methods to the unpenalised MLE using $J = \{1, \dots r \}$:
+We compare all proposed methods to the unpenalized MLE using $J = \{1, \dots r \}$:
 %Recall that fixing $\lambda = 0$ gives us the unpenalized likelihood in Equation \eqref{eq:objective}, and so by denoting the $C$ estimate obtained using Method 0 as $\widehat{C}_{[0]}$, then
@@
 
-In Method 2 we propose to partition the data into two subsets, and evaluate the estimates based on the ``training'' subset using the likelihood based on the ``evaluation'' subset.
+In Method 2 we propose to repeatedly partition the data into two subsets, and evaluate the estimates based on the ``training'' subset using the likelihood based on the ``evaluation'' subset.
   % use one subset of the data to estimate $C$ and $\eta$.  Then we evaluate those estimates using the likelihood with the data from the other subset.  As before, w
@@
 \sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{f}_k \right)^2 }{\widehat{f}_k} %&=\sum_{k=1}^{\infty} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}}  \label{eq:chi_sq_compact} \\
-= \sum_{k=1}^{k_{max}} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k) \label{eq:chi_sq_expanded}
+&= \sum_{k=1}^{\kmax} \frac{\left(f_k - \widehat{C}p_{\widehat{\eta}}(k) \right)^2 }{\widehat{C} p_{\widehat{\eta}(k)}} \ + \ \widehat{C} \sum_{k = k_{max}+1}^{\infty} p_{\widehat{\eta}}(k)
+\label{eq:chi_sq_expanded}
 \end{align}
@@
 
-\amy{up to here with editing}
+% \amy{up to here with editing}
 
@@
 \end{align}
-In Equation \eqref{eq:selected_lambda_3} note the estimated counts $\left( \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)$ are the same over all replicates ($j$).
+% In Equation \eqref{eq:selected_lambda_3} note the estimated counts $\left( \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)$ are the same over all replicates ($j$).
 This follows because we do not have independent $C, \eta$ estimates from each replicate, but a shared estimate for the sample.
@@
 
-In Method 4 we return to the partitioning scheme of Method 2.
-Rather than using the likelihood in the evaluation step, we hypothesize that the goodness of fit metric may be a better choice.
-Partition the data $p$ times into $T(l)$, the training set in partition $l$, and $E(l)$, the evaluation set in partition $l$.
-For all $\lambda \in \lambdagrid$ and $l \in \{1, \dots , p \}$, we generate estimates exactly as in Method 2,
+In Method 4 we return to the partitioning scheme of Method 2.  Rather than using the likelihood in the evaluation step, we hypothesize that the goodness-of-fit metric may be a better choice.
+We partition the data $p$ times into a training set $T(l)$ and a evaluations set $E(l)$.
+For all $\lambda \in \lambdagrid$ and $l \in \{1, \dots , p \}$, we generate estimates exactly as in Method 2:
 \begin{equation}
-\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right) \label{eq:c_hat_lambda_method_4}
+\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right). \label{eq:c_hat_lambda_method_4}
 \end{equation}
-To select $\lambda$ we evaluate the goodness of fit metric using only the evaluation subset data, $j \in E(l)$.  The $\lambda$ which produces the best fitting estimates is $\widetilde{\lambda}_{[4]}$,
+To select $\lambda$ we evaluate the goodness of fit metric using only the evaluation subset data, $j \in E(l)$.  The $\lambda$ which produces the best fitting estimates is $\widetilde{\lambda}_{[4]}$:
 \begin{align}
@@
 \end{align}
-Compared to Method 2, this method generates estimates $\widehat{C}^{T(l)}_{\lambda}$ in the same fashion but evaluates in a completely different way.
-Compared with Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all the replicates are used in both steps.
+
+Similar to Method 2, this method generates training set-based estimates $\widehat{C}^{T(l)}_{\lambda}$, however, it evaluates these estimates using goodness-of-fit rather than likelihood maximisation.
+Compared to Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all the replicates are used in both steps.
 
@@
 
-We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the RMSE of estimates from each method in simulations.
+We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the performance of each estimator via simulations.
+We simulate zero-truncated gamma-mixed Poisson data using the same parameters used in Section \ref{sec:efficacy_sims}.
+Based on the results of Section \ref{sec:efficacy_sims} we know that estimation can be improved on this set of parameters through the use of a fixed penalization parameter.
+The purpose of this section is to determine if any method can reliably select a $\lambda$ that reduces RMSE compared to unpenalized maximum likelihood estimation.
 
-To evaluate which method has the best performance we will apply them to gamma-Poisson simulated data with the same parameters used in Section \ref{sec:efficacy_sims}.
-Based on the results in Section \ref{sec:efficacy_sims} we know that estimation can be improved on this set of parameters through the use of a fixed penalization parameter.
-The purpose of this section is to determine if any method can reliably select a $\lambda$ that improves the RMSE.
+\amy{make sure to say zero truncated negative binomial throughout}
 
-We present two simulation below, and in each it is important to select an appropriate $\lambdagrid$.
-In particular our grid needs to be large enough to reveal the negative aspects of methods which over-penalize.
+We present two simulations below, and in each it is important to select an appropriate $\lambdagrid$.
+% In particular our grid needs to be large enough to reveal the negative aspects of methods which over-penalize.
 For instance, if $\lambdagrid = \{0, \dots , 50\}$ and the optimal value is $\lambda = 45$, a method which deterministically selects the maximum value in $\lambdagrid$ (an extreme over-penalizer) would perform well.
 This would lead us to spuriously conclude this hypothetical over-penalizer is effective.
-One of the important foundations laid in Section \ref{sec:efficacy_sims} is that we now know approximately what the optimal values are for each simulation setting (Table \ref{tab:fixed_lambda_results}).
-We select a large enough grid by extending to approximately double the optimal value in both simulations presented.
+In Section \ref{sec:efficacy_sims}  we established the approximate optimal values for each simulation setting (Table \ref{tab:fixed_lambda_results}).
+% One of the important foundations laid in Section \ref{sec:efficacy_sims} is that we now know approximately what the optimal values are for each simulation setting (Table \ref{tab:fixed_lambda_results}).
+We select a large enough grid by extending to at least double the optimal value in both simulations presented.
 
-\subsection{Primary simulation}
+\subsection{Primary simulation -- TODO Rename section}
 \label{sec:tuning_simulation_1}
 
-\subsubsection{Setup}
+% \subsubsection{Setup}
 
@@
 Recall these $\eta$ are the values chosen based on our motivating examples.
-We know that the optimal $\lambda$ values are between 10 and 20 under these $\eta$ parameters for $r$ frequency count tables, so we chose $\lambdagrid = \{0, 5, 10, \dots 60\}$ for this simulation.
+We know that the optimal $\lambda$ choice for these $r$ and $\eta$ are between 10 and 20, so we chose $\lambdagrid = \{0, 5, 10, \dots 60\}$ for this simulation.
 Methods 0-4 are all evaluated over the same random draws for each parameter choice.
 
-\subsubsection{Results}
+% \subsubsection{Results}
 
-In Table \ref{tab:tuning_sim_1} we show the RMSE over all simulations for each method and each $\eta, r$ combination.
-Over all parameter choices Method 3 performs at least as well as Method 0, with a RMSE which was 0-23\% lower depending on $\eta$ and $r$.
-Method 3 is the only method which is at least as good as Method 0 for all parameter choices.
+In Table \ref{tab:tuning_sim_1} we show the RMSE over all simulations for each method and each combination of $\eta$ and $r$.
+Over all parameter choices, Method 3 performs at least as well as Method 0, with a RMSE which was between 0 and 23\% lower.
+Method 3 is the only method which performs at least as well as Method 0 for all parameter choices.
 
-Universally superior RMSE is the most desirable property for a method, but we can characterize the performance of the methods based on the simulation parameters as well.
+We note some differences between the performance of the methods for different $\eta.$ When $\eta = (10^{-2},10^{-5})$, both Methods 2 also outperforms Method 0 for $r \in \{6, 10, 14\}$, while Method 2 does not outperform Method 0 for any $r$ when $\eta = (10^{-1},10^{-1})$.
+Method 4 outperforms Method 0 for $r \in \{6, 14\}$ when $\eta = (10^{-2},10^{-5})$, but never when $\eta = (10^{-1},10^{-1})$.
+We conjecture that the differing performance across $\eta$ is due to the differing rare species structures implied by the different $\eta$'s.
+For example $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
+
+% Universally superior RMSE is the most desirable property for a method, but we can characterize the performance of the methods based on the simulation parameters as well.
 For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
@@
 
+[Method 1 has better performance only for $r = 6$ on both $\eta$ choices.
+In general, we would expect that any truly effective method would improve relative to Method 0 as $r$ increases, as this means more data was collected.
+The fact that Method 1 loses its edge over Method 0 as we collect more data is a discouraging finding.
+Method 3 shows the opposite trend, which is especially clear in Figure \ref{fig:tuning_sim_1} on $\eta = (10^{-2},10^{-5})$.
+As we increase $r$ the simulations appear closer to the truth in Method 3, and Method 3 has the best RMSE for $r = 14$ under both $\eta$ values.]
+
 \begin{table}[ht]
-\tblcaption{RMSE$(\widehat{C}_{[\text{method}]})$ over all simulations.
-Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.
-Under each $\eta, r$ combination 100 simulations were run.
-Methods with RMSE better than Method 0 have a \textcolor{blue!50}{blue} background, and the best method is \textbf{bolded}.
+\tblcaption{RMSE$(\widehat{C})$ for Methods 0-4 based on a zero-truncated gamma-mixed Poisson data generating process.
+$C = 1000$ is constant for all simulations.
+Under each $\eta, r$ combination, 100 simulations were run.
+Methods with RMSE better than Method 0 have a \textcolor{blue!50}{blue} background, and the best method for each $\eta, r$ combination is \textbf{bolded}.
 \label{tab:tuning_sim_1}}
@@
 
-Method 1 has better performance only for $r = 6$ on both $\eta$ choices.
-In general, we would expect that any truly effective method would improve relative to Method 0 as $r$ increases, as this means more data was collected.
-The fact that Method 1 loses its edge over Method 0 as we collect more data is a discouraging finding.
-Method 3 shows the opposite trend, which is especially clear in Figure \ref{fig:tuning_sim_1} on $\eta = (10^{-2},10^{-5})$.
-As we increase $r$ the simulations appear closer to the truth in Method 3, and Method 3 has the best RMSE for $r = 14$ under both $\eta$ values.
-
 In Figure \ref{fig:tuning_sim_1} we display the $C$ estimates for each method.
 Recall that as we increase $\lambda$, $\widehat{C}_\lambda$ monotonically decreases.
-Therefore we can discuss whether a method is over-penalizing (selecting $\widetilde{\lambda}$ which is larger than optimal) or under-penalizing (selecting $\widetilde{\lambda}$ which is smaller than optimal) without additional information.
-If $\widehat{C}$ is too high then the method has under-penalized and if $\widehat{C}$ is too low then the method has over-penalized.
+Therefore we can discuss whether a method is over-penalizing (selecting $\widetilde{\lambda}$ which is larger than optimal) or under-penalizing (selecting $\widetilde{\lambda}$ which is smaller than optimal) without additional information: if $\widehat{C} > C$, then the method has under-penalized while if $\widehat{C} < C$ the method has over-penalized.
 
@@
 
-It is clear in Figure \ref{fig:tuning_sim_1} that Method 1 over-penalizes for all parameter choices, as all of the estimates are far below the truth.
-We can make sense of this observation based on what we observed in Figure \ref{fig:fixed_lambda}: for very high values of $\lambda$ the estimates have low variance because they are driven down to the observed richness $\max_j c_j$.
-However the observed richness has a large negative bias.
-Method 1 inherits this large bias as it minimizes variance, so RMSE is not effectively minimized.
+We see from Figure \ref{fig:tuning_sim_1} that Method 1 over-penalizes for all parameter choices, as all of the estimates are far below the truth.
+We can understand this result given Figure \ref{fig:fixed_lambda}: for very high values of $\lambda$ the estimates are equal to $\max_j c_j$, and so have low variance.
+However, the observed richness is severely negatively biased for $C$.
+Given its large and consistent negative bias and high RMSE, we do not discuss  Method 1 further.
 
-Method 2 has the opposite behavior: the estimates tend to be too high, so Method 2 underpenalizes.
-This is especially noticeable in $\eta = (10^{-2}, 10^{-5})$, where the Method 0 bias is substantial.
-Under this $\eta$ value Method 2 does better than Method 0, but the estimates are almost all too high.
-
+Method 2 has the opposite behavior: the estimates tend to be too high, especially when $\eta = (10^{-2}, 10^{-5})$.
+This is similarly the case for Method 0, though we see that Method 2 has slightly lower bias compared to Method 0.
+%Under this $\eta$ value Method 2 does better than Method 0, but the estimates are almost all too high.
 Even for $\eta = (10^{-1}, 10^{-1})$, where the Method 0 bias is smaller, we see that Method 2 has poor performance (Table \ref{tab:tuning_sim_1}).
@@
 As a consequence Method 2 has slightly higher variance when compared with Method 0, as evidenced by the interquartile range in Figure \ref{fig:tuning_sim_1} for $\eta = (10^{-1}, 10^{-1})$.
-In conclusion, Method 2 does not seem to be effectively tuning $\lambda$, making the cost of using only half the data to generate estimates especially problematic.
+In conclusion, Method 2 does not appear to be effectively
+tuning $\lambda$, and sample splitting to generate estimates may be leading to high variance.
 
-Methods 3 and 4 have different penalization tendencies depending on the $\eta$ and $r$ values of the simulation.
-This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.
+Methods 3 and 4, which are both based on goodness-of-fit criteria, display different patterns depending on the $\eta$ and $r$ values of the simulation.
+% This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.
 In Methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.
-Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.
-
-While we find the indications that Method 4 is adaptable to the dataset encouraging, under $\eta = (10^{-1}, 10^{-1})$ the RMSE for Method 4 is 15-75\% greater than for Method 0, while Method 3 is at least as good as Method 0.
-Method 3 is adaptive while being simpler, faster and having better performance than Method 4.
+While we find the indications that Method 4 is adaptable to the dataset encouraging, under $\eta = (10^{-1}, 10^{-1})$ the RMSE for Method 4 is 15-75\% greater than for Method 0.
+In comparison, Method 3 is at least as good as Method 0 with respect to RMSE.
+In addition,  Method 3 is simpler and faster than Method 4.
 Given deficits noted for Method 1 and 2, Method 3 is by far the most promising method based on this simulation.
 We propose to conduct a secondary simulation to further challenge Method 3, testing whether it remains superior to Method 0 over a wider range of $\eta$ values.
+
+\amy{move later -- Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.}
 
@@
 
-\subsubsection{Setup}
+% \subsubsection{Setup}
 
-In this simulation we will fix $C = 1000$, $r \in \{6, 10, 14\}$ and do 100 simulations for each combination of $\eta, r$ as before.
+In this simulation we will fix $C = 1000$, $r \in \{6, 10, 14\}$ and perform 100 simulations for each combination of $\eta$ and $r$.
 We let $\eta \in \{ (10^{-1}, 10^{-3}), (10^{-1}, 10^{-5}) \}$ and only consider methods 0 and 3.
@@
 
-\subsubsection{Results}
+% \subsubsection{Results}
 
-In Table \ref{tab:tuning_sim_2} we display the RMSE for both methods and each $\eta, r$ combination.
+In Table \ref{tab:tuning_sim_2} we display the RMSE for both methods and each combination of $r$ and $\eta$.
 Method 3 has slightly worse performance on all $r$ values for $\eta = (10^{-1}, 10^{-3})$, with a RMSE which was 14-36\% greater than Method 0.
-This $\eta$ generates more rare species so the relatively poor performance of Method 3 is consistent with the results our previous simulation.
+This $\eta$ generates more rare species, so the comparatively poor performance of Method 3 is consistent with the results our previous simulation.
 In Section \ref{sec:tuning_simulation_1}, Method 3 had greatly improved estimation in data with many abundant species, but the improvements over Method 0 were weaker in datasets that had many rare species.
@@
 In Figure \ref{fig:tuning_sim_2} we display boxplots of the $C$ estimates over all simulations.
-For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is noticeably inflated on a few simulations.
-Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species (while still having some singletons and unobserved species).
-Therefore this is a excellent example of a dataset which will cause the instability problem in maximum likelihood estimation we reviewed in Section \ref{sec:literature_review}.
+For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is very large for a few simulations.
+Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species.
+Therefore, this is a excellent example of a simulation structure which will cause the instability problem in maximum likelihood estimation discussed in Section \ref{sec:literature_review}.
 Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.
@@
 
-As a result of these simulations, we conclude that none of the methods (including method 0) are best for all simulation settings.
-Method 1 resulted in a strong negative bias no matter which data are being analyzed.
-Method 3 had better performance than Method 0 when there are highly abundant species present in the data, which makes Method 0 highly unstable.
-When there are many rare species Method 3 is better than methods 1, 2 and 4 but not necessarily Method 0.
-Method 3 also improved relative to Method 0 with greater $r$ and appears to better adapt to the data.
-Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but we cannot recommend them over Method 3 due to their inferior performance on data dominated by rare species.
+As a result of these simulations, we conclude that no method (including Method 0) is best for all simulation settings, but Method 3 has advantages in many settings.
+Method 3 had improved performance compared to Method 0 when there are highly abundant species present in the data, in which case Method 0 is highly unstable.
+When there are many rare species Method 3 is better than Methods 1, 2 and 4, but is not necessarily better than Method 0.
+Method 3 also improved relative to Method 0 with greater $r$.
+Methods 2 and 4 had lower RMSE than Method 0 when highly abundant species exist, but have inferior performance compared to Method 3 on data dominated by rare species.
+Method 1 resulted in a strong negative bias in all simulations, and we do not recommend its use.
 
@@
 
-To illustrate the performance of these methods on real data we will use samples collected in consecutive years from Lake Champlain (\citet{tromas_2017}).
-We limit our samples to replicates from the summer months and the littoral zone of the lake  (near the shore).
-There were 8 replicates meeting these criteria from 2009, 6 replicates from 2010 and 6 from 2011.
-The maximum observed richness (most species observed in any replicate that year) was 560 in 2009, 572 in 2010 and 715 in 2011.
+To illustrate the performance of our methods on ecological data, we estimate strain-level microbial diversity in Lake Champlain, a large eutrophic lake in Canada.
+We analyse data from \citet{tromas_2017}, considering samples from the littoral zone in the summer season of the same year as replicates.
+This gives us 8 replicates from 2009, 6 replicates from 2010 and 6 replicates from 2011.
+%The maximum observed richness in any given sample was 560 in 2009, 572 in 2010 and 715 in 2011.
 
-The richness estimates from each method are displayed in Table \ref{tab:data_analysis}.
-Our estimates from each method are at least 10 times as high as the observed richness, and in some cases they reach the maximum value in our grid search for $C$ (which is 20 times the observed richness).
-These extreme estimates suggest that the Gamma-Poisson model which underlies all methods is a poor fit to these data.
-In simulations penalization had the intended effect of lowering inflated estimates.
+% These extreme estimates suggest that the Gamma-Poisson model which underlies all methods is a poor fit to these data.
+% In simulations penalization had the intended effect of lowering inflated estimates.
 However, we always simulated under a Gamma-Poisson model in simulations.
@@
 
-\begin{table}[t]
-\tblcaption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
+% \begin{table}[t]
+% \tblcaption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
 $\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
 The number of replicates in each sample is given by $r$.
-\label{tab:data_analysis}}
-\centering
-\begin{tabular}{|r|r|r|r|}
-  \hline
-& 2009  & 2010 & 2011 \\
-  \hline
-$\widehat{C}_{[0]}$ & 11200 & 11440 & 14014 \\
-  $\widehat{C}_{[1]}$ & 9645 & 11139 & 12246 \\
-  $\widehat{C}_{[2]}$ & 10612 & 11120 & 11946 \\
-  $\widehat{C}_{[3]}$ & 11200 & 11440 & 14014 \\
-  $\widehat{C}_{[4]}$ & 10744 & 11165 & 10328 \\
-   \hline
-     $r$ & 8 & 6 & 6 \\
-  \hline
-\end{tabular}
-\end{table}
+% \label{tab:data_analysis}}
+% \centering
+% \begin{tabular}{|r|r|r|r|}
+%   \hline
+% & 2009  & 2010 & 2011 \\
+%   \hline
+% $\widehat{C}_{[0]}$ & 11200 & 11440 & 14014 \\
+%   $\widehat{C}_{[1]}$ & 9645 & 11139 & 12246 \\
+%   $\widehat{C}_{[2]}$ & 10612 & 11120 & 11946 \\
+%   $\widehat{C}_{[3]}$ & 11200 & 11440 & 14014 \\
+%   $\widehat{C}_{[4]}$ & 10744 & 11165 & 10328 \\
+%    \hline
+%      $r$ & 8 & 6 & 6 \\
+%   \hline
+% \end{tabular}
+% \end{table}
 
-In Figure \ref{fig:data_analysis} we display the same results visually.
+The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.
+We see that Methods 0 and 3 produce identical estimates for all samples, and these estimates are larger than estimates for Methods 1, 2 and 4.
+We note that Methods 1, 2 and 4 use partitioning, but the trend of lower estimates for partition-based estimates was not observed in simulations (see Figure \ref{fig:tuning_sim_1}).
+
+Our estimates from each method are all at least 10 times as high as the observed richness, indicating a significant fraction of rare and unobserved species.
+Based on our previous results, estimation error is lowest for Methods 0 and 3 when there are many rare species.
+We therefore estimate $\widehat{C}^{2009} = 11200$, $\widehat{C}^{2010} = 11440$ and $\widehat{C}^{2011} = 14014$.
+% , and in some cases they reach the maximum value in our grid search for $C$ (which is 20 times the observed richness).
+
+% In Figure \ref{fig:data_analysis} we display the same results visually.
 The methods which use partitioning (Methods 1, 2 and 4) consistently produce lower estimates than those which do not (Methods 0 and 3).
@@
 \begin{figure}[t]
-\caption{A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.
+\caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) based on our proposed methods.
+%$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
+The number of replicates in each sample is given by $r$.
+%
+% A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.
 \label{fig:data_analysis}}
@@
 
-\noindent \alex{Next: Modify the discussion to harmonious with what you've written here.}
+% \noindent \alex{Next: Modify the discussion to harmonious with what you've written here.}
 
-\section{Discussion}
+\section{Discussion and Conclusion}
 \label{sec:discussion}
 
-In this paper we have proposed an extension of the penalization in \citet{wang_2005} to data with biological replicates and proposed several methods for tuning the penalization parameter.
-This extends the existing body of literature on species richness by demonstrating that penalization is useful in analyzing replicate data.
+In this paper we discussed an extension of the penalized maximum likelihood procedure of \citet{wang_2005} for species richness estimation to data with biological replicates, and proposed several methods for tuning the penalization parameter.
+We demonstrated that penalization can reduce estimation error (defined as square root mean squared error) when analyzing replicate data.
 Our goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
-In this section we note promising directions for future work.{}
+On microbial abundance data \amy{TODO}
 
 Our results from Section \ref{sec:fixed_lambda} hint at a positive correlation between $r$ and the optimal choice of $\lambda$.
-We hypothesized that this may be the case and considered other ways of extending the objective function given in Equation \eqref{eq:objective}.
-For instance, using $-h(J)\lambda \log p_{\eta}(0)$ where $h(J)$ is $|J|$ or $\sqrt{|J|}$ seemed like plausible extensions, whereas the definition we give uses $h(J) = 1$.
-Our investigations revealed that the trend is unfortunately not this simple as $|J|$ becomes large.
-A parsimonious extension which accounts for this observed trend would be a valuable modification to our work.
+To investigate this, we investigated whether the optimal choice of $\lambda$ remained constant for penalties of the form $-h(J)\lambda \log p_{\eta}(0)$ (instead of $-\lambda \log p_{\eta}(0)$; see Equation \eqref{eq:objective}).
+We tested $h(J) = J$ and $h(J) = \sqrt{J}$, but found that the trend in optimal $\lambda$ is not so simple.
+We leave further investigation into how the optimal $\lambda$ varies with $J$ to future work.
 
-In \citet{wang_2005} the gamma-Poisson model is explicitly considered but their results apply to many other Poisson mixture models.
-Our most promising method, Method 3, is also amenable to other modeling choices provided that we have some way to calculate the fitted counts $\widehat{f_k}$.
-It would be interesting to explore whether our results hold for a more flexible mixed Poisson distribution, such as the proposal by \citet{norris_1998}.
-Flexible models often have better performance under model misspecification, and we simulated under correctly specified models.
-Therefore, this variation on  Method 3 would be especially interesting to study under model misspecification.
+The approach of \citet{wang_2005} is considerably more general than the gamma-Poisson model.
+The gamma-Poisson model is common for modeling microbiome data \citep{deseq2,mcmurdie2013phyloseq}, and for this reason we focused on it in our investigation.
+However, our most promising method, Method 3, is also amenable to other modeling choices provided that there is some way to calculate the fitted counts $\widehat{f_k}$.
+We leave investigation of tuning parameter selection under different models to future work.
 
-We intentionally chose a likelihood optimization algorithm which was stable and exhaustive but admittedly slow.
-Confirming that our results hold when a faster approach is used would be another valuable extension to our work.
-We did not derive theoretical standard errors in this paper, and the long computation times precluded the consideration of a bootstrapping approach.
-However, if our results hold when a faster likelihood maximization algorithm is used then a standard error procedure could become feasible.
+For our investigations, we intentionally chose a likelihood optimization algorithm which was stable and exhaustive.
+However, refining the optimisation algorithm would be a valuable extension to our work.
+We also did not construct standard errors for our estimates, and the long computation times precluded the consideration of a bootstrapping approach.
+A faster optimisation algorithm would facilitate a resample-based standard error procedure.
 
-We have been careful to try to create methods which only depend on $\lambdagrid$ being large enough to contain the optimal value, with no change in results if it is too large.
+% We have been careful to try to create methods which only depend on $\lambdagrid$ being large enough to contain the optimal value, with no change in results if it is too large.
 This makes $\lambdagrid$ a more forgiving parameter than fixed $\lambda$ values, but we still need to make a choice to analyze data.
@@
 
-Method 3 was our top performer and it is the only tuning method we proposed which can be run without replicates ($r = 1$).
+% Method 3 was our top performer and it is the only tuning method we proposed which can be run without replicates ($r = 1$).
 The ever-decreasing cost of high-throughput sequencing suggests biological replicates might be more available in the future.
@@
 \section*{Acknowledgments}
-List of acknowledgements for formalize:
+The authors of this manuscript are grateful to Jim Hughes for many helpful suggestions that improved content and exposition.
 
-{\it Conflict of Interest}:
+{\it Conflict of Interest}: None to declare
 

### 2019-08-06 rre_manuscript/paper.tex
round two of editing; add refs; JCGS style file

@@
-\documentclass[oupdraft]{bio}
-\usepackage{url}
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib, bbm}
+\usepackage{url} % not crucial - just used below for the URL
 
-\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{lipsum}
+
 \usepackage{xparse}
@@
 
+\newtheoremstyle{break}
+  {\topsep}{\topsep}%
+  {\itshape}{}%
+  {\bfseries}{}%
+  {\newline}{}%
+\theoremstyle{break}
+\newtheorem{theorem}{Theorem}
+
+\newtheoremstyle{break}
+  {\topsep}{\topsep}%
+  {\itshape}{}%
+  {\bfseries}{}%
+  {\newline}{}%
+\theoremstyle{break}
+\newtheorem{prop}{Proposition}
+
 \newboolean{DEBUG}
 \setboolean{DEBUG}{true}
-
 \newrgbcolor{amycolor}{.7 .1 .1}
@@
 
-\newrgbcolor{alexcolor}{.1 .7 .7}
-\ifthenelse {\boolean{DEBUG}}
-{\newcommand{\alex}[1]{{\bf \alexcolor{[\@Alex: #1]}}}}
-{\newcommand{\alex}[1]{}}
+\RequirePackage[OT1]{fontenc}
+\RequirePackage{amsthm,amsmath, bbm, xspace, graphics, graphicx}
+%\RequirePackage{breqn}
+%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
 
-\newrgbcolor{editcolor}{.1 .5 .7}
-\ifthenelse {\boolean{DEBUG}}
-{\newcommand{\edit}[1]{{\bf \editcolor{[\@edit: #1]}}}}
-{\newcommand{\edit}[1]{}}
+\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{-.3in}%
+\addtolength{\topmargin}{-.8in}%
 
@@
 
-% Title of paper
-\title{Tuning parameter selection for a penalized estimator of species richness}
+%\bibliographystyle{natbib}
 
-% List of authors, with corresponding author marked by asterisk
-\author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
-% Author addresses
-\textit{Department of Biostatistics,
-University of Washington,
-Health Sciences Building, Box 357232,
-1705 NE Pacific St., Seattle, WA 98195}
-% \\[2pt]
-% % E-mail address for correspondence
-% {payntera@uw.edu, adwillis@uw.edu}
-}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
-% Running headers of paper:
-\markboth%
-% First field is the short list of authors
-{A. Paynter and A.D. Willis}
-% Second field is the short title of the paper
-{Tuning parameter selection for species richness}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if1\blind
+{
+  \title{\bf Tuning parameter selection for a penalized estimator of species richness}
+  \author{Alex Paynter and Amy D Willis\\ \\
+  Department of Biostatistics, University of Washington,\\
+  Health Sciences Building, Box 357232,\\
+  1705 NE Pacific St., Seattle, WA 98195 \\ \\
+  adwillis@uw.edu}
 
   \maketitle
+} \fi
 
-% Add a footnote for the corresponding author if one has been
-% identified in the author list
-\footnotetext{adwillis@uw.edu}
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Tuning parameter selection for a penalized estimator of species richness}
+\end{center}
+  \medskip
+} \fi
 
+\bigskip
 \begin{abstract}
-{Our goal is to estimate the true number of classes in a population, called the species richness.
+  Our goal is to estimate the true number of classes in a population, called the species richness.
 We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
   Because high probabilities of unobserved classes increase the variance of species richness estimates,
-our method penalizes the probability of a class being unobserved, since high probabilities of unobserved classes increase the variance of species richness estimates.
+  our method penalizes the probability of a class being unobserved.
 Tuning the penalization parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods for tuning the penalization parameter.
-We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.}
+We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.
 \end{abstract}
+
+\noindent%
+{\it Keywords: diversity, biodiversity, microbiome, regularization, maximum likelihood}
+\vfill
+
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
+
+% % List of authors, with corresponding author marked by asterisk
+% \author{ALEX PAYNTER, AMY D WILLIS$^\ast$\\[4pt]
+% % Author addresses
+% \textit{Department of Biostatistics,
+% University of Washington,
+% Health Sciences Building, Box 357232,
+% 1705 NE Pacific St., Seattle, WA 98195}
+% % \\[2pt]
+% % % E-mail address for correspondence
+% % {payntera@uw.edu, adwillis@uw.edu}
+% }
+%
+% % Running headers of paper:
+% \markboth%
+% % First field is the short list of authors
+% {A. Paynter and A.D. Willis}
+% % Second field is the short title of the paper
+% {Tuning parameter selection for species richness}
+%
+% \maketitle
+%
+% % Add a footnote for the corresponding author if one has been
+% % identified in the author list
+% \footnotetext{adwillis@uw.edu}
 
@@
 Microbial abundance data typically contain many species observed infrequently (\textit{rare species}) and some which are observed a large number of times (\textit{abundant species}).
-In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{wang_2005,willis_2015}, motivating our regularised approach to estimation.
+In data with both many rare and many abundant species, richness estimation methods often have high variance \citep{wang_2005,willis_2015}, motivating our regularized approach to estimation.
 
@@
 we evaluate our proposals.
-\amy{Section \ref{sec:data_analysis} applies our methods to microbiome data on Lake Champlain...} Section \ref{sec:discussion} closes with an expanded discussion of our results and suggested directions for future work.
-Software implementing the methods is available in the \url{R} package \url{rre}, available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.
+In Section \ref{sec:data_analysis} we apply our methods to estimate microbial diversity in Lake Champlain, and Section \ref{sec:discussion} closes with a discussion of our results and suggested directions for future work.
+Software implementing the methods is available in the \url{R} package \url{rre} (\textit{regularized richness estimation}), available at \url{github.com/statdivlab/rre}, and code to reproduce the simulation results is available at \url{github.com/statdivlab/rre_sims}.
 
@@
 The classical model for estimating species abundance is a Poisson mixture model.
-Under this model the number of times we observe an species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
+Under this model the number of times we observe species $i$ is $X_i \sim \text{Poisson}(\Lambda)$ where $\Lambda$ is a random variable distributed according to some mixing distribution.
 Many mixing distributions have been proposed, including by  \cite{bulmer_1974,ord_1986,norris_1998} (see \cite{bunge_1993} for a comprehensive review).
-While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{fisher_1943}.
+While our proposal may be easily generalized to other mixing distributions, in this paper we consider the Gamma  distribution for $\Lambda$, first considered by \citet{greenwood_yule} and \citet{fisher_1943}.
+Note that we only observe $X_i | X_i > 0$, and therefore we are interested in estimating the parameters of a Poisson mixture model using \textit{zero-truncated} Poisson-mixed data.
 
-Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequencies} or \textit{frequency counts}.
+Let $f_k = \#\{i: X_i = k\}$ be the number of classes observed $k$ times, called the \textit{frequency counts}.
 Then the set $\{f_k\}_{k \geq 1}$ is a \textit{frequency count table}, a common way to represent species abundance data.
-Because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$, the species problem can be framed as predicting $f_0$ to obtain the species richness estimate $\hat{C} = \hat{f}_0 + c$.
+Because $C = f_0 + f_1 + f_2 + \dots = f_0 + c$, the species problem can be framed as predicting $f_0$ given $f_1, f_2, \ldots$ to obtain a species richness estimate $\hat{C} = \hat{f}_0 + c$.
 
@@
 %
-There exist both frequentist \citep{fisher_1943} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model.
+There exist both frequentist \citep{fisher_1943,chao_2002} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model.
 We will focus on frequentist maximum likelihood in this paper.
-\amy{introduce Gamma later?}
 
@@
 \end{align}
-By fixing $\lambda = 0$ we obtain the \textit{unpenalized log likelihood} for all replicates $j \in J$.
-We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate for based on tuning parameter $\lambda$, that is,
+By fixing $\lambda = 0$ we obtain the unpenalized log likelihood for the set $J$.
+We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate of $C$ based on tuning parameter $\lambda$, that is,
 \begin{equation}
@@
 
-We use root-mean-square error as the primary criterion for evaluating the performance of estimators in simulations.
-Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.
-Then we define
-\begin{equation}
- \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }
-\end{equation}
-
 To maximize the likelihood we will use the direct approach, rather than the profile and conditional approaches reviewed in Section \ref{sec:literature_review}.
-While it may be faster to use the profile or conditional approach, our investigations revealed that the results are not always the same in finite samples.
-Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed.
+While it may be faster to use the profile or conditional approach, the results are not always the same in finite samples.
+Since our objective is to evaluate approaches to tuning, we place priority on the accuracy of optimization over speed of optimization.
 
@@
 
-\section{Efficacy of Penalization}
+\section{Evaluating penalized species richness estimates}
 \label{sec:efficacy_sims}
@@
 \citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.
-We propose simulating data under the gamma-Poisson model and maximizing the objective function with a set of fixed $\lambda$ values.
-If there exists some $\lambda > 0$ where RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective for the given data generating mechanism.
+
+\amy{note that it's not a penalty, it's a bonus}
+
+We use square root mean square error (RMSE) as the primary criterion for evaluating the performance of estimators.
+Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.
+Then we define
+\begin{equation}
+ \text{RMSE}\left( \widehat{C}_{\lambda} \right) = \sqrt{ \frac{1}{n_{\text{sim}}} \sum_{s=1}^{n_{\text{sim}}} \left( \widehat{C}_{\lambda \langle s \rangle} - C\right)^2 }.
+\end{equation}
+In this simulation, if there exists some $\lambda > 0$ for which RMSE($\widehat{C}_{\lambda}$) is less than RMSE($\widehat{C}_0)$, we conclude that penalization is effective.
 
 We simulated using $C = 1000$, and $r \in \{6, 10, 14\}$.
-We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017}.
+We found that $\eta = (\alpha, \delta) = (10^{-1}, 10^{-1})$  provided a good fit to the data of  \citet{walsh2014restricted}, and $\eta = (10^{-2}, 10^{-5})$ provided a good fit to the data of \citet{tromas_2017} (see Section \ref{sec:data_analysis}).
 For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
@@
 Under each combination of $\eta$ and $r$ 100 simulations were completed.
-In Table \ref{tab:eta_intuition} we provide a few summary measures for frequency count tables generated under each choice of $\eta$.
+In Table \ref{tab:eta_intuition} we provide  summary measures for frequency count tables generated under each choice of $\eta$.
 This characterizes the $\eta$ values by the proportion of unobserved, rare or abundant species we would expect to see.
@@
 \begin{table}[t]
-\tblcaption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.
+\caption{The expected proportion of unobserved ($k=0$), singleton ($k=1$), rare ($k=1,2,3$), and abundant  ($k \geq 10$) species for 4 choices of $\eta$.
 We also give the expected frequency count of the most abundant species when $C = 1000$.
@@
   Proportion abundant ($\sum_{k=10}^{\infty} p_{\eta}(k)$) & 0.028 & 0.083 & 0.340 & 0.583 \\
-Expected max abundance ($\mathbb{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\
+Expected max abundance ($\mathbbm{E}\left[k_{\text{max}}\right]$) & $4.1 \times 10^{1}$ & $2.0 \times 10^{5}$ & $4.0 \times 10^{3}$ & $3.9 \times 10^{5}$ \\
    \hline
@@
 \begin{table}[h]
-\tblcaption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
+\caption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
 $\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
@@
 The results of the simulation are shown in Table \ref{tab:fixed_lambda_results}.
-For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) is found for some $\lambda > 0$.
+For every parameter choice a reduction in RMSE($\widehat{C}_{\lambda}$) was found for some $\lambda > 0$.
 Depending on simulation parameters the reduction varied from 22\% to 70\% compared to RMSE($\widehat{C}_{0}$) for the best $\lambda$.
@@
 For small values of $\lambda$ we acquire lower variance in the estimates while incurring some negative bias, and the RMSE improves as a result.
-Eventually the bias difference dominates and the RMSE increases as $\lambda \to \infty$.
+Eventually the bias term dominates and the RMSE increases as $\lambda \to \infty$.
 
@@
 
-We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different penalization parameters ($\lambda$) to minimize RMSE.
-Therefore a method which can tune the parameter based on a sample is warranted.
+We have established that penalization can improve richness estimates, but different abundance structures ($\eta$) require different values of $\lambda$ to minimize RMSE.
+We therefore develop methods to tune $\lambda$ based on a sample.
 We propose several novel methods in this section.
-Each is evaluated via simulation in Section \ref{sec:tuning_simulations}.
+Each is evaluated in Section \ref{sec:tuning_simulations}.
 
@@
 \end{equation}
-This method is fast and simple, making it an ideal baseline for comparison
+This method is fast and simple, making it an ideal baseline for comparison.
 
@@
 
-Ideally, our estimator for $C$ will have low variance, that is, there should be low variance in estimates from different resamples from the population.
+Since large variance in $C$ is a major concern in species richness estimation, ideally an estimator will have low variance.
+If this is the case, there should be low variance in estimates from equally sized subsets of $J$.
 % Method 1 is motivated by the belief that if we resample from the same population, an ideal $C$ estimator should have low variance.
-Exploiting the fact that we have replicate data, we repeatedly partition the replicates into two subsets and calculate two estimates.
+For Method 1, we exploit the fact that we have replicate data by repeatedly partitioning the replicates into two subsets and calculating two estimates.
 We then select the $\lambda$ which yields the lowest between-subset variance.
@@
 In our simulations we chose an equal split of the indices for each partition: $|T_1(l)| = |T_2(l)|$.
-The subsets of each partition are selected at random, so we could select the same subset split more than once.
+The subsets of each partition are selected at random, and we sample with replacement.
 We partition a total 10 times ($p = 10$).
-For example, if $r = 4$ then $T_1(1) = \{1,4\}$ and $T_2(1) = \{2, 3\}$ would be a valid subset selection.
-This holds in methods 2 and 4 as well, which also use partitioning.
+For example, if $r = 4$ then $T_1(1) = \{1,4\}$ and $T_2(1) = \{2, 3\}$ would be valid subsets.
+This approach to partitioning is also used in Methods 2 and 4.
 
@@
 
-In Method 2 we propose to repeatedly partition the data into two subsets, and evaluate the estimates based on the ``training'' subset using the likelihood based on the ``evaluation'' subset.
+In Method 2 we propose to repeatedly partition the data into subsets and evaluate the estimates based on the ``training'' subset using the likelihood based on the ``evaluation'' subset.
   % use one subset of the data to estimate $C$ and $\eta$.  Then we evaluate those estimates using the likelihood with the data from the other subset.  As before, w
@@
 \begin{equation}
-\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right) \label{eq:part_c_hats_method_2}
+\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; T(l) \right), \label{eq:part_c_hats_method_2}
 \end{equation}
-so that we have an estimate using only the training subset for each $\lambda$ value and each partition.
-A $\lambda$ value is selected by plugging those estimates into the unpenalized likelihood with $J = E(l)$:
+that is, $\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right)$ is the penalized maximum likelihood estimate evaluated on the \textit{training} subset.
+The $\lambda$ value which maximises the unpenalized likelihood calculated using the \textit{evaluation} subset is selected, and the average species richness estimate at this $\lambda$ is the estimated richness from Method 2:
 \begin{align}
@@
 Given that $c \sim$ Binomial $\left(C, 1- p_{\eta}(0)\right)$ and $(f_1, f_2, \ldots)|c \sim$  Multinomial $\left(c,  \frac{(p_{\eta}(1), p_{\eta}(2), \ldots)}{1-p_{\eta}(0)}\right)$, we have that
-% Using Equation \eqref{eq:likelihood} we can find the expected frequency counts $f_k$ given $C$ and $\eta$,
 \begin{equation}
- \mathbb{E}\left[f_k\right] = \mathbb{E}\left[\mathbb{E}\left[f_k|c \right]\right] = \mathbb{E}\left[ c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right] = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k).
+ \mathbbm{E}\left[f_k\right] = \mathbbm{E}\left[\mathbbm{E}\left[f_k|c \right]\right] = \mathbbm{E}\left[ c \frac{p_{\eta}(k)}{1-p_{\eta}(0)} \right] = C\left(1-p_{\eta}(0) \right) \frac{p_{\eta}(k)}{1-p_{\eta}(0)} = Cp_{\eta}(k).
 \end{equation}
-Then, given estimates $\widehat{C}$ and $\widehat{\eta}$, we consider a plug-in estimate for the expected frequency counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.
+Therefore, given estimates $\widehat{C}$ and $\widehat{\eta}$, we consider a plug-in estimate for the expected frequency counts: $\widehat{f_k} = \widehat{C}p_{\widehat{\eta}}(k)$.
 The usual $\chi^2$ goodness of fit statistic is then
@@
 \begin{equation}
-\left(\widehat{C}_{\lambda}, \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; \left\{1, \dots, r \right\} \right) \label{eq:c_hat_lambdas_method_3}
+\left(\widehat{C}_{\lambda}, \widehat{\eta}_{\lambda} \right) = \argmax_{C, \, \eta} \mathcal{O}_\lambda \left(C, \eta; \left\{1, \dots, r \right\} \right). \label{eq:c_hat_lambdas_method_3}
 \end{equation}
-$\widehat{C}_{[3]}$ is determined by evaluating the goodness of fit metric shown above for each choice of $\lambda$.
-$\widetilde{\lambda}_{[3]}$ is the $\lambda$ value which provides $\widehat{C}_{\lambda}$ estimates which fit the data best.
+$\widetilde{\lambda}_{[3]}$ is the $\lambda$ value with the best-fitting estimates of $C$ and $\eta$:
 \begin{align}
-\widetilde{\lambda}_{[3]} &= \argmin_{\lambda} \sum_{j=1}^r \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)^2}{\widehat{C}_{\lambda}p_{\widehat{\eta}_{\lambda}}(k)} \label{eq:selected_lambda_3} \\
+\widetilde{\lambda}_{[3]} &= \argmin_{\lambda} \sum_{j=1}^r \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda} p_{\widehat{\eta}_{\lambda}}(k) \right)^2}{\widehat{C}_{\lambda}p_{\widehat{\eta}_{\lambda}}(k)} \label{eq:selected_lambda_3}
+\end{align}
+and $\widehat{C}_{[3]}$ is the estimated value of $C$ at this choice of $\lambda$:
+\begin{align}
 \widehat{C}_{[3]} &= \widehat{C}_{\widetilde{\lambda}_{[3]}}
@@
 
-In Method 4 we return to the partitioning scheme of Method 2.  Rather than using the likelihood in the evaluation step, we hypothesize that the goodness-of-fit metric may be a better choice.
+In Method 4 we return to the partitioning scheme of Method 2.  Rather than using the likelihood in the evaluation step, we hypothesize that the goodness of fit metric may be a better choice.
 We partition the data $p$ times into a training set $T(l)$ and a evaluations set $E(l)$.
@@
 
-Similar to Method 2, this method generates training set-based estimates $\widehat{C}^{T(l)}_{\lambda}$, however, it evaluates these estimates using goodness-of-fit rather than likelihood maximisation.
+Similar to Method 2, this method generates training set-based estimates $\widehat{C}^{T(l)}_{\lambda}$, however, it evaluates these estimates using goodness of fit rather than likelihood maximisation.
 Compared to Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all the replicates are used in both steps.
@@
 We conjecture that the differing performance across $\eta$ is due to the differing rare species structures implied by the different $\eta$'s.
-For example $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
+For example $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbbm{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
 
 % Universally superior RMSE is the most desirable property for a method, but we can characterize the performance of the methods based on the simulation parameters as well.
-For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbb{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
+For example, $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbbm{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
 The RMSE for Method 2 under this $\eta$ is 14-24\% smaller than Method 0, and for Method 4 the RMSE is slightly better for 2 of the 3 $r$ values.
@@
 \begin{table}[ht]
-\tblcaption{RMSE$(\widehat{C})$ for Methods 0-4 based on a zero-truncated gamma-mixed Poisson data generating process.
+\caption{RMSE$(\widehat{C})$ for Methods 0-4 based on a zero-truncated gamma-mixed Poisson data generating process.
 $C = 1000$ is constant for all simulations.
@@
 
-Methods 3 and 4, which are both based on goodness-of-fit criteria, display different patterns depending on the $\eta$ and $r$ values of the simulation.
+Methods 3 and 4, which are both based on goodness of fit criteria, display different patterns depending on the $\eta$ and $r$ values of the simulation.
 % This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.
@@
 \begin{table}[t]
-\tblcaption{RMSE in $\widehat{C}_{[\text{method}]}$.
+\caption{RMSE in $\widehat{C}_{[\text{method}]}$.
 Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.
@@
 To illustrate the performance of our methods on ecological data, we estimate strain-level microbial diversity in Lake Champlain, a large eutrophic lake in Canada.
-We analyse data from \citet{tromas_2017}, considering samples from the littoral zone in the summer season of the same year as replicates.
+We analyze data from \citet{tromas_2017}, considering samples from the littoral zone in the summer season of the same year as replicates.
 This gives us 8 replicates from 2009, 6 replicates from 2010 and 6 replicates from 2011.
@@
 % \begin{table}[t]
-% \tblcaption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
+% \caption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
 $\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
@@
 For our investigations, we intentionally chose a likelihood optimization algorithm which was stable and exhaustive.
-However, refining the optimisation algorithm would be a valuable extension to our work.
+However, refining the optimization algorithm would be a valuable extension to our work.
 We also did not construct standard errors for our estimates, and the long computation times precluded the consideration of a bootstrapping approach.
-A faster optimisation algorithm would facilitate a resample-based standard error procedure.
+A faster optimization algorithm would facilitate a resample-based standard error procedure.
 

### 2019-08-08 rre_manuscript/paper.tex
edits throughout

@@
 \noindent%
-{\it Keywords: diversity, biodiversity, microbiome, regularization, maximum likelihood}
+{\it Keywords: diversity, microbiome, regularization, maximum likelihood, ecology}
 \vfill
@@
 The multinomial cell probabilities are proportional to $p_{\eta}(k)$.
-\citet{sanathanan_1977}'s conditional approach to optimization involves first maximising  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximising $L_b(C, \widehat{\eta}_c)$ over $C$.
+\citet{sanathanan_1977}'s conditional approach to optimization involves first maximizing  $L_c(\eta)$ to obtain a conditional abundance estimate $\widehat{\eta}_c$, then maximizing $L_b(C, \widehat{\eta}_c)$ over $C$.
 The conditional estimate of $C$ is then
@@
 Furthermore, for a large enough $\lambda$, $\hat{C}_\lambda = c$; that is, the penalized maximum likelihood estimate shrinks to the observed richness $c$.
-\citet{wang_2005} also consider two other penalty functions of similar form, which we do not discuss here except to note that our tuning parameter selection methods would equally apply.
+
+We note that $\log p_{\eta}(0) < 0,$ and so the addition of the ``penalty'' term in fact increases the objective function \eqref{eq:wang_lindsay} over the (unpenalized) likelihood $\ell(C,\eta)$.
+While technically smaller $p_{\eta}(0)$ adds a larger reward to the objective function (for a fixed $\lambda$), we refer to the term $ - \lambda \log p_{\eta}(0)$ as a penalty
+to be consistent with the terminology of \citet{wang_2005}.
+\citet{wang_2005} also consider two other penalty functions, which we do not discuss here except to note that our tuning parameter selection methods would equally apply to these other penalty functions.
 
@@
 Before evaluating potential tuning parameter selection methods, we first establish that penalization can improve richness estimation.
-\citet{wang_2005} established that it can improve estimates for $r = 1$, this needs to be verified for $r> 1$.
-
-\amy{note that it's not a penalty, it's a bonus}
+While \citet{wang_2005} established that penalization can improve estimates when $r = 1$, this needs to be investigated when $r> 1$.
 
 We use square root mean square error (RMSE) as the primary criterion for evaluating the performance of estimators.
-Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$ with  $\lambda$.
-Then we define
+Let $n_{\text{sim}}$ be the number of simulations completed for one choice of $(C, \eta, r)$, and $\widehat{C}_{\lambda \langle s \rangle}$ be the solution of Equation \eqref{eq:ccc_hat_lambda} obtained in simulation number $s$.
+We define
 \begin{equation}
@@
 
+\amy{check all captions}
+
 \begin{table}[h]
@@
 $\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
-$\widehat{C}_0$ is the $C$ estimate with $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the $C$ estimate using $\lambda = \lambda_{\text{opt}}$.
+$\widehat{C}_0$ is the estimate of $C$ when $\lambda = 0$, and $\widehat{C}_{\lambda_{\text{opt}}}$ is the estimate of $C$ when $\lambda = \lambda_{\text{opt}}$.
 \label{tab:fixed_lambda_results}}
@@
 
-In order to demonstrate the effect on $\widehat{C}_{\lambda}$ as we increase $\lambda$, we plot the results for two $\eta$ values in Figures \ref{fig:fixed_lambda} and \ref{fig:fixed_lambda_2}.
-For small values of $\lambda$ we acquire lower variance in the estimates while incurring some negative bias, and the RMSE improves as a result.
-Eventually the bias term dominates and the RMSE increases as $\lambda \to \infty$.
+To illustrate the bias-variance trade-off in this problem, we show the  effect on $\widehat{C}_{\lambda}$ of increasing $\lambda$ in Figure \ref{fig:fixed_lambda} (when $\eta = (10^{-1}, 10^{-1})$) and in Figure \ref{fig:fixed_lambda_2} (when $\eta = (10^{-2}, 10^{-5})$).
+In both cases, for small values of $\lambda$ we observe lower variance in the estimates while incurring some negative bias, and the RMSE improves.
+However, as $\lambda$ continues to increase, the bias term becomes large and the RMSE increases.
 
@@
 that is, $\left(\widehat{C}_{\lambda}^{ \ T(l)}, \widehat{\eta}_{\lambda}^{ \ T(l)} \right)$ is the penalized maximum likelihood estimate evaluated on the \textit{training} subset.
-The $\lambda$ value which maximises the unpenalized likelihood calculated using the \textit{evaluation} subset is selected, and the average species richness estimate at this $\lambda$ is the estimated richness from Method 2:
+The $\lambda$ value which maximizes the unpenalized likelihood calculated using the \textit{evaluation} subset is selected, and the average species richness estimate at $\widetilde{\lambda}_{[2]}$ is the estimated richness from Method 2:
 \begin{align}
@@
 % Therefore the statistic is a convergent sum, because the infinite sum is the tail of a distribution multiplied by a finite number $\widehat{C}$.
-Equation \eqref{eq:chi_sq_expanded} is useful in software implementation as we can make use of built-in CDF functions for the tail.
+Equation \eqref{eq:chi_sq_expanded} is useful in software implementation as we can make use of precomputed tail probabilities $\sum_{k_{max}+1}^{\infty} p_{\widehat{\eta}}(k)$.
 %Nonetheless we use Equation \eqref{eq:chi_sq_compact} below in extending to replicates for more concise notation.
@@
 
-%
-%
-%
+An advantage of Method 3 that is not shared by Methods 1, 2 and 4 is that it can be used when $r=1$, that is, when no repeated measurements are available.
+
 \subsection{Method 4: Cross-validated goodness of fit}
 
-In Method 4 we return to the partitioning scheme of Method 2.  Rather than using the likelihood in the evaluation step, we hypothesize that the goodness of fit metric may be a better choice.
-We partition the data $p$ times into a training set $T(l)$ and a evaluations set $E(l)$.
+\amy{up to here}
+
+In Method 4 we return to the partitioning scheme of Method 2, but rather than using the likelihood in the evaluation step, we hypothesize that the goodness of fit metric may be a better choice.
+We partition the data $p$ times, indexing the partitions by $l$. For each partition we have a training set $T(l)$ and a evaluation set $E(l)$.
 For all $\lambda \in \lambdagrid$ and $l \in \{1, \dots , p \}$, we generate estimates exactly as in Method 2:
@@
 \end{equation}
-To select $\lambda$ we evaluate the goodness of fit metric using only the evaluation subset data, $j \in E(l)$.  The $\lambda$ which produces the best fitting estimates is $\widetilde{\lambda}_{[4]}$:
+To select $\lambda$ we evaluate the goodness of fit metric using only the evaluation subset data, $j \in E(l)$.
+The $\lambda$ which produces the best fitting estimates on the evaluation subset is $\widetilde{\lambda}_{[4]}$:
 \begin{align}
-\widetilde{\lambda}_{[4]} &= \argmin_{\lambda} \sum_{l = 1}^p \sum_{j \in E(l)} \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda}^{T(l)} p_{\widehat{\eta}_{\lambda}^{T(l)}}(k) \right)^2}{\widehat{C}_{\lambda}^{T(l)}p_{\widehat{\eta}_{\lambda}^{T(l)}}(k)} \\
+\widetilde{\lambda}_{[4]} &= \argmin_{\lambda} \sum_{l = 1}^p \sum_{j \in E(l)} \sum_{k=1}^{\infty} \frac{ \left( f_{kj} - \widehat{C}_{\lambda}^{T(l)} p_{\widehat{\eta}_{\lambda}^{T(l)}}(k) \right)^2}{\widehat{C}_{\lambda}^{T(l)}p_{\widehat{\eta}_{\lambda}^{T(l)}}(k)}
+\end{align}
+and $\widehat{C}_{[4]}$ is the mean estimate of $C$ at  $\widetilde{\lambda}_{[4]}$, averaged all partitions $l$:
+\begin{align}
 \widehat{C}_{[4]} &= \frac{1}{p} \sum_{l=1}^p \widehat{C}_{\widetilde{\lambda}_{[4]}}^{T(l)}
@@
 
-Similar to Method 2, this method generates training set-based estimates $\widehat{C}^{T(l)}_{\lambda}$, however, it evaluates these estimates using goodness of fit rather than likelihood maximisation.
-Compared to Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all the replicates are used in both steps.
+Similar to Method 2, this method generates training set-based estimates $\widehat{C}^{T(l)}_{\lambda}$, however, it evaluates these estimates using goodness of fit rather than likelihood maximization.
+Compared to Method 3, this method uses each replicate to either generate an estimate or evaluate the fit, while in Method 3 all replicates are used in both steps.
 
@@
 We simulate zero-truncated gamma-mixed Poisson data using the same parameters used in Section \ref{sec:efficacy_sims}.
-Based on the results of Section \ref{sec:efficacy_sims} we know that estimation can be improved on this set of parameters through the use of a fixed penalization parameter.
+Based on the results of Section \ref{sec:efficacy_sims} we know that estimation can be improved through penalization, at least for the choices of $\eta$ that we propose to simulate from.
 The purpose of this section is to determine if any method can reliably select a $\lambda$ that reduces RMSE compared to unpenalized maximum likelihood estimation.
+In each simulation below, we select the largest value in $\lambdagrid$ by doubling the optimal value of $\lambda$ found in Table \ref{tab:fixed_lambda_results}.
 
@@
 
-We present two simulations below, and in each it is important to select an appropriate $\lambdagrid$.
-% In particular our grid needs to be large enough to reveal the negative aspects of methods which over-penalize.
+% We present two simulations below, and in each it is important to select an appropriate $\lambdagrid$.
+% % In particular our grid needs to be large enough to reveal the negative aspects of methods which over-penalize.
 For instance, if $\lambdagrid = \{0, \dots , 50\}$ and the optimal value is $\lambda = 45$, a method which deterministically selects the maximum value in $\lambdagrid$ (an extreme over-penalizer) would perform well.
 This would lead us to spuriously conclude this hypothetical over-penalizer is effective.
-In Section \ref{sec:efficacy_sims}  we established the approximate optimal values for each simulation setting (Table \ref{tab:fixed_lambda_results}).
-% One of the important foundations laid in Section \ref{sec:efficacy_sims} is that we now know approximately what the optimal values are for each simulation setting (Table \ref{tab:fixed_lambda_results}).
-We select a large enough grid by extending to at least double the optimal value in both simulations presented.
+% In Section \ref{sec:efficacy_sims}  we established the approximate optimal values for each simulation setting (Table \ref{tab:fixed_lambda_results}).
+% % One of the important foundations laid in Section \ref{sec:efficacy_sims} is that we now know approximately what the optimal values are for each simulation setting (Table \ref{tab:fixed_lambda_results}).
+% We select a large enough grid by extending to at least double the optimal value in both simulations presented.
 
-\subsection{Primary simulation -- TODO Rename section}
+\subsection{Comparison of methods' performance}
 \label{sec:tuning_simulation_1}
@@
 We know that the optimal $\lambda$ choice for these $r$ and $\eta$ are between 10 and 20, so we chose $\lambdagrid = \{0, 5, 10, \dots 60\}$ for this simulation.
-Methods 0-4 are all evaluated over the same random draws for each parameter choice.
+Methods 0-4 are all evaluated over the same random draws for each parameter choice using the \texttt{R}
+package \texttt{simulator} \citep{Bien:2016tg}.
 
@@
 
-We note some differences between the performance of the methods for different $\eta.$ When $\eta = (10^{-2},10^{-5})$, both Methods 2 also outperforms Method 0 for $r \in \{6, 10, 14\}$, while Method 2 does not outperform Method 0 for any $r$ when $\eta = (10^{-1},10^{-1})$.
+We note some differences between the performance of the methods for different $\eta.$ When $\eta = (10^{-2},10^{-5})$, Method 2 also outperforms Method 0 for $r \in \{6, 10, 14\}$, while Method 2 does not outperform Method 0 for any $r$ when $\eta = (10^{-1},10^{-1})$.
 Method 4 outperforms Method 0 for $r \in \{6, 14\}$ when $\eta = (10^{-2},10^{-5})$, but never when $\eta = (10^{-1},10^{-1})$.
@@
 
-[Method 1 has better performance only for $r = 6$ on both $\eta$ choices.
+% [Method 1 has better performance only for $r = 6$ on both $\eta$ choices.
 In general, we would expect that any truly effective method would improve relative to Method 0 as $r$ increases, as this means more data was collected.
@@
 Recall that as we increase $\lambda$, $\widehat{C}_\lambda$ monotonically decreases.
-Therefore we can discuss whether a method is over-penalizing (selecting $\widetilde{\lambda}$ which is larger than optimal) or under-penalizing (selecting $\widetilde{\lambda}$ which is smaller than optimal) without additional information: if $\widehat{C} > C$, then the method has under-penalized while if $\widehat{C} < C$ the method has over-penalized.
+Therefore we can discuss whether a method is on average over-penalizing (selecting $\widetilde{\lambda}$ which is larger than optimal) or under-penalizing (selecting $\widetilde{\lambda}$ which is smaller than optimal): if $\widehat{C} > C$, then the method has under-penalized while if $\widehat{C} < C$ the method has over-penalized.
 
@@
 
-\subsection{Secondary simulation}
+\subsection{Performance over wider range of parameter values}
 \label{sec:tuning_simulation_2}
@@
 
-In this simulation we will fix $C = 1000$, $r \in \{6, 10, 14\}$ and perform 100 simulations for each combination of $\eta$ and $r$.
-We let $\eta \in \{ (10^{-1}, 10^{-3}), (10^{-1}, 10^{-5}) \}$ and only consider methods 0 and 3.
+In this simulation we will fix $C = 1000$, $r \in \{6, 10, 14\}$ and perform 100 simulations for two additional choices of $\eta$.
+We let $\eta \in \{ (10^{-1}, 10^{-3}), (10^{-1}, 10^{-5}) \}$ and only consider Methods 0 and 3, based on the results of the previous section.
 To account for the fact that a larger $\lambda$ was optimal for these $\eta$ values (see Table \ref{tab:fixed_lambda_results}) we use $\lambdagrid = \{0,10,20, \dots 140 \}$.
@@
 \begin{table}[t]
-\caption{RMSE in $\widehat{C}_{[\text{method}]}$.
+\caption{RMSE for Methods 0 and 3.
 Simulation parameters are grouped by $\eta$ and $r$, with $C = 1000$ fixed throughout.
-Under each $\eta, r$ combination 100 simulations were run.
+For each $\eta$ and $r$ combination, 100 simulations were run.
 If Method 3 has better performance than Method 0 is has a \textcolor{blue!50}{blue} background and it is \textbf{bolded}.
-This is redundant but consistent with Table \ref{tab:tuning_sim_1}.
 \label{tab:tuning_sim_2}}
@@
 \cline{2-7}
-\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (0.1,0.001)$} & \multicolumn{3}{|c|}{$\eta = (0.1,0.00001)$} \\
+\multicolumn{1}{c}{} & \multicolumn{3}{|c|}{$\eta = (10^{-1},10^{-3})$} & \multicolumn{3}{|c|}{$\eta = (10^{-1},10^{-5})$} \\
 \cline{2-7}
@@
 In Figure \ref{fig:tuning_sim_2} we display boxplots of the $C$ estimates over all simulations.
-For $\eta = (10^{-2}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is very large for a few simulations.
+For $\eta = (10^{-1}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is very large for a few simulations.
 Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species.
 Therefore, this is a excellent example of a simulation structure which will cause the instability problem in maximum likelihood estimation discussed in Section \ref{sec:literature_review}.
-Method 3 succeeds on this $\eta$ primarily by selecting lower estimates on these problematic simulations.
+Method 3 succeeds on this $\eta$ by selecting lower estimates.
 
 \begin{figure}[t]
-\caption{Simulation results for methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
+\caption{Simulation results for Methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
 \label{fig:tuning_sim_2}}
@@
 
-\subsection{Simulation Conclusions}
+While Table \ref{tab:tuning_sim_2} suggests that Method 3 has larger RMSE than Method 0, Figure \ref{fig:tuning_sim_2} suggests that when  $\eta = (10^{-1}, 10^{-3})$, the estimates produced by the methods are very similar.
+The advantage of Method 3 over Method 0 when $\eta = (10^{-1}, 10^{-5})$ for increasing $r$ is also clearer, since we see that Method 3 reduces the incidence of outlying estimates.
 
@@
 \label{sec:data_analysis}
+
+\subsection{Estimating microbial richness in Lake Champlain}
 
@@
 
-Our estimates from each method are all at least 10 times as high as the observed richness, indicating a significant fraction of rare and unobserved species.
+Our estimates from each method are all at least 10 times as high as the observed richness, indicating that a significant fraction of species were unobserved.
 Based on our previous results, estimation error is lowest for Methods 0 and 3 when there are many rare species.
@@
 
+\subsection{Estimating global human host-associated microbial richness}
+
+We also applied our method to estimate the species-level diversity of human host associated microbes.
+\cite{pasolli2019extensive} assembled 4,930 species-level genome bins (SGBs) using publicly available shotgun metagenomic data, but we expect that many SGBs were not observed due to undersampling and challenges in genome assembly.
+The frequency counts of each SGB are available at \cite[Table S4]{pasolli2019extensive}.
+
+In this dataset $r=1$, and so it is not possible to sample split replicate frequency count tables.
+Therefore, only Methods 0 and Method 3 can be applied.
+With $\lambdagrid = \{0, 0.25, \ldots, 1100\}$ we found that Method 3 selected $\lambda = 913.5$.
+However, $\widehat{C}_{[3]} = 98600$
+
 \section{Discussion and Conclusion}
@@
 
-In this paper we discussed an extension of the penalized maximum likelihood procedure of \citet{wang_2005} for species richness estimation to data with biological replicates, and proposed several methods for tuning the penalization parameter.
-We demonstrated that penalization can reduce estimation error (defined as square root mean squared error) when analyzing replicate data.
-Our goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
-On microbial abundance data \amy{TODO}
-
-Our results from Section \ref{sec:fixed_lambda} hint at a positive correlation between $r$ and the optimal choice of $\lambda$.
-To investigate this, we investigated whether the optimal choice of $\lambda$ remained constant for penalties of the form $-h(J)\lambda \log p_{\eta}(0)$ (instead of $-\lambda \log p_{\eta}(0)$; see Equation \eqref{eq:objective}).
-We tested $h(J) = J$ and $h(J) = \sqrt{J}$, but found that the trend in optimal $\lambda$ is not so simple.
-We leave further investigation into how the optimal $\lambda$ varies with $J$ to future work.
+\subsection{Limitations and future work}
 
@@
 The gamma-Poisson model is common for modeling microbiome data \citep{deseq2,mcmurdie2013phyloseq}, and for this reason we focused on it in our investigation.
-However, our most promising method, Method 3, is also amenable to other modeling choices provided that there is some way to calculate the fitted counts $\widehat{f_k}$.
+However, the best performing method, Method 3, is also amenable to other modeling choices provided that there is some way to calculate the fitted counts $\widehat{f_k}$.
 We leave investigation of tuning parameter selection under different models to future work.
 
+Our results from Section \ref{sec:fixed_lambda} hint at a possible positive correlation between $r$ and the optimal choice of $\lambda$.
+We investigated whether the optimal choice of $\lambda$ remained constant for penalties of the form $-h(r)\lambda \log p_{\eta}(0)$ (instead of $-\lambda \log p_{\eta}(0)$; see Equation \eqref{eq:objective}).
+We tested $h(r) = r$ and $h(r) = \sqrt{r}$, but found that the trend in optimal $\lambda$ is not so simple.
+We leave further investigation into how the optimal $\lambda$ varies with $r$ to future work.
+Understanding of $h(r)$ would allow us to consider unequally sized evaluation and training partitions for Methods 2 and 4.
+
 For our investigations, we intentionally chose a likelihood optimization algorithm which was stable and exhaustive.
-However, refining the optimization algorithm would be a valuable extension to our work.
 We also did not construct standard errors for our estimates, and the long computation times precluded the consideration of a bootstrapping approach.
-A faster optimization algorithm would facilitate a resample-based standard error procedure.
+Refining the optimization algorithm would be a valuable extension, and a faster optimization algorithm would facilitate a resample-based variance estimation procedure.
+
+\subsection{Conclusions}
+
+In this paper we discussed an extension of the penalized maximum likelihood procedure of \citet{wang_2005} for species richness estimation to data with biological replicates, and proposed several methods for tuning the penalization parameter.
+We demonstrated that penalization can reduce estimation error (defined as square root mean squared error) when analyzing replicate data.
+We found that tuning the penalization parameter is challenging, but that a goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
+However, on a microbial abundance dataset with many rare species, it selected the same estimate as unregularized maximum likelihood estimation.
+
+Our investigation highlights the challenges of selecting tuning parameters in the absence of a ground truth, since we never observe $C$.
+However, even in the absence of information with which to calibrate $\lambda$, we showed that
+with a parametric model for species abundance data we can use goodness of fit in conjunction with maximum likelihood to select $\lambda$.
+While simulation results suggest that regularisation can improve estimation, in practice we found no difference between regularised and unregularised estimates.
 

### 2019-08-08 rre_manuscript/paper.tex
mike's inline suggestions

@@
 We also applied our method to estimate the species-level diversity of human host associated microbes.
-\cite{pasolli2019extensive} assembled 4,930 species-level genome bins (SGBs) using publicly available shotgun metagenomic data, but we expect that many SGBs were not observed due to undersampling and challenges in genome assembly.
+\cite{pasolli2019extensive} assembled $c=4,930$ species-level genome bins (SGBs) using publicly available shotgun metagenomic data, but we expect that many SGBs were not observed due to undersampling and challenges in genome assembly.
 The frequency counts of each SGB are available at \cite[Table S4]{pasolli2019extensive}.
@@
 However, $\widehat{C}_{[3]} = 98600$
+
+We found that $\widehat{C}_{[0]} = 420056$ with $(\hat{\alpha}, \hat{\delta}) = (0.00234, 0.00641).$
 

### 2019-08-08 rre_manuscript/paper.tex
final numbers for Pasolli analysis

@@
 \cite{pasolli2019extensive} assembled $c=4,930$ species-level genome bins (SGBs) using publicly available shotgun metagenomic data, but we expect that many SGBs were not observed due to undersampling and challenges in genome assembly.
-The frequency counts of each SGB are available at \cite[Table S4]{pasolli2019extensive}.
+The frequency counts of each SGB are available at \citet[Table S4]{pasolli2019extensive}.
 
@@
 Therefore, only Methods 0 and Method 3 can be applied.
-With $\lambdagrid = \{0, 0.25, \ldots, 1100\}$ we found that Method 3 selected $\lambda = 913.5$.
-However, $\widehat{C}_{[3]} = 98600$
+We found that $\widehat{C}_{[0]} = 420056$ with $(\hat{\alpha}, \hat{\delta}) = (0.00234, 0.00641).$ In contrast, $\widehat{C}_{[3]} = 163587$ with $\tilde{\lambda}_{[3]} = 905$ and $(\hat{\alpha}, \hat{\delta}) = (0.006045614, 0.006352191).$
+% With $\lambdagrid = \{0, 0.25, \ldots, 1100\}$ we found that Method 3 selected $\lambda = 913.5$. However, $\widehat{C}_{[3]} = 98600$
+% We found that $\widehat{C}_{[3]} = 163587$ with $\tilde{\lambda}_{[3]} = 905$ and $(\hat{\alpha}, \hat{\delta}) = (0.006045614, 0.006352191).$ % GOF = 11196564679
 
-We found that $\widehat{C}_{[0]} = 420056$ with $(\hat{\alpha}, \hat{\delta}) = (0.00234, 0.00641).$
+\amy{TODO: Discussion}
 

### 2019-08-08 data_analysis/pasolli/pasolli_investigation.R
add pasolli data analysis script

@@ -1,0 +1,167 @@
+library(tidyverse)
+pasolli_et_al <- openxlsx::read.xlsx("https://ars.els-cdn.com/content/image/1-s2.0-S0092867419300017-mmc4.xlsx", sheet = 2)
+ft <- pasolli_et_al$`#.Samples` %>% make_frequency_count_table
+ft %>% head
+m0 <- rre::unregularized_mle(list(ft))
+# m1 <- rre::minimum_subset_distance(list(ft))
+m3 <- rre::gof_criterion(list(ft))
+# m4 <- rre::cv_replicates(list(ft))
+
+
+m3$best
+m0$best
+
+m3
+m3_b <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(20, 100, by= 5))
+m3_b
+
+m3_c <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(100, 1100, by= 200))
+m3_c
+
+
+m3_d <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(700, 1100, by= 50))
+m3_d
+
+m3_e <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(850, 950, by= 10))
+m3_e
+
+m3_f <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(900, 920, by= 5))
+m3_f
+
+m3_g <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(905, 915, by= 1))
+m3_g
+
+m3_h <- rre::gof_criterion(list(ft),
+                           lambda_vec = seq(912, 914, by= 0.25))
+m3_h
+
+m3_i <- rre::gof_criterion(list(ft),
+                           starts = data.frame(alpha = c(0.0102, 0.0123),
+                                               delta = c(0.00647, 0.00650)),
+                           lambda_vec = 913.5)
+m3_i
+
+
+m3_j <- rre::gof_criterion(list(ft),
+                           starts = data.frame(alpha = c(0.0102, 0.0123, 0.0102, 0.0123),
+                                               delta = c(0.00647, 0.00647, 0.00650, 0.00650)),
+                           lambda_vec = seq(912, 914, by= 0.5))
+m3_j # 90712, but on boundary with lambda = 914
+
+m0$best # MLE is 98600
+ft %>% breakaway::sample_richness # 4930
+
+### noooo this is close to the multiplier!
+
+expand.grid("alpha" = unique(c(0.0102, 0.0123, 0.005)),
[…]

### 2019-08-10 data_analysis/data_analysis_expanded.R
start for expanded analysis

@@ -1,0 +1,93 @@
+library(magrittr)
+library(rre)
+setwd("/Users/alex/Dropbox/classGrad/IS_Wil/rre_sims/data_analysis/")
+
+lakes <- readr::read_tsv("./lakes_data.txt") %>% as.data.frame()
+lakes_metadata <- readr::read_tsv("./lakes_metadata.txt") %>% as.data.frame()
+
+# Deciding which samples to extract:
+lakes_metadata %>% 
+  dplyr::group_by(Site, Years, Period) %>%
+  dplyr::summarize(number = dplyr::n()) %>%
+  dplyr::arrange(Site, Period, Years) %>%
+  print(n = 500)
+
+# Based on the above we want the following rows to obtain a series of samples:
+lakes_metadata %>%
+  dplyr::filter(Years %in% seq(2009, 2011), Period == "Summer", Site == "Littoral") %>%
+  dplyr::arrange(Site, Years)
+
+# sort(colnames(lakes))
+
+# goal is to grab a set of samples from the same year and site
+get_samples <- function(yr, zone, per = NULL, month = NULL) {
+  quo_period <- dplyr::enquo(per)
+  sample_list <- lakes_metadata %>%
+    dplyr::filter(Site == zone) %>%
+    dplyr::filter(Years == yr)
+  if (!is.null(per)) {
+    sample_list %<>%  dplyr::filter(Period == per) 
+  }
+  if (!is.null(month)) {
+    sample_list %<>%  dplyr::filter(Months == month) 
+  }
+  sample_list %<>% dplyr::select(SampleID) %>% unlist
+  tables <- lakes[,colnames(lakes) %in% sample_list] %>%
+    lapply(., rre::make_frequency_count_table)
+  return(tables)
+}
+
+# Note that one of the metadata files matches no lakes sample ID:  12.07.11.1
+list_of_samples <- mapply(get_samples, seq(2009, 2011), "Littoral", "Summer")
+
+list_of_samples %>% lapply(.,rre::get_cc_max) 
+
+s1_m0 <- rre::unregularized_mle(list_of_samples[[1]], 
+                                multiplier = 200,
+                                c_seq_len = 200) # Chat = 23904 which is 42*cc
+s1_m0$best
+
+s2_m0 <- rre::unregularized_mle(list_of_samples[[2]], 
+                                multiplier = 200,
+                                c_seq_len = 200) # Chat = 48500 which is 84*cc
+s2_m0$best
+
[…]

### 2019-08-13 data_analysis/data_analysis_expanded.Rmd
Expanded data analysis update

---
title: "Data Analysis Expanded"
author: "Alex Paynter"
date: "August 9, 2019"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, include = T, cache = T, message = F, results = 'hide')
knitr::opts_knit$set(root.dir = here::here("data_analysis"))
```

```{r load_data}
library(magrittr)
library(rre)
library(dplyr);
#setwd("/Users/alex/Dropbox/classGrad/IS_Wil/rre_sims/data_analysis/")

lakes <- readr::read_tsv("./lakes_data.txt") %>% as.data.frame()
lakes_metadata <- readr::read_tsv("./lakes_metadata.txt") %>% as.data.frame()

# goal is to grab a set of samples from the same year and site
get_samples <- function(yr, zone, per = NULL, month = NULL) {
  quo_period <- dplyr::enquo(per)
  sample_list <- lakes_metadata %>%
    dplyr::filter(Site == zone) %>%
    dplyr::filter(Years == yr)
  if (!is.null(per)) {
    sample_list %<>%  dplyr::filter(Period == per) 
  }
  if (!is.null(month)) {
    sample_list %<>%  dplyr::filter(Months == month) 
  }
  sample_list %<>% dplyr::select(SampleID) %>% unlist
  tables <- lakes[,colnames(lakes) %in% sample_list] %>%
    lapply(., rre::make_frequency_count_table)
  return(tables)
}

#test <- rre::unregularized_mle(list_of_samples[[1]], 
#                               multiplier = 2,
#                               c_seq_len = 2)

report_results_2 <- function(method_output) {
  best <- method_output[["best"]]
  c_hat <- best$ccc
  m <- c_hat/(best$cc_max)
  cat(paste0("C_hat: ", c_hat, "\n\n", round(m,2), " times greater than observed richness \n\nSelected lambda: ", best$selected_lambda, "\n\n"))
}

# Note that one of the metadata files matches no lakes sample ID:  12.07.11.1
list_of_samples <- mapply(get_samples, seq(2009, 2011), "Littoral", "Summer")

big_starts = data.frame(alpha = c(0.1, 0.01, 0.01, 0.001), 
                        delta = c(0.1, 0.01, 0.0001, 0.001))
```

## Method 0

```{r s1_m0_a}
s1_m0_a <- rre::unregularized_mle(list_of_samples[[1]], 
                                  multiplier = 200,
                                  c_seq_len = 200,
                                  starts = big_starts)
```
```{r s1_m0_a_print, results = 'asis'}
s1_m0_a %>% report_results_2(.)
```
<br>
<br>

```{r s2_m0_a}
s2_m0_a <- rre::unregularized_mle(list_of_samples[[2]], 
                                  multiplier = 200,
                                  c_seq_len = 200,
                                  starts = big_starts)
```
```{r s2_m0_a_print, results = 'asis'}
s2_m0_a %>% report_results_2(.)
```
<br>
<br>

```{r s3_m0_a}
s3_m0_a <- rre::unregularized_mle(list_of_samples[[3]],
                                  multiplier = 200,
                                  c_seq_len = 200,
                                  starts = big_starts)
```
```{r s3_m0_a_print, results = 'asis'}
s3_m0_a %>% report_results_2(.)
```
<br>
<br>

<br><br>

## Method 1

### Sample 1

```{r s1_m1_a}
# big grids and few partitions to start:
set.seed(12039)
s1_m1_a <- rre::minimum_subset_distance(list_of_samples[[1]],
                                        partitions = 5,
                                        lambda_vec = seq(0,1000,by=200),
                                        multiplier = 100,
                                        c_seq_len = 20)
```
```{r s1_m1_a_print, results = 'asis'}
s1_m1_a %>% report_results_2(.)
```
<br><br>

<br><br>

### Sample 2

```{r s2_m1_a}
# big grids and few partitions to start:
set.seed(2345629)
s2_m1_a <- rre::minimum_subset_distance(list_of_samples[[2]],
                                        partitions = 5,
                                        lambda_vec = seq(0,1000,by=200),
                                        multiplier = 100,
                                        c_seq_len = 50)
```
```{r s2_m1_a_print, results = 'asis'}
s2_m1_a %>% report_results_2(.)
```
<br><br>

<br><br>

### Sample 3

```{r s3_m1_a}
# big grids and few partitions to start:
set.seed(45698)
s3_m1_a <- rre::minimum_subset_distance(list_of_samples[[3]],
                                        partitions = 5,
                                        lambda_vec = seq(0,1000,by=200),
                                        multiplier = 100,
                                        c_seq_len = 50)
```
```{r s3_m1_a_print, results = 'asis'}
s3_m1_a %>% report_results_2(.)
```
<br><br>

<br><br>

## Method 3

### Sample 1

```{r s1_m3_a}
s1_m3_a <- rre::gof_criterion(list_of_samples[[1]], 
                              lambda_vec = seq(0,1000,by=100),
                              multiplier = 200,
                              c_seq_len = 10)
```
```{r s1_m3_a_print, results = 'asis'}
s1_m3_a %>% report_results_2(.)
```
<br>
<br>

```{r s1_m3_b}
# Zero in on lambda, use a more dense C grid:
s1_m3_b <- rre::gof_criterion(list_of_samples[[1]], 
                              lambda_vec = seq(300,500,by=10),
                              multiplier = 200,
                              c_seq_len = 20)
```
```{r s1_m3_b_print, results = 'asis'}
s1_m3_b %>% report_results_2(.)
```
<br>
<br>

```{r s1_m3_c}
s1_m3_c <- rre::gof_criterion(list_of_samples[[1]], 
                              lambda_vec = seq(450,530,by=5),
                              multiplier = 200,
                              c_seq_len = 50,
                              starts = big_starts)
```
```{r s1_m3_c_print, results = 'asis'}
s1_m3_c %>% report_results_2(.)
```
<br>
<br>

```{r s1_m3_d}
# Feed in more alpha and delta, narrow lambdagrid more.
#s1_m3_d <- rre::gof_criterion(list_of_samples[[1]], 
#                              lambda_vec = seq(505,525,by=2),
#                              multiplier = 100,
#                              c_seq_len = 50,
#                              starts = data.frame(alpha = c(0.1, 0.01, 0.01, 0.001), delta = c(0.1, 0.01, 0.0001, 0.001)))
```
```{r s1_m3_d_print, results = 'asis'}
#s1_m3_d %>% report_results_2(.)
```
<br>
<br>

<br><br>

### Sample 2
```{r s2_m3_a}
s2_m3_a <- rre::gof_criterion(list_of_samples[[2]], 
                              lambda_vec = seq(0,1000,by=100),
                              multiplier = 200,
                              c_seq_len = 10)
```
```{r s2_m3_a_print, results = 'asis'}
s2_m3_a %>% report_results_2(.)
```
<br>
<br>

```{r s2_m3_b}
# Zero in on lambda, use a more dense C grid:
s2_m3_b <- rre::gof_criterion(list_of_samples[[2]], 
                              lambda_vec = seq(0,200,by=10),
                              multiplier = 200,
                              c_seq_len = 20)
```
```{r s2_m3_b_print, results = 'asis'}
s2_m3_b %>% report_results_2(.)
```
<br>
<br>

```{r s2_m3_c}
# Zero in on lambda, use a more dense C grid:
s2_m3_c <- rre::gof_criterion(list_of_samples[[2]], 
                              lambda_vec = seq(120,200,by=5),
                              multiplier = 200,
                              c_seq_len = 50)
```
```{r s2_m3_c_print, results = 'asis'}
s2_m3_c %>% report_results_2(.)
```
<br>
<br>

```{r s2_m3_d}
# Selected lambda was on the edge of previous grid, lets try again:
s2_m3_d <- rre::gof_criterion(list_of_samples[[2]], 
                              lambda_vec = seq(150,300,by=10),
                              multiplier = 100,
                              c_seq_len = 50,
                              starts = big_starts)
```
```{r s2_m3_d_print, results = 'asis'}
s2_m3_d %>% report_results_2(.)
```
<br>
<br>

<br><br>

### Sample 3
```{r s3_m3_a}
s3_m3_a <- rre::gof_criterion(list_of_samples[[3]], 
                              lambda_vec = seq(0,1000,by=100),
                              multiplier = 200,
                              c_seq_len = 10)
```
```{r s3_m3_a_print, results = 'asis'}
s3_m3_a %>% report_results_2(.)
```
<br>
<br>

```{r s3_m3_b}
# Zero in on lambda, use a more dense C grid:
s3_m3_b <- rre::gof_criterion(list_of_samples[[3]], 
                              lambda_vec = seq(0,100,by=10),
                              multiplier = 200,
                              c_seq_len = 20)
```
```{r s3_m3_b_print, results = 'asis'}
s3_m3_b %>% report_results_2(.)
```
<br>
<br>

```{r s3_m3_c}
s3_m3_c <- rre::gof_criterion(list_of_samples[[3]], 
                              lambda_vec = seq(50,130,by=10),
                              multiplier = 200,
                              c_seq_len = 50)
```
```{r s3_m3_c_print, results = 'asis'}
s3_m3_c %>% report_results_2(.)
```
<br>
<br>

```{r s3_m3_d}
# Selected lambda was on the edge of previous grid, lets try again:
s3_m3_d <- rre::gof_criterion(list_of_samples[[3]], 
                              lambda_vec = seq(100, 200, by=10),
                              multiplier = 100,
                              c_seq_len = 50,
                              starts = big_starts)
```
```{r s3_m3_d_print, results = 'asis'}
s3_m3_d %>% report_results_2(.)
```

<br><br>

### 2019-08-14 data_analysis/data_analysis_expanded.Rmd
Methods 2/4 started

@@
 
+<br><br>
+
+## Method 2
+
+### Sample 1
+
+```{r s1_m2_a}
+# big grids and few partitions to start:
+set.seed(12039)
+s1_m2_a <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s1_m2_a_print, results = 'asis'}
+s1_m2_a %>% report_results_2(.)
+```
+
+<br><br>
+
+### Sample 2
+
+```{r s2_m2_a}
+# big grids and few partitions to start:
+set.seed(12023439)
+s2_m2_a <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s2_m2_a_print, results = 'asis'}
+s2_m2_a %>% report_results_2(.)
+```
+
+<br><br>
+
+```{r s3_m2_a}
+# big grids and few partitions to start:
+set.seed(2342039)
+s3_m2_a <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s3_m2_a_print, results = 'asis'}
+s3_m2_a %>% report_results_2(.)
+```
+
+<br><br>
 <br><br>
@@
 <br><br>
+
+## Method 4
+
+### Sample 1
+
+```{r s1_m4_a}
+# big grids and few partitions to start:
+set.seed(87124)
+s1_m4_a <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s1_m4_a_print, results = 'asis'}
+s1_m4_a %>% report_results_2(.)
+```
+
+<br><br>
+
+### Sample 2
+
+```{r s2_m4_a}
+# big grids and few partitions to start:
+set.seed(23456)
+s2_m4_a <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s2_m4_a_print, results = 'asis'}
+s2_m4_a %>% report_results_2(.)
+```
+
+<br><br>
+
+```{r s3_m4_a}
+# big grids and few partitions to start:
+set.seed(7654)
+s3_m4_a <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s3_m4_a_print, results = 'asis'}
+s3_m4_a %>% report_results_2(.)
+```
+
+<br><br>
+<br><br>

### 2019-08-18 data_analysis/data_analysis_expanded.Rmd
Data analysis update (m0 done)

--- 2019-08-13 data_analysis/data_analysis_expanded.Rmd
@@
 library(rre)
-library(dplyr);
-#setwd("/Users/alex/Dropbox/classGrad/IS_Wil/rre_sims/data_analysis/")
+library(dplyr)
 
@@
 
-#test <- rre::unregularized_mle(list_of_samples[[1]], 
-#                               multiplier = 2,
-#                               c_seq_len = 2)
-
-report_results_2 <- function(method_output) {
+report_results_2 <- function(method_output, sample, report_eta = F) {
   best <- method_output[["best"]]
   c_hat <- best$ccc
-  m <- c_hat/(best$cc_max)
-  cat(paste0("C_hat: ", c_hat, "\n\n", round(m,2), " times greater than observed richness \n\nSelected lambda: ", best$selected_lambda, "\n\n"))
+  m <- c_hat/(rre::get_cc_max(sample))
+  out_text <- paste0("C_hat: ", c_hat, "\n\n", 
+             round(m,2), " times greater than observed richness \n\nSelected lambda: ", 
+             best$selected_lambda, "\n\n")
+  if (report_eta) {
+    out_text <- paste0(out_text, "alpha = ", best$alpha,
+                       ", delta = ", best$delta)
+  }
+  cat(out_text)
 }
@@
 
+### Sample 1
+
 ```{r s1_m0_a}
@@
 ```{r s1_m0_a_print, results = 'asis'}
-s1_m0_a %>% report_results_2(.)
+report_results_2(s1_m0_a, list_of_samples[[1]], report_eta = T)
 ```
@@
 <br>
+
+```{r s1_m0_b}
+# Use a very dense C grid, see if previous result is stable:
+s1_m0_b <- rre::unregularized_mle(list_of_samples[[1]], 
+                                  multiplier = 200,
+                                  c_seq_len = 1000,
+                                  starts = data.frame(alpha = 0.0008818,
+                                           delta = 0.001802))
+```
+```{r s1_m0_b_print, results = 'asis'}
+report_results_2(s1_m0_b, list_of_samples[[1]], report_eta = T)
+```
+<br>
+<br>
+
+### Sample 2
 
@@
 ```{r s2_m0_a_print, results = 'asis'}
-s2_m0_a %>% report_results_2(.)
+report_results_2(s2_m0_a,  list_of_samples[[2]], report_eta = T)
 ```
@@
 <br>
+
+```{r s2_m0_b}
+# Use a very dense C grid, see if previous result is stable:
+s2_m0_b <- rre::unregularized_mle(list_of_samples[[2]], 
+                                  multiplier = 200,
+                                  c_seq_len = 1000,
+                                  starts = data.frame(alpha = 0.001846,
+                                                      delta = 0.002530))
+```
+```{r s2_m0_b_print, results = 'asis'}
+report_results_2(s2_m0_b,  list_of_samples[[2]], report_eta = T)
+```
+<br>
+<br>
+
+### Sample 3
 
@@
 ```{r s3_m0_a_print, results = 'asis'}
-s3_m0_a %>% report_results_2(.)
+report_results_2(s3_m0_a, list_of_samples[[3]], report_eta = T)
+```
+<br>
+<br>
+
+```{r s3_m0_b}
+# Use a very dense C grid, see if previous result is stable:
+s3_m0_b <- rre::unregularized_mle(list_of_samples[[3]],
+                                  multiplier = 200,
+                                  c_seq_len = 1000,
+                                  starts = data.frame(alpha = 0.001606,
+                                                      delta = 0.001396))
+```
+```{r s3_m0_b_print, results = 'asis'}
+report_results_2(s3_m0_b, list_of_samples[[3]], report_eta = T)
 ```
@@
 ```{r s1_m1_a_print, results = 'asis'}
-s1_m1_a %>% report_results_2(.)
+s1_m1_a %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+```
+<br><br>
+
+```{r s1_m1_b}
+# more starts, lambda refinement, smaller multiplier:
+set.seed(12439)
+s1_m1_b <- rre::minimum_subset_distance(list_of_samples[[1]],
+                                        partitions = 5,
+                                        lambda_vec = seq(400,800, by = 50),
+                                        multiplier = 20,
+                                        c_seq_len = 50, # mean to to do 50 here last time, and did not.
+                                        starts = big_starts)
+```
+```{r s1_m1_b_print, results = 'asis'}
+s1_m1_b %>% report_results_2(.,  list_of_samples[[1]], report_eta = T)
+```
+<br><br>
+
+```{r s1_m1_c}
+# lambda refinement, see if the result is stable, full partitions:
+set.seed(12439)
+s1_m1_c <- rre::minimum_subset_distance(list_of_samples[[1]],
+                                        partitions = 10,
+                                        lambda_vec = seq(650, 750, by = 5),
+                                        multiplier = 10,
+                                        c_seq_len = 100,
+                                        starts = data.frame(alpha = 0.3463,
+                                                            delta = 0.003673))
+```
+```{r s1_m1_c_print, results = 'asis'}
+s1_m1_c %>% report_results_2(.,  list_of_samples[[1]], report_eta = T)
+```
+<br><br>
+
+```{r s1_m1_d}
+
+set.seed(2339)
+s1_m1_d <- rre::minimum_subset_distance(list_of_samples[[1]],
+                                        partitions = 10,
+                                        lambda_vec = 700:740,
+                                        multiplier = 2,
+                                        c_seq_len = 11,
+                                        starts = data.frame(alpha = 0.28863,
+                                                            delta = 0.0032607))
+```
+```{r s1_m1_d_print, results = 'asis'}
+s1_m1_d %>% report_results_2(.,  list_of_samples[[1]], report_eta = T)
 ```
@@
 ```{r s2_m1_a_print, results = 'asis'}
-s2_m1_a %>% report_results_2(.)
+s2_m1_a %>% report_results_2(., list_of_samples[[2]])
+```
+<br><br>
+
+```{r s2_m1_b}
+# more starts, lambda refinement, smaller multiplier:
+set.seed(239)
+s2_m1_b <- rre::minimum_subset_distance(list_of_samples[[2]],
+                                        partitions = 5,
+                                        lambda_vec = seq(400,800,by = 50),
+                                        multiplier = 20,
+                                        c_seq_len = 50,
+                                        starts = big_starts)
+```
+```{r s2_m1_b_print, results = 'asis'}
+s2_m1_b %>% report_results_2(., list_of_samples[[2]], report_eta = T)
+```
+<br><br>
+
+```{r s2_m1_c}
+set.seed(239234)
+s2_m1_c <- rre::minimum_subset_distance(list_of_samples[[2]],
+                                        partitions = 10,
+                                        lambda_vec = seq(450,550,by=5),
+                                        multiplier = 2,
+                                        c_seq_len = 11,
+                                        starts = data.frame(alpha = 0.56337,
+                                                            delta = 0.0067572))
+```
+```{r s2_m1_c_print, results = 'asis'}
+s2_m1_c %>% report_results_2(., list_of_samples[[2]], report_eta = T)
 ```
@@
 ```{r s3_m1_a_print, results = 'asis'}
-s3_m1_a %>% report_results_2(.)
+s3_m1_a %>% report_results_2(., list_of_samples[[3]])
 ```
@@
 
+```{r s3_m1_b}
+# more starts, lambda refinement, smaller multiplier:
+set.seed(45698)
+s3_m1_b <- rre::minimum_subset_distance(list_of_samples[[3]],
+                                        partitions = 5,
+                                        lambda_vec = seq(200,600,by=50),
+                                        multiplier = 20,
+                                        c_seq_len = 50,
+                                        starts = big_starts)
+```
+```{r s3_m1_b_print, results = 'asis'}
+s3_m1_b %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+```{r s3_m1_c}
+# refine lambda, full partitions, see if the previous result is stable:
+set.seed(45698)
+s3_m1_c <- rre::minimum_subset_distance(list_of_samples[[3]],
+                                        partitions = 10,
+                                        lambda_vec = seq(500, 600 ,by=5),
+                                        multiplier = 10,
+                                        c_seq_len = 100,
+                                        starts = data.frame(alpha = 0.4392,
+                                                            delta = 0.003637))
+```
+```{r s3_m1_c_print, results = 'asis'}
+s3_m1_c %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+```{r s3_m1_c}
+set.seed(452348)
+s3_m1_c <- rre::minimum_subset_distance(list_of_samples[[3]],
+                                        partitions = 10,
+                                        lambda_vec = seq(500, 600 ,by=5),
+                                        multiplier = 2,
+                                        c_seq_len = 11,
+                                        starts = data.frame(alpha = 0.5143,
+                                                            delta = 0.003997))
+```
+```{r s3_m1_c_print, results = 'asis'}
+s3_m1_c %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+<br><br>
+
+## Method 2
+
+### Sample 1
+
+```{r s1_m2_a}
+# big grids and few partitions to start:
+set.seed(12039)
+s1_m2_a <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s1_m2_a_print, results = 'asis'}
+s1_m2_a %>% report_results_2(., list_of_samples[[1]])
+```
+<br><br>
+
+```{r s1_m2_b}
+# narrow lambda, more starts:
+set.seed(12039)
+s1_m2_b <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,300,by=20),
+                              multiplier = 200,
+                              c_seq_len = 20,
+                              starts = big_starts)
+```
+```{r s1_m2_b_print, results = 'asis'}
+s1_m2_b %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+```
+<br><br>
+
+### Sample 2
+
+```{r s2_m2_a}
+# big grids and few partitions to start:
+set.seed(120239)
+s2_m2_a <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s2_m2_a_print, results = 'asis'}
+s2_m2_a %>% report_results_2(., list_of_samples[[2]])
+```
+<br><br>
+
+```{r s2_m2_b}
+# narrow lambda, more starts:
+set.seed(12039)
+s2_m2_b <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,300,by=20),
+                              multiplier = 200,
+                              c_seq_len = 20,
+                              starts = big_starts)
+```
+```{r s2_m2_b_print, results = 'asis'}
+s2_m2_b %>% report_results_2(., list_of_samples[[2]], report_eta = T)
+```
+<br><br>
+
+### Sample 3
+
+```{r s3_m2_a}
+# big grids and few partitions to start:
+set.seed(223439)
+s3_m2_a <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s3_m2_a_print, results = 'asis'}
+s3_m2_a %>% report_results_2(., list_of_samples[[3]])
+```
+<br><br>
+
+```{r s3_m2_b}
+# narrow lambda, more starts:
+set.seed(23439)
+s3_m2_b <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,300,by=20),
+                              multiplier = 200,
+                              c_seq_len = 20,
+                              starts = big_starts)
+```
+```{r s3_m2_b_print, results = 'asis'}
+s3_m2_b %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
 <br><br>
@@
 ```{r s1_m3_a_print, results = 'asis'}
-s1_m3_a %>% report_results_2(.)
+s1_m3_a %>% report_results_2(., list_of_samples[[1]])
 ```
@@
 ```{r s1_m3_b_print, results = 'asis'}
-s1_m3_b %>% report_results_2(.)
+s1_m3_b %>% report_results_2(., list_of_samples[[1]])
 ```
@@
 ```{r s1_m3_c_print, results = 'asis'}
-s1_m3_c %>% report_results_2(.)
+s1_m3_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```
@@
 ```{r s2_m3_a_print, results = 'asis'}
-s2_m3_a %>% report_results_2(.)
+s2_m3_a %>% report_results_2(., list_of_samples[[2]])
 ```
@@
 ```{r s2_m3_b_print, results = 'asis'}
-s2_m3_b %>% report_results_2(.)
+s2_m3_b %>% report_results_2(., list_of_samples[[2]])
 ```
@@
 ```{r s2_m3_c_print, results = 'asis'}
-s2_m3_c %>% report_results_2(.)
+s2_m3_c %>% report_results_2(., list_of_samples[[2]])
 ```
@@
 ```{r s2_m3_d_print, results = 'asis'}
-s2_m3_d %>% report_results_2(.)
+s2_m3_d %>% report_results_2(., list_of_samples[[2]], report_eta =T)
 ```
@@
 ```{r s3_m3_a_print, results = 'asis'}
-s3_m3_a %>% report_results_2(.)
+s3_m3_a %>% report_results_2(., list_of_samples[[3]])
 ```
@@
 ```{r s3_m3_b_print, results = 'asis'}
-s3_m3_b %>% report_results_2(.)
+s3_m3_b %>% report_results_2(., list_of_samples[[3]])
 ```
@@
 ```{r s3_m3_c_print, results = 'asis'}
-s3_m3_c %>% report_results_2(.)
+s3_m3_c %>% report_results_2(., list_of_samples[[3]])
 ```
@@
 ```{r s3_m3_d_print, results = 'asis'}
-s3_m3_d %>% report_results_2(.)
+s3_m3_d %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+## Method 4
+
+### Sample 1
+
+```{r s1_m4_a}
+# big grids and few partitions to start:
+set.seed(87124)
+s1_m4_a <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s1_m4_a_print, results = 'asis'}
+s1_m4_a %>% report_results_2(., list_of_samples[[1]])
 ```
@@
 <br><br>
+
+```{r s1_m4_b}
+set.seed(24)
+s1_m4_b <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,300,by=20),
+                              multiplier = 200,
+                              c_seq_len = 50,
+                              starts = big_starts)
+```
+```{r s1_m4_b_print, results = 'asis'}
+s1_m4_b %>% report_results_2(., list_of_samples[[1]])
+```
+
+<br><br>
+
+### Sample 2
+
+```{r s2_m4_a}
+# big grids and few partitions to start:
+set.seed(23456)
+s2_m4_a <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s2_m4_a_print, results = 'asis'}
+s2_m4_a %>% report_results_2(., list_of_samples[[2]])
+```
+
+<br><br>
+
+```{r s2_m4_b}
+set.seed(2345)
+s2_m4_b <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,200,by=20),
+                              multiplier = 200,
+                              c_seq_len = 50,
+                              starts = big_starts)
+```
+```{r s2_m4_b_print, results = 'asis'}
+s2_m4_b%>% report_results_2(., list_of_samples[[2]])
+```
+
+<br><br>
+
+```{r s3_m4_a}
+# big grids and few partitions to start:
+set.seed(7654)
+s3_m4_a <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,1000,by=200),
+                              multiplier = 200,
+                              c_seq_len = 20)
+```
+```{r s3_m4_a_print, results = 'asis'}
+s3_m4_a %>% report_results_2(., list_of_samples[[3]])
+```
+
+<br><br>
+
+```{r s3_m4_b}
+# big grids and few partitions to start:
+set.seed(7654)
+s3_m4_b <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,200,by=20),
+                              multiplier = 200,
+                              c_seq_len = 50,
+                              starts = big_starts)
+```
+```{r s3_m4_b_print, results = 'asis'}
+s3_m4_b %>% report_results_2(., list_of_samples[[3]])
+```
+
+<br><br>
+<br><br>

### 2019-08-19 data_analysis/data_analysis_expanded.Rmd
Rdata output added (results not complete)

@@
 
-```{r s3_m1_c}
+```{r s3_m1_d}
 set.seed(452348)
-s3_m1_c <- rre::minimum_subset_distance(list_of_samples[[3]],
+s3_m1_d <- rre::minimum_subset_distance(list_of_samples[[3]],
                                         partitions = 10,
@@
 ```
-```{r s3_m1_c_print, results = 'asis'}
-s3_m1_c %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```{r s3_m1_d_print, results = 'asis'}
+s3_m1_d %>% report_results_2(., list_of_samples[[3]], report_eta = T)
 ```
@@
 ```{r s1_m3_d}
-# Feed in more alpha and delta, narrow lambdagrid more.
 #s1_m3_d <- rre::gof_criterion(list_of_samples[[1]], 
-#                              lambda_vec = seq(505,525,by=2),
-#                              multiplier = 100,
-#                              c_seq_len = 50,
-#                              starts = data.frame(alpha = c(0.1, 0.01, 0.01, 0.001), delta = c(0.1, 0.01, 0.0001, 0.001)))
+#                              lambda_vec = seq(450,550,by=10),
+#                              multiplier = 200,
+#                              c_seq_len = 200,
+#                              starts = data.frame(alpha = 0.001738,
+#                                                  delta = 0.001744))
 ```
 ```{r s1_m3_d_print, results = 'asis'}
-#s1_m3_d %>% report_results_2(.)
+#s1_m3_d %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```
-<br>
-<br>
-
+<br><br>
 <br><br>
@@
 ```
-<br>
-<br>
+<br><br>
 
+```{r s2_m3_e}
+# Selected lambda was on the edge of previous grid, lets try again:
+#s2_m3_e <- rre::gof_criterion(list_of_samples[[2]], 
+#                              lambda_vec = seq(150,300,by=5),
+#                              multiplier = 100,
+#                              c_seq_len = 499,
+#                              starts = data.frame(alpha = 0.004656,
+#                                                  delta = 0.0025668))
+```
+```{r s2_m3_e_print, results = 'asis'}
+#s2_m3_e %>% report_results_2(., list_of_samples[[2]], report_eta =T)
+```
+<br><br>
 <br><br>
@@
 <br><br>
+
+```{r s3_m3_e}
+# Selected lambda was on the edge of previous grid, lets try again:
+#s3_m3_e <- rre::gof_criterion(list_of_samples[[3]], 
+#                              lambda_vec = seq(150, 250, by=5),
+#                              multiplier = 100,
+#                              c_seq_len = 499,
+#                              starts = data.frame(alpha = 0.0014137,
+ #                                                 delta = 0.001367))
+```
+```{r s3_m3_e_print, results = 'asis'}
+#s3_m3_e %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br><br><br>
 
@@
 <br><br>
+
+## Output data
+
+For each method and sample we save the most refined or best estimate.
+```{r output_data}
+sample_1 <- list(method_0 = s1_m0_b$best,
+                 method_1 = s1_m1_d$best,
+                 method_2 = s1_m2_b$best,
+                 method_3 = s1_m3_c$best,
+                 method_4 = s1_m4_b$best)
+
+sample_2 <- list(method_0 = s2_m0_b$best,
+                 method_1 = s2_m1_c$best,
+                 method_2 = s2_m2_b$best,
+                 method_3 = s2_m3_d$best,
+                 method_4 = s2_m4_b$best)
+
+sample_3 <- list(method_0 = s3_m0_b$best,
+                 method_1 = s3_m1_d$best,
+                 method_2 = s3_m2_b$best,
+                 method_3 = s3_m3_d$best,
+                 method_4 = s3_m4_b$best)
+
+sample_outputs <- list(sample_1 = sample_1, 
+                       sample_2 = sample_2, 
+                       sample_3 = sample_3)
+
+save(sample_outputs, file = "data_analysis_expanded_results.RData")
+
+```

### 2019-08-20 data_analysis/figures_tables_expanded.R
Figures and tables script added

@@ -1,0 +1,72 @@
+# Figure and table creation for the expanded data analysis.
+
+load("data_analysis_expanded_results.RData")
+
+sample_outputs$sample_1 %>% lapply(., (function(x) c(x$ccc, x$selected_lambda, x$alpha, x$delta))) %>%
+  do.call(rbind, .) %>%
+  'colnames<-'(., c("ccc", "selected_lambda", "alpha", "delta")) %>%
+  as.data.frame(.) %>%
+  tibble::add_column(method = 0:4)
+
+one_sample <- function(sample) {
+  lapply(sample, (function(x) c(x$ccc, x$selected_lambda, x$alpha, x$delta))) %>%
+    do.call(rbind, .) %>%
+    'colnames<-'(., c("ccc", "selected_lambda", "alpha", "delta")) %>%
+    as.data.frame(.) %>%
+    tibble::add_column(method = 0:4)
+}
+
+all_samples <- function(sample_outputs) {
+  lapply(sample_outputs, one_sample) %>% 
+    do.call(rbind, .) %>%
+    tibble::add_column(year = rep(2009:2011, each = 5)) %>%
+    'rownames<-'(., NULL) %>%
+    dplyr::select(year, method, ccc, selected_lambda, alpha, delta)
+}
+
+
+plot_fn <- function(output_df, 
+                    output_name = NA,
+                    output_height = 4,
+                    output_width = 7) {
+  # match method names to simulations:
+  output_df %<>% dplyr::mutate(method_display = dplyr::case_when(
+    method == 0 ~ "[0] MLE", # method 00
+    method == 1 ~ "[1] Min Var", # method 01
+    method == 2 ~ "[2] CV likelihood", # method 02
+    method == 3 ~ "[3] GOF", # method 03
+    method == 4 ~ "[4] CV GOF", # method 04 
+    TRUE ~ "error")) %>%
+    dplyr::mutate(method_display = factor(method_display))
+  
+  if (any(output_df$method_display == "error")) stop("Coding not recognized")
+  
+  pd <- position_dodge(0.1)
+  gg <- ggplot(output_df, aes(x = year, y = ccc, 
+                              group = method_display, 
+                              color = method_display)) +
+    geom_point(position = pd) + 
+    geom_line(position = pd) +
+    labs(y = latex2exp::TeX("$\\hat{C}$"), x = "Year") + 
+    scale_color_viridis_d("Method", option = "viridis", begin = 0.15, end = 0.9) +    
[…]

### 2019-08-20 data_analysis/data_analysis_expanded.Rmd
another data analysis update

@@
 ```{r s1_m3_d}
-#s1_m3_d <- rre::gof_criterion(list_of_samples[[1]], 
-#                              lambda_vec = seq(450,550,by=10),
-#                              multiplier = 200,
-#                              c_seq_len = 200,
-#                              starts = data.frame(alpha = 0.001738,
-#                                                  delta = 0.001744))
+s1_m3_d <- rre::gof_criterion(list_of_samples[[1]],
+                             lambda_vec = seq(500,650,by=10),
+                             multiplier = 200,
+                             c_seq_len = 200,
+                             starts = data.frame(alpha = 0.001738,
+                                                 delta = 0.001744))
 ```
 ```{r s1_m3_d_print, results = 'asis'}
-#s1_m3_d %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+s1_m3_d %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```
@@
 # Selected lambda was on the edge of previous grid, lets try again:
-#s2_m3_e <- rre::gof_criterion(list_of_samples[[2]], 
-#                              lambda_vec = seq(150,300,by=5),
-#                              multiplier = 100,
-#                              c_seq_len = 499,
-#                              starts = data.frame(alpha = 0.004656,
-#                                                  delta = 0.0025668))
+s2_m3_e <- rre::gof_criterion(list_of_samples[[2]],
+                             lambda_vec = seq(150,300,by=10),
+                             multiplier = 100,
+                             c_seq_len = 100,
+                             starts = data.frame(alpha = 0.004656,
+                                                 delta = 0.0025668))
 ```
 ```{r s2_m3_e_print, results = 'asis'}
-#s2_m3_e %>% report_results_2(., list_of_samples[[2]], report_eta =T)
+s2_m3_e %>% report_results_2(., list_of_samples[[2]], report_eta =T)
 ```
@@
 # Selected lambda was on the edge of previous grid, lets try again:
-#s3_m3_e <- rre::gof_criterion(list_of_samples[[3]], 
-#                              lambda_vec = seq(150, 250, by=5),
-#                              multiplier = 100,
-#                              c_seq_len = 499,
-#                              starts = data.frame(alpha = 0.0014137,
- #                                                 delta = 0.001367))
+s3_m3_e <- rre::gof_criterion(list_of_samples[[3]],
+                             lambda_vec = seq(150, 250, by=10),
+                             multiplier = 100,
+                             c_seq_len = 100,
+                             starts = data.frame(alpha = 0.0014137,
+                                                delta = 0.001367))
 ```
 ```{r s3_m3_e_print, results = 'asis'}
-#s3_m3_e %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+s3_m3_e %>% report_results_2(., list_of_samples[[3]], report_eta = T)
 ```

### 2019-08-20 data_analysis/data_analysis_expanded.Rmd
routine update

@@
 ```{r s1_m4_b_print, results = 'asis'}
-s1_m4_b %>% report_results_2(., list_of_samples[[1]])
+s1_m4_b %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```
@@
 ```{r s2_m4_b_print, results = 'asis'}
-s2_m4_b%>% report_results_2(., list_of_samples[[2]])
+s2_m4_b%>% report_results_2(., list_of_samples[[2]], report_eta = T)
 ```
@@
 
+### Sample 3
 ```{r s3_m4_a}
@@
 ```{r s3_m4_b_print, results = 'asis'}
-s3_m4_b %>% report_results_2(., list_of_samples[[3]])
+s3_m4_b %>% report_results_2(., list_of_samples[[3]], report_eta = T)
 ```

### 2019-08-22 data_analysis/data_analysis_expanded.Rmd
another update

@@
 
+```{r s1_m2_c}
+set.seed(4439)
+s1_m2_c <- rre::cv_replicates(list_of_samples[[1]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(150,250,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.001879,
+                                                  delta = 0.002208))
+```
+```{r s1_m2_c_print, results = 'asis'}
+s1_m2_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+```
+<br><br>
+
 ### Sample 2
@@
 s2_m2_b %>% report_results_2(., list_of_samples[[2]], report_eta = T)
+```
+<br><br>
+
+```{r s2_m2_c}
+# narrow lambda, more starts:
+set.seed(12044)
+s2_m2_c <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,100,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.002191,
+                                                  delta = 0.002789))
+```
+```{r s2_m2_c_print, results = 'asis'}
+s2_m2_c %>% report_results_2(., list_of_samples[[2]], report_eta = T)
 ```
@@
 <br><br>
+
+```{r s3_m2_c}
+# narrow lambda, more starts:
+set.seed(23444)
+s3_m2_c <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "neg_unreg_like",
+                              partitions = 5,
+                              lambda_vec = seq(0,100,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.005138,
+                                                  delta = 0.002135))
+```
+```{r s3_m2_c_print, results = 'asis'}
+s3_m2_c %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+_Note_ The 'c' round of method 2 took over a day to run for these 3 samples.
+
 <br><br>
@@
 
+```{r s1_m4_c}
+set.seed(2423)
+s1_m4_c <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(150,250,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.005779,
+                                                  delta = 0.002530))
+```
+```{r s1_m4_c_print, results = 'asis'}
+s1_m4_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+```
+
+<br><br>
+
 ### Sample 2
@@
 s2_m4_b%>% report_results_2(., list_of_samples[[2]], report_eta = T)
+```
+
+<br><br>
+
+```{r s2_m4_c}
+set.seed(23451)
+s2_m4_c <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,100,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.0016934,
+                                                  delta = 0.0027209))
+```
+```{r s2_m4_c_print, results = 'asis'}
+s2_m4_c%>% report_results_2(., list_of_samples[[2]], report_eta = T)
 ```
@@
 <br><br>
+
+```{r s3_m4_c}
+# big grids and few partitions to start:
+set.seed(7654)
+s3_m4_c <- rre::cv_replicates(list_of_samples[[3]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(0,120,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.002450,
+                                                  delta = 0.001773))
+```
+```{r s3_m4_c_print, results = 'asis'}
+s3_m4_c %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+
+<br><br>
 <br><br>

### 2019-08-24 data_analysis/data_analysis_expanded.Rmd
third pass for all methods

@@
                               partitions = 5,
-                              lambda_vec = seq(150,250,by=5),
+                              lambda_vec = seq(50,150,by=5),
                               multiplier = 200,
@@
                  method_1 = s1_m1_d$best,
-                 method_2 = s1_m2_b$best,
+                 method_2 = s1_m2_c$best,
                  method_3 = s1_m3_c$best,
-                 method_4 = s1_m4_b$best)
+                 method_4 = s1_m4_c$best)
 
@@
                  method_1 = s2_m1_c$best,
-                 method_2 = s2_m2_b$best,
+                 method_2 = s2_m2_c$best,
                  method_3 = s2_m3_d$best,
-                 method_4 = s2_m4_b$best)
+                 method_4 = s2_m4_c$best)
 
@@
                  method_1 = s3_m1_d$best,
-                 method_2 = s3_m2_b$best,
+                 method_2 = s3_m2_c$best,
                  method_3 = s3_m3_d$best,
-                 method_4 = s3_m4_b$best)
+                 method_4 = s3_m4_c$best)
 

### 2019-08-25 data_analysis/data_analysis_expanded.Rmd
bug fix

@@
 ```{r s1_m4_c}
-set.seed(2423)
-s1_m4_c <- rre::cv_replicates(list_of_samples[[2]],
-                              eval_function = "gof_chi_sq",
-                              partitions = 5,
-                              lambda_vec = seq(50,150,by=5),
-                              multiplier = 200,
-                              c_seq_len = 200,
-                              starts = data.frame(alpha = 0.005779,
-                                                  delta = 0.002530))
+#set.seed(2423)
+#s1_m4_c <- rre::cv_replicates(list_of_samples[[2]],
+#                              eval_function = "gof_chi_sq",
+#                              partitions = 5,
+#                              lambda_vec = seq(50,150,by=5),
+#                              multiplier = 200,
+#                              c_seq_len = 200,
+#                              starts = data.frame(alpha = 0.005779,
+#                                                  delta = 0.002530))
 ```
 ```{r s1_m4_c_print, results = 'asis'}
-s1_m4_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+#s1_m4_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```
@@
                  method_2 = s1_m2_c$best,
-                 method_3 = s1_m3_c$best,
-                 method_4 = s1_m4_c$best)
+                 method_3 = s1_m3_d$best,
+                 method_4 = s1_m4_b$best)
 
@@
                  method_2 = s2_m2_c$best,
-                 method_3 = s2_m3_d$best,
+                 method_3 = s2_m3_e$best,
                  method_4 = s2_m4_c$best)
@@
                  method_2 = s3_m2_c$best,
-                 method_3 = s3_m3_d$best,
+                 method_3 = s3_m3_e$best,
                  method_4 = s3_m4_c$best)

### 2019-08-26 rre_manuscript/data_analysis_extended.tex
Manuscript data analysis first draft

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib, bbm}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{amsthm}
\usepackage{lipsum}

\usepackage{xparse}
\usepackage{mathtools}
\usepackage[table,dvipsnames]{xcolor}

\usepackage{ifthen}
\usepackage{pstricks}
\usepackage[normalem]{ulem}

\usepackage{booktabs}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\kmax}{k_{\text{max}}}
\newcommand{\lambdagrid}{\lambda^{\text{grid}}}

\usepackage{multirow}

\newtheoremstyle{break}
  {\topsep}{\topsep}%
  {\itshape}{}%
  {\bfseries}{}%
  {\newline}{}%
\theoremstyle{break}
\newtheorem{theorem}{Theorem}

\newtheoremstyle{break}
  {\topsep}{\topsep}%
  {\itshape}{}%
  {\bfseries}{}%
  {\newline}{}%
\theoremstyle{break}
\newtheorem{prop}{Proposition}

\newboolean{DEBUG}
\setboolean{DEBUG}{true}
\newrgbcolor{amycolor}{.7 .1 .1}
\ifthenelse {\boolean{DEBUG}}
{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
{\newcommand{\amy}[1]{}}

\RequirePackage[OT1]{fontenc}
\RequirePackage{amsthm,amsmath, bbm, xspace, graphics, graphicx}
%\RequirePackage{breqn}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}

\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Data Analysis}
\label{sec:data_analysis}

\subsection{Estimating microbial richness in Lake Champlain}

To illustrate the performance of our methods on ecological data, we estimate strain-level microbial diversity in Lake Champlain, a large eutrophic lake in Canada.  We analyze data from \citet{tromas_2017}, considering samples from the littoral zone in the summer season of the same year as replicates. This gives us 8 replicates from 2009, 6 replicates from 2010 and 6 replicates from 2011.

The richness estimates from each method are displayed in Figure \ref{fig:data_analysis} and Table \ref{tab:data_analysis}.  These estimates are high as 160 times the observed richness, suggesting that the negative binomial model may be a poor fit to this data set.  As a result, applying our methods to real data required some manual tuning of $\lambdagrid$ and the search grid for $C$.

Despite the estimates being significantly more inflated than in our simulations, many of the trends observed in simulations hold true here.  Method 0 produces the highest $\widehat{C}$ for 2009 and 2010 and it was second highest to Method 2 in 2011.  Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness $c$.  These are exactly the tendencies these methods showed when we generate simulations from a correctly specified model (see Figure \ref{fig:tuning_sim_1}).

We model species richness because the observed richness is an underestimate of the truth.  The original motivation for penalization is that the unpenalized MLE estimate can be much larger than the truth.  In data analysis we don't know the truth, but we would expect an ideal method to be situated between these two extremes.  Only the methods making use of a goodness of fit metric (3 and 4) are between Method 0 and the observed richness in all samples.  As a result we conclude Methods 3 and 4 show the greatest promise on real data containing replicates.

\noindent\large{\textbf{Note from Alex:  I've given two versions of the table, the first is much bigger and probably more suited to being a supplement.  The second is more compact and could be crammed into a corner of the paper if needed.  Happy to make revisions to tables and figures if that's helpful. }}

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis}}
\centering
\begin{tabular}{llcccc}
  \toprule
Year & Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
2009 & {[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 \\ 
  2010 & {[0]} Unpenalized MLE & 47,631 & \textemdash & 0.00185 & 0.00253 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 572 & 500 & 0.62668 & 0.00724 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 24,799 & 0 & 0.00379 & 0.00270 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 13,156 & 220 & 0.00685 & 0.00258 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 47,098 & 0 & 0.00208 & 0.00277 \\ 
  2011 & {[0]} Unpenalized MLE & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  \multirow{5}{*} & {[1]} Minimum Variance & 718 & 500 & 0.40112 & 0.00355 \\ 
  \multirow{5}{*} & {[2]} CV Likelihood & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  \multirow{5}{*} & {[3]} Goodness of fit & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  \multirow{5}{*} & {[4]} CV G.O.F. & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
\label{tab:data_analysis_compact}}
\centering
\tiny
\begin{tabular}{lllllllllllll}
  \toprule
      & \multicolumn{4}{c}{2009} & \multicolumn{4}{c}{2010} & \multicolumn{4}{c}{2011} \\ \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(lr){10-13}
Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\ 
  \midrule
{[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 & 47,631 & \textemdash & 0.00185 & 0.00253 & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
  {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\ 
  {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\ 
  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 220 & 0.00685 & 0.00258 & 40,040 & 230 & 0.00231 & 0.00137 \\ 
  {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) based on our proposed methods.
%$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.  The number of replicates in each sample is given by $r$.
%
% A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.
\label{fig:data_analysis}}
\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/data_analysis_expanded.pdf}}
\end{figure}

\bibliographystyle{biorefs}
\bibliography{refs,amys-papers-library}

\end{document}

### 2019-08-26 data_analysis/figures_tables_expanded.R

@@ -2,12 +2,6 @@
 
 load("data_analysis_expanded_results.RData")
 
-sample_outputs$sample_1 %>% lapply(., (function(x) c(x$ccc, x$selected_lambda, x$alpha, x$delta))) %>%
-  do.call(rbind, .) %>%
-  'colnames<-'(., c("ccc", "selected_lambda", "alpha", "delta")) %>%
-  as.data.frame(.) %>%
-  tibble::add_column(method = 0:4)
-
 one_sample <- function(sample) {
   lapply(sample, (function(x) c(x$ccc, x$selected_lambda, x$alpha, x$delta))) %>%
     do.call(rbind, .) %>%
@@ -41,7 +35,7 @@ plot_fn <- function(output_df,
   
   if (any(output_df$method_display == "error")) stop("Coding not recognized")
   
-  pd <- position_dodge(0.1)
+  pd <- position_dodge(0)
   gg <- ggplot(output_df, aes(x = year, y = ccc, 
                               group = method_display, 
                               color = method_display)) +
@@ -67,6 +61,88 @@ plot_fn <- function(output_df,
   }
 }
 
+sample_outputs %>% all_samples %>% str
 
-all_samples(sample_outputs) %>% plot_fn(.)
+# compact table output:
+table_to_console_1 <- function(outs) {
+  outs %<>% all_samples  
+  outs %<>% dplyr::mutate(method = dplyr::case_when(
+    method == 0 ~ "{[0]} Unpenalized MLE", # method 00
+    method == 1 ~ "{[1]} Minimum Variance", # method 01
+    method == 2 ~ "{[2]} CV Likelihood", # method 02
+    method == 3 ~ "{[3]} Goodness of fit", # method 03
+    method == 4 ~ "{[4]} CV G.O.F.", # method 04 
+    TRUE ~ "error")) %>%
+    dplyr::mutate(method = factor(method))
+  
+  if (any(outs$method_display == "error")) stop("Coding not recognized")
+  
+  outs %<>% dplyr::mutate(ccc = formatC(ccc, format="d", big.mark=","),
+                         selected_lambda = ifelse(is.na(selected_lambda), 
+                                                  "\\textemdash", 
+                                                  as.character(selected_lambda)),
+                         alpha = formatC(alpha, format = "f", digits = 5),
+                         delta = formatC(delta, format = "f", digits = 5))
[…]

### 2019-08-26 data_analysis/data_analysis_expanded.Rmd
Touch ups on method 3

@@
 <br><br>
+
+```{r s1_m3_e}
+s1_m3_e <- rre::gof_criterion(list_of_samples[[1]],
+                             lambda_vec = seq(520,580,by=5),
+                             multiplier = 200,
+                             c_seq_len = 200,
+                             starts = data.frame(alpha = 0.001738,
+                                                 delta = 0.001744))
+```
+```{r s1_m3_e_print, results = 'asis'}
+s1_m3_e %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+```
+<br><br>
 <br><br>
@@
 <br><br>
+
+```{r s2_m3_f}
+# Selected lambda was on the edge of previous grid, lets try again:
+s2_m3_f <- rre::gof_criterion(list_of_samples[[2]],
+                             lambda_vec = seq(170,270,by=5),
+                             multiplier = 100,
+                             c_seq_len = 100,
+                             starts = data.frame(alpha = 0.004656,
+                                                 delta = 0.0025668))
+```
+```{r s2_m3_f_print, results = 'asis'}
+s2_m3_f %>% report_results_2(., list_of_samples[[2]], report_eta =T)
+```
+<br><br>
 <br><br>
@@
 ```{r s3_m3_e}
-# Selected lambda was on the edge of previous grid, lets try again:
 s3_m3_e <- rre::gof_criterion(list_of_samples[[3]],
@@
 ```
-<br><br><br><br>
+<br><br>
+
+```{r s3_m3_f}
+# Dense lambda grid:
+s3_m3_f <- rre::gof_criterion(list_of_samples[[3]],
+                             lambda_vec = seq(180, 280, by=5),
+                             multiplier = 100,
+                             c_seq_len = 100,
+                             starts = data.frame(alpha = 0.0014137,
+                                                delta = 0.001367))
+```
+```{r s3_m3_f_print, results = 'asis'}
+s3_m3_f %>% report_results_2(., list_of_samples[[3]], report_eta = T)
+```
+<br><br>
+
+<br><br>
 
@@
 ```{r s1_m4_c}
-#set.seed(2423)
-#s1_m4_c <- rre::cv_replicates(list_of_samples[[2]],
-#                              eval_function = "gof_chi_sq",
-#                              partitions = 5,
-#                              lambda_vec = seq(50,150,by=5),
-#                              multiplier = 200,
-#                              c_seq_len = 200,
-#                              starts = data.frame(alpha = 0.005779,
-#                                                  delta = 0.002530))
+set.seed(2423)
+s1_m4_c <- rre::cv_replicates(list_of_samples[[2]],
+                              eval_function = "gof_chi_sq",
+                              partitions = 5,
+                              lambda_vec = seq(50,150,by=5),
+                              multiplier = 200,
+                              c_seq_len = 200,
+                              starts = data.frame(alpha = 0.005779,
+                                                  delta = 0.002530))
 ```
 ```{r s1_m4_c_print, results = 'asis'}
-#s1_m4_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
+s1_m4_c %>% report_results_2(., list_of_samples[[1]], report_eta = T)
 ```

### 2019-08-27 data_analysis/data_analysis_expanded.Rmd
Slight change to m3s3

@@
 
+<!--
 ```{r s1_m1_d}
+# This seemed like a good idea until realizing how long it would take to do all the methods to this level of precision:
 
-set.seed(2339)
-s1_m1_d <- rre::minimum_subset_distance(list_of_samples[[1]],
-                                        partitions = 10,
-                                        lambda_vec = 700:740,
-                                        multiplier = 2,
-                                        c_seq_len = 11,
-                                        starts = data.frame(alpha = 0.28863,
-                                                            delta = 0.0032607))
+#set.seed(2339)
+#s1_m1_d <- rre::minimum_subset_distance(list_of_samples[[1]],
+#                                        partitions = 10,
+#                                        lambda_vec = 700:740,
+#                                        multiplier = 2,
+#                                        c_seq_len = 11,
+#                                        starts = data.frame(alpha = 0.28863,
+#                                                            delta = 0.0032607))
 ```
 ```{r s1_m1_d_print, results = 'asis'}
-s1_m1_d %>% report_results_2(.,  list_of_samples[[1]], report_eta = T)
+#s1_m1_d %>% report_results_2(.,  list_of_samples[[1]], report_eta = T)
 ```
 <br><br>
+-->
 
@@
 ```{r s3_m1_d}
+# Previous result on the edge of lambdagrid
 set.seed(452348)
@@
                                         partitions = 10,
-                                        lambda_vec = seq(500, 600 ,by=5),
+                                        lambda_vec = seq(450, 550 ,by=5),
                                         multiplier = 2,
@@
 sample_1 <- list(method_0 = s1_m0_b$best,
-                 method_1 = s1_m1_d$best,
+                 method_1 = s1_m1_c$best,
                  method_2 = s1_m2_c$best,

### 2019-08-28 data_analysis/data_analysis_expanded.Rmd
output edit

@@
 ---
-title: "Data Analysis Expanded"
+title: "Data Analysis"
 author: "Alex Paynter"
@@
 ---
+
+# Introduction
+
+This document shows the code used to analyze the Lake Champlain data originally published by _Tromas et.
+al.
+(2017)_.
+Our methods utilize a grid search for $\lambda$ and $C$, and some manual tuning of these grids was needed to analyze these data.
+We started with large, sparse grids and narrowed them down iteratively.
+The manual tuning is shown for each method and sample in chronological order, and we denote the order using letter suffixes.
+For example, `s1_m0_a` indicates the first run of sample 1 using method 0.
+Then `s1_m0_b` is the second run, using the `s1_m0_a` result to inform our grid choices.
+
+Our likelihood search was selected to be exhaustive rather than fast, so be warned that running all  code in this document will take days on a typical personal computer.
 
@@
 
-_Note_ The 'c' round of method 2 took over a day to run for these 3 samples.
-
 <br><br>
@@
                  method_2 = s1_m2_c$best,
-                 method_3 = s1_m3_d$best,
-                 method_4 = s1_m4_b$best)
+                 method_3 = s1_m3_e$best,
+                 method_4 = s1_m4_c$best)
 
@@
                  method_2 = s2_m2_c$best,
-                 method_3 = s2_m3_e$best,
+                 method_3 = s2_m3_f$best,
                  method_4 = s2_m4_c$best)
@@
                  method_2 = s3_m2_c$best,
-                 method_3 = s3_m3_e$best,
+                 method_3 = s3_m3_f$best,
                  method_4 = s3_m4_c$best)

### 2019-08-28 rre_manuscript/data_analysis_extended.tex

@@
 
+\noindent \textbf{Note from Alex:}\textit{ Included is my data analysis draft, a figure, a table and some blue comments which may or may not be useful in discussion.
+I did the table two ways - the same information is contained in both versions (The \texttt{mutlirow} package is needed for the first one.)  As always happy to edit text, figures or tables if that's helpful.}
+
 \section{Data Analysis}
@@
 
-The richness estimates from each method are displayed in Figure \ref{fig:data_analysis} and Table \ref{tab:data_analysis}.
-These estimates are high as 160 times the observed richness, suggesting that the negative binomial model may be a poor fit to this data set.
-As a result, applying our methods to real data required some manual tuning of $\lambdagrid$ and the search grid for $C$.
+The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.
+Method 0 produces the highest estimate for 2009 ($\widehat{C}_{[0]} = 73,404$) and 2010 ($\widehat{C}_{[0]}$ = 47,631) and it was second highest to Method 2 in 2011 ($\widehat{C}_{[2]} = 118,547$).
+Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness, $C_{[1]} = 584$ in 2009,  $572$ in 2010 and $718$ in 2011.
+Methods 3 and 4, which both use goodness of fit, were between these two extremes.
+These results mirror what we observed in simulations, where Methods 0 and 2 were the highest estimates and Method 1 was the lowest (see Figure \ref{fig:tuning_sim_1}).
 
-Despite the estimates being significantly more inflated than in our simulations, many of the trends observed in simulations hold true here.
-Method 0 produces the highest $\widehat{C}$ for 2009 and 2010 and it was second highest to Method 2 in 2011.
-Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness $c$.
-These are exactly the tendencies these methods showed when we generate simulations from a correctly specified model (see Figure \ref{fig:tuning_sim_1}).
+These estimates are up to 160 times the observed richness, suggesting that a majority of the species were unobserved in each sample.
+The relatively inflated estimates suggest that these data may be poorly fit by the negative binomial model.
+\\
 
-We model species richness because the observed richness is an underestimate of the truth.
-The original motivation for penalization is that the unpenalized MLE estimate can be much larger than the truth.
-In data analysis we don't know the truth, but we would expect an ideal method to be situated between these two extremes.
-Only the methods making use of a goodness of fit metric (3 and 4) are between Method 0 and the observed richness in all samples.
-As a result we conclude Methods 3 and 4 show the greatest promise on real data containing replicates.
+\noindent \textcolor{blue}{\textbf{Notes for discussion:}
+The reason we model species richness is that the observed richness is an underestimate of the actual richness.
+The motivation for penalization is that the unpenalized MLE tends to be much larger than the actual richness.
+These results show that only the methods making use of a goodness of fit metric (3 and 4) were consistently between Method 0 and the observed richness in all samples.
+We conclude Methods 3 and 4 show the greatest promise on real data containing replicates.}
 
-\noindent\large{\textbf{Note from Alex:  I've given two versions of the table, the first is much bigger and probably more suited to being a supplement.
-The second is more compact and could be crammed into a corner of the paper if needed.
-Happy to make revisions to tables and figures if that's helpful.
-}}
+\textcolor{blue}{
+In order to apply our methods to real data, some manual tuning of the optimization search grids ($\lambdagrid$ and the search grid for $C$) was required.}
+
+\normalsize
 
@@
 2009 & {[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 \\ 
-  \multirow{5}{*} & {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 \\ 
+  \multirow{5}{*} & {[1]} Minimum Variance & 584 & 700 & 0.28863 & 0.00326 \\ 
   \multirow{5}{*} & {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 \\ 
@@
 {[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 & 47,631 & \textemdash & 0.00185 & 0.00253 & 57,686 & \textemdash & 0.00161 & 0.00140 \\ 
-  {[1]} Minimum Variance & 593 & 700 & 0.27720 & 0.00342 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\ 
+  {[1]} Minimum Variance & 584 & 700 & 0.28863 & 0.00326 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\ 
   {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\ 
@@
 
-\bibliographystyle{biorefs}
-\bibliography{refs,amys-papers-library}
+%\bibliographystyle{biorefs}
+%\bibliography{refs,amys-papers-library}
 

### 2019-08-28 rre_manuscript/data_analysis_extended.tex
draft updated

@@
 \noindent \textbf{Note from Alex:}\textit{ Included is my data analysis draft, a figure, a table and some blue comments which may or may not be useful in discussion.
-I did the table two ways - the same information is contained in both versions (The \texttt{mutlirow} package is needed for the first one.)  As always happy to edit text, figures or tables if that's helpful.}
+I did the table two ways - the same information is contained in both versions (The \texttt{mutlirow} package is needed for the first one.)  I'm happy to edit text, figures or tables if that's helpful.}
 
@@
 \subsection{Estimating microbial richness in Lake Champlain}
-
+\normalsize
 To illustrate the performance of our methods on ecological data, we estimate strain-level microbial diversity in Lake Champlain, a large eutrophic lake in Canada.
@@
 
-The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.
-Method 0 produces the highest estimate for 2009 ($\widehat{C}_{[0]} = 73,404$) and 2010 ($\widehat{C}_{[0]}$ = 47,631) and it was second highest to Method 2 in 2011 ($\widehat{C}_{[2]} = 118,547$).
-Method 1 consistently estimated $\widehat{C}$ to be at or near the observed richness, $C_{[1]} = 584$ in 2009,  $572$ in 2010 and $718$ in 2011.
+The richness estimates from each method are displayed in Figure \ref{fig:data_analysis} and Table \ref{tab:data_analysis}.
+Method 0 produces the highest estimate for 2009 ($\widehat{C}_{[0]} = 73,404$) and 2010 ($\widehat{C}_{[0]}$ = 47,631) and it was second highest to Method 2 in 2011 ($\widehat{C}_{[0]} = 57,686$ and $\widehat{C}_{[2]} = 118,547$).
+Method 1 consistently estimated $C$ to be at or near the observed richness, $\widehat{C}_{[1]} = 584$ in 2009,  $572$ in 2010 and $718$ in 2011.
 Methods 3 and 4, which both use goodness of fit, were between these two extremes.
-These results mirror what we observed in simulations, where Methods 0 and 2 were the highest estimates and Method 1 was the lowest (see Figure \ref{fig:tuning_sim_1}).
+These results mirror what we observed in simulations, where Methods 0 and 2 generated the highest estimates and Method 1 was the lowest (see Figure \ref{fig:tuning_sim_1}).
 
-These estimates are up to 160 times the observed richness, suggesting that a majority of the species were unobserved in each sample.
-The relatively inflated estimates suggest that these data may be poorly fit by the negative binomial model.
+These estimates are up to 165 times the observed richness, suggesting that a majority of the species were unobserved in each sample.
+The relatively large estimates suggest that these data may be poorly fit by the negative binomial model.
 \\
@@
 \noindent \textcolor{blue}{\textbf{Notes for discussion:} 
-The reason we model species richness is that the observed richness is an underestimate of the actual richness.
-The motivation for penalization is that the unpenalized MLE tends to be much larger than the actual richness.
-These results show that only the methods making use of a goodness of fit metric (3 and 4) were consistently between Method 0 and the observed richness in all samples.
-We conclude Methods 3 and 4 show the greatest promise on real data containing replicates.}
-
-\textcolor{blue}{
-In order to apply our methods to real data, some manual tuning of the optimization search grids ($\lambdagrid$ and the search grid for $C$) was required.}
-
-\normalsize
+\begin{itemize}
+\item In order to apply our methods to real data, some manual tuning of the optimization search grids ($\lambdagrid$ and the search grid for $C$) was required.
+\item $\widetilde{\lambda}$ was similar for each method in 2010 and 2011 but larger in 2009, suggesting that $\widetilde{\lambda}$ may increase with $r$.
+\item The observed richness is an underestimate of the actual richness and the motivation for penalization is that the unpenalized MLE tends to overestimate the actual richness.
+Our data analysis shows that only the methods making use of a goodness of fit metric (3 and 4) were consistently between Method 0 and the observed richness in all samples.
+We conclude that Methods 3 and 4 show the greatest promise on real data containing replicates.
+\end{itemize}}
 
 \begin{table}[ht]
-\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
+\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) using our proposed methods. 
 \label{tab:data_analysis}}
@@
   \multirow{5}{*} & {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 \\ 
-  \multirow{5}{*} & {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 \\ 
+  \multirow{5}{*} & {[4]} CV G.O.F. & 36,893 & 85 & 0.00246 & 0.00251 \\ 
   2010 & {[0]} Unpenalized MLE & 47,631 & \textemdash & 0.00185 & 0.00253 \\ 
@@
   \multirow{5}{*} & {[2]} CV Likelihood & 24,799 & 0 & 0.00379 & 0.00270 \\ 
-  \multirow{5}{*} & {[3]} Goodness of fit & 13,156 & 220 & 0.00685 & 0.00258 \\ 
+  \multirow{5}{*} & {[3]} Goodness of fit & 13,156 & 225 & 0.00685 & 0.00257 \\ 
   \multirow{5}{*} & {[4]} CV G.O.F. & 47,098 & 0 & 0.00208 & 0.00277 \\ 
@@
 \begin{table}[ht]
-\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) from our proposed methods. 
+\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) using our proposed methods. 
 \label{tab:data_analysis_compact}}
@@
   {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\ 
-  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 220 & 0.00685 & 0.00258 & 40,040 & 230 & 0.00231 & 0.00137 \\ 
-  {[4]} CV G.O.F. & 39,997 & 100 & 0.00578 & 0.00253 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\ 
+  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 225 & 0.00685 & 0.00257 & 40,040 & 230 & 0.00231 & 0.00137 \\ 
+  {[4]} CV G.O.F. & 36,893 & 85 & 0.00246 & 0.00251 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\ 
    \bottomrule
@@
 \begin{figure}[t]
-\caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) based on our proposed methods.
+\caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) using our proposed methods.
 %$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.

### 2019-09-03 rre_manuscript/paper.tex
final proofreading and edits to data analysis

@@
 
+\usepackage{booktabs}
+\usepackage{multirow}
+
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 \newcommand{\lambdagrid}{\lambda^{\text{grid}}}
-
-\newtheoremstyle{break}
-  {\topsep}{\topsep}%
-  {\itshape}{}%
-  {\bfseries}{}%
-  {\newline}{}%
-\theoremstyle{break}
-\newtheorem{theorem}{Theorem}
-
-\newtheoremstyle{break}
-  {\topsep}{\topsep}%
-  {\itshape}{}%
-  {\bfseries}{}%
-  {\newline}{}%
-\theoremstyle{break}
-\newtheorem{prop}{Proposition}
-
-\newboolean{DEBUG}
-\setboolean{DEBUG}{true}
-\newrgbcolor{amycolor}{.7 .1 .1}
-\ifthenelse {\boolean{DEBUG}}
-{\newcommand{\amy}[1]{{\bf \amycolor{[\@Amy: #1]}}}}
-{\newcommand{\amy}[1]{}}
 
@@
 \newcommand{\blind}{1}
+% \newcommand{\blind}{0}
 
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{-.3in}%
+\addtolength{\oddsidemargin}{-.75in}%
+\addtolength{\evensidemargin}{-.75in}%
+\addtolength{\textwidth}{1.5in}%
+\addtolength{\textheight}{1.3in}%
 \addtolength{\topmargin}{-.8in}%
@@
 We consider the case where multiple frequency count tables have been collected from a homogeneous population, and investigate a penalized maximum likelihood estimator under a negative binomial model.
-  Because high probabilities of unobserved classes increase the variance of species richness estimates,
-  our method penalizes the probability of a class being unobserved.
+Because high probabilities of unobserved classes increase the variance of species richness estimates, our method penalizes the probability of a class being unobserved.
 Tuning the penalization parameter is challenging because the true species richness is never known, and so we propose and validate four novel methods for tuning the penalization parameter.
-We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years.
+We illustrate and contrast the performance of the proposed methods by estimating the strain-level microbial diversity of Lake Champlain over 3 consecutive years, and global human host-associated species-level microbial richness.
 \end{abstract}
@@
 \noindent%
-{\it Keywords: diversity, microbiome, regularization, maximum likelihood, ecology}
+{\it Keywords: diversity, regularization, maximum likelihood, ecology,  microbiome}
 \vfill
@@
 
-% TODO(Amy)
-% - go through all references and confirm used correctly
-
 \section{Introduction}
@@
 
-The \textit{species problem} concerns estimating the number of classes that are present in a population, which we call $C$.
+The \textit{species problem} concerns estimating $C$, the number of classes that are present in a population.
 $n$ individuals from the population can be sampled to find which classes they belong to, but since the  number of individuals in the sample is less than the total population size (which may be infinite), only $c$ classes are observed ($c \leq C$).
@@
 There exist both frequentist \citep{fisher_1943,chao_2002} and Bayesian \citep{efron_1976,barger_2010} approaches to parameter estimation under this model.
-We will focus on frequentist maximum likelihood in this paper.
+We will focus on frequentist maximum likelihood (penalized and unpenalized) in this paper.
 
@@
 The conditional and profile approaches have been shown to be asymptotically equivalent to the direct approach.
-%\amy{Need to justify why we are bringing this up}
 
@@
 A perennial issue in the species problem, especially for microbial datasets, is the instability of species richness estimators \citep{Rocchetti:2011tv,willis_2015}.
-%
-% Under the gamma-Poisson model,
-%
-% To illustrate, consider that a first-order approximation gives $$Var(\hat{C}_c | c) \approx \frac{c^2}{(1-\hat{p}(0))^4}Var(\hat{p}(0))$$ \amy{check this}
+% To illustrate, consider that a first-order approximation gives $$Var(\hat{C}_c | c) \approx \frac{c^2}{(1-\hat{p}(0))^4}Var(\hat{p}(0))$$ TODO check this
 % Therefore,
@@
 A proposal which encourages stability of estimates under Poisson-mixture models is due to \citet{wang_2005}.
-Their proposal is to add a penalty term to the log likelihood that penalizes the probability of observing a species zero times.
+Their proposal is to add a penalty term to the log-likelihood that penalizes the probability of observing a species zero times.
 They consider the
-penalized log likelihood
+penalized log-likelihood
 \begin{equation}
@@
 \end{equation}
-where $\ell(C,\eta) = \log L(C, \eta)$ is the log likelihood and $\lambda > 0$ is a penalization parameter.
+where $\ell(C,\eta) = \log L(C, \eta)$ is the log-likelihood and $\lambda > 0$ is a penalization parameter.
 Let $\hat{C}_\lambda = \arg \max_{C} \ell_\lambda(C, \eta)$.
@@
 
-\section{Extension to Replicate Data}
+\section{Extension to Biological Replicates}
 \label{sec:fixed_lambda}
@@
 $J$ indicates the data being used in the evaluation of the objective function, with $J = \{1, \dots r\}$ meaning we use all replicates.
-Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log likelihood for replicate $j$.
-Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log likelihoods for each individual replicate.
-We define the \textit{penalized log likelihood} for multiple samples as
+Let $\ell \left(C, \eta; \{f_{kj}\}_{k \geq 1} \right)$ be the unpenalized log-likelihood for replicate $j$.
+Using the fact that the draws are independent, our objective function for a fixed $\lambda \geq 0$ is the sum of the log-likelihoods for each individual replicate.
+We define the \textit{penalized log-likelihood} for multiple samples as
 \begin{align}
@@
 \end{align}
-By fixing $\lambda = 0$ we obtain the unpenalized log likelihood for the set $J$.
+By fixing $\lambda = 0$ we obtain the unpenalized log-likelihood for the set $J$.
 We define $\widehat{C}_{\lambda}$ to be the penalized maximum likelihood estimate of $C$ based on tuning parameter $\lambda$, that is,
@@
 \label{sec:efficacy_sims}
-\subsection{Simulation setup}
+% \subsection{Simulation setup}
 
@@
 For completeness we also investigated nearby $\eta$ values: $\eta = \left( 10^{-1}, 10^{-3} \right)$ and $\eta = \left( 10^{-1}, 10^{-5}\right)$.
-% To select $\eta$, we investigated the unregularised
+% To select $\eta$, we investigated the unregularized
 %  two microbial datasets
@@
 
-\amy{check all captions}
-
-\begin{table}[h]
-\caption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
+\begin{table}[t]
+\caption{The penalized maximum likelihood estimate of $C$ has lower RMSE for all investigated choices of $\eta$ and $r$ under a zero-truncated Gamma-mixed Poisson model for species abundances based on 100 simulations for each choice of $\eta$ and $r$.
 $\lambda_{\text{opt}}$ is the value of $\lambda$ which produced the lowest RMSE.
@@
 
-\subsection{Simulation results} \label{sec:first_sim}
+% \subsection{Simulation results} \label{sec:first_sim}
 
@@
 \begin{figure}[p]
-\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-1}, 10^{-1})$, over 100 simulations. The black line indicates the true value $C = 1000$.
+\caption{Estimates of $C$ and their root-MSE over $\lambda$ when $\eta = (10^{-1}, 10^{-1})$ and $C = 1000$. Results are based on 100 simulations per $\lambda$.
 \label{fig:fixed_lambda}}
@@
 \begin{figure}[p]
-\caption{Simulation results for estimation of $C$ under a range of fixed $\lambda$ constants, $\eta = (10^{-2}, 10^{-5})$, over 100 simulations.  The black line indicates the true value $C = 1000$.
+\caption{Estimates of $C$ and their root-MSE over $\lambda$ when $\eta = (10^{-2}, 10^{-5})$ and $C = 1000$. Results are based on 100 simulations per $\lambda$.
 \label{fig:fixed_lambda_2}}
@@
 
-%
-%
-%
 \section{Methods for Tuning $\lambda$}
@@
 \setcounter{subsection}{-1}
-\subsection{Method 0: MLE without penalization}
+\subsection{Method 0: No penalization}
 
@@
 
-For all of the remaining methods (1-4) we generate estimates $\widehat{C}_\lambda$ over $\lambda \in \lambda^{\text{grid}}$, where $\lambdagrid$ is user-specified (e.g., $\lambdagrid = \{0, 5, 10, \dots 120\}$ in Section \ref{sec:first_sim}).
+For all of the remaining methods (1-4) we generate estimates $\widehat{C}_\lambda$ over $\lambda \in \lambda^{\text{grid}}$, where $\lambdagrid$ is user-specified (e.g., $\lambdagrid = \{0, 5, 10, \dots 120\}$ in Section \ref{sec:efficacy_sims}).
 %The methods differ in how they subset the data to create these estimates, and in the steps they take to arrive at a final estimate $\widehat{C}_{\text{[method]}}$.
@@
 
-% \amy{up to here with editing}
-
 In Method 3 we make use of this goodness of fit metric by first generating estimates using all replicates.  For each $\lambda \in \lambdagrid$ let
@@
 \subsection{Method 4: Cross-validated goodness of fit}
-
-\amy{up to here}
 
@@
 
-\section{Tuning Method Simulations}
+\section{Comparison of methods for selecting $\lambda$}
 \label{sec:tuning_simulations}
 
-We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the performance of each estimator via simulations.
+We have proposed four tuning methods motivated by properties which would be desirable in an estimator of $C$.  In this chapter we compare the performance of each estimator.
 We simulate zero-truncated gamma-mixed Poisson data using the same parameters used in Section \ref{sec:efficacy_sims}.
@@
 In each simulation below, we select the largest value in $\lambdagrid$ by doubling the optimal value of $\lambda$ found in Table \ref{tab:fixed_lambda_results}.
-
-\amy{make sure to say zero truncated negative binomial throughout}
 
@@
 
-\subsection{Comparison of methods' performance}
+\subsection{Initial comparison of methods' performance}
 \label{sec:tuning_simulation_1}
@@
 We conjecture that the differing performance across $\eta$ is due to the differing rare species structures implied by the different $\eta$'s.
-For example $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbbm{E}(k_{\text{max}})$, but relatively few rare species (see Table \ref{tab:eta_intuition}).
+For example $\eta = (10^{-2},10^{-5})$ has more abundant species and larger expected maximum abundance $\mathbbm{E}(k_{\text{max}})$, but relatively few rare species (Table \ref{tab:eta_intuition}).
 
@@
 
-\begin{table}[ht]
+\begin{table}[t]
 \caption{RMSE$(\widehat{C})$ for Methods 0-4 based on a zero-truncated gamma-mixed Poisson data generating process.
@@
 \begin{figure}[t]
-\caption{Simulation results for all proposed tuning methods over $\eta = (10^{-1}, 10^{-1})$ and $\eta = (10^{-2}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
+\caption{Simulation results for all proposed methods when $\eta = (10^{-1}, 10^{-1})$, and when $\eta = (10^{-2}, 10^{-5})$.
 \label{fig:tuning_sim_1}}
@@
 This is similarly the case for Method 0, though we see that Method 2 has slightly lower bias compared to Method 0.
-%Under this $\eta$ value Method 2 does better than Method 0, but the estimates are almost all too high.
 Even for $\eta = (10^{-1}, 10^{-1})$, where the Method 0 bias is smaller, we see that Method 2 has poor performance (Table \ref{tab:tuning_sim_1}).
@@
 Methods 3 and 4, which are both based on goodness of fit criteria, display different patterns depending on the $\eta$ and $r$ values of the simulation.
-% This is a desirable property because we know the optimal $\lambda$ value depends on the structure of the data, and we want a method which can adapt.
-In Methods 3 and 4 we generate estimates using the penalized likelihood and evaluate the fit with the goodness of fit metric.
-While we find the indications that Method 4 is adaptable to the dataset encouraging, under $\eta = (10^{-1}, 10^{-1})$ the RMSE for Method 4 is 15-75\% greater than for Method 0.
-In comparison, Method 3 is at least as good as Method 0 with respect to RMSE.
+Method 4 outperforms Method 0 when $\eta = (10^{-2}, 10^{-5})$ (Table \ref{tab:tuning_sim_1}), but under $\eta = (10^{-1}, 10^{-1})$ the RMSE for Method 4 is 15-75\% greater than for Method 0.
+In comparison, Method 3 is at least as good as Method 0 with respect to RMSE for all parameter combinations tested.
 In addition,  Method 3 is simpler and faster than Method 4.
-Given deficits noted for Method 1 and 2, Method 3 is by far the most promising method based on this simulation.
-We propose to conduct a secondary simulation to further challenge Method 3, testing whether it remains superior to Method 0 over a wider range of $\eta$ values.
-
-\amy{move later -- Incorporating two distinct functions may explain why these methods have more adaptable behavior, as opposed to Method 2 which uses likelihood exclusively.}
+We conclude that Method 3 is the most promising method based on this simulation.
+We propose to conduct a secondary simulation to further investigate Method 3, testing whether it remains superior to Method 0 over a wider range of $\eta$ values.
 
@@
 Method 3 has slightly worse performance on all $r$ values for $\eta = (10^{-1}, 10^{-3})$, with a RMSE which was 14-36\% greater than Method 0.
-This $\eta$ generates more rare species, so the comparatively poor performance of Method 3 is consistent with the results our previous simulation.
-In Section \ref{sec:tuning_simulation_1}, Method 3 had greatly improved estimation in data with many abundant species, but the improvements over Method 0 were weaker in datasets that had many rare species.
+$\eta = (10^{-1}, 10^{-3})$ generates more rare species than $\eta = (10^{-1}, 10^{-5})$
+(assessed as $\sum_{k=0}^3 p_\eta(k)$), so the comparatively poor performance of Method 3 is consistent with the results our previous simulation: in Section \ref{sec:tuning_simulation_1}, we found that Method 3 improved estimation in data with abundant species ($\eta = (10^{-2}, 10^{-5})$),
+but the improvements over Method 0 were less pronounced in datasets that had more rare species ($\eta = (10^{-1}, 10^{-1})$).
 
@@
 For $\eta = (10^{-1}, 10^{-5})$ we note that $\widehat{C}_{[0]}$ is very large for a few simulations.
-Referencing Table \ref{tab:eta_intuition} we can recall this $\eta$ choice has the highest proportion of abundant species.
-Therefore, this is a excellent example of a simulation structure which will cause the instability problem in maximum likelihood estimation discussed in Section \ref{sec:literature_review}.
-Method 3 succeeds on this $\eta$ by selecting lower estimates.
+We recall from Table \ref{tab:eta_intuition} that this $\eta$ choice has the highest proportion of abundant species.
+This is an example of a simulation structure which will cause the instability problem in maximum likelihood estimation discussed in Section \ref{sec:literature_review}.
+Method 3 outperforms Method 0 on this $\eta$ by selecting lower estimates.
 
 \begin{figure}[t]
-\caption{Simulation results for Methods 0 and 3 over $\eta = (10^{-1}, 10^{-3})$ and $\eta = (10^{-1}, 10^{-5})$ are shown, with a black line indicating the true value, $C = 1000$.
+\caption{Simulation results for Methods 0 and 3 when $\eta = (10^{-1}, 10^{-3})$, and when and $\eta = (10^{-1}, 10^{-5})$.
 \label{fig:tuning_sim_2}}
@@
 This gives us 8 replicates from 2009, 6 replicates from 2010 and 6 replicates from 2011.
-%The maximum observed richness in any given sample was 560 in 2009, 572 in 2010 and 715 in 2011.
+Given our results from \ref{sec:tuning_simulations}, we focus on Methods 0 and 3.
 
-% These extreme estimates suggest that the Gamma-Poisson model which underlies all methods is a poor fit to these data.
-% In simulations penalization had the intended effect of lowering inflated estimates.
-However, we always simulated under a Gamma-Poisson model in simulations.
-Using wild data our methods still produce estimates which are suspiciously high.
+Method 3 produces lower estimates of $C$ than Method 0, as expected.
+The 2009 and 2010 estimates were approximately $3.5$ times lower for Method 3, while the 2011 estimate was approximately $1.5$ times lower for Method 3.
+We see that the estimates of $\delta$ are comparable across the two methods, but that the estimates of $\alpha$ differ, and may be higher or lower depending on the dataset.
 
-% \begin{table}[t]
-% \caption{Estimates using a series of three samples from 2009, 2010 and 2011 in the lakes data.
-$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
-The number of replicates in each sample is given by $r$.
-% \label{tab:data_analysis}}
-% \centering
-% \begin{tabular}{|r|r|r|r|}
-%   \hline
-% & 2009  & 2010 & 2011 \\
-%   \hline
-% $\widehat{C}_{[0]}$ & 11200 & 11440 & 14014 \\
-%   $\widehat{C}_{[1]}$ & 9645 & 11139 & 12246 \\
-%   $\widehat{C}_{[2]}$ & 10612 & 11120 & 11946 \\
-%   $\widehat{C}_{[3]}$ & 11200 & 11440 & 14014 \\
-%   $\widehat{C}_{[4]}$ & 10744 & 11165 & 10328 \\
-%    \hline
-%      $r$ & 8 & 6 & 6 \\
-%   \hline
-% \end{tabular}
-% \end{table}
+% The richness estimates from each method are displayed in Figure \ref{fig:data_analysis} and Table \ref{tab:data_analysis_compact}.
+Method 0 produces the highest estimate for 2009 ($\widehat{C}_{[0]} = 73,404$) and 2010 ($\widehat{C}_{[0]}$ = 47,631) and it was second highest to Method 2 in 2011
+% ($\widehat{C}_{[0]} = 57,686$ and $\widehat{C}_{[2]} = 118,547$).
+Method 1 consistently estimated $C$ to be at or near the observed richness, $\widehat{C}_{[1]} = 584$ in 2009,  $572$ in 2010 and $718$ in 2011.
+Methods 3 and 4, which both use goodness of fit, were between these two extremes.
+These results mirror what we observed in simulations, where Methods 0 and 2 generated the highest estimates and Method 1 was the lowest (see Figure \ref{fig:tuning_sim_1}).
+%
+% These estimates are up to 165 times the observed richness, suggesting that a majority of the species were unobserved in each sample.
+The relatively large estimates suggest that these data may be poorly fit by the negative binomial model.
 
-The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.
+\begin{table}[t]
+\caption{Diversity estimates from the Lake Champlain data analysis from 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) using our proposed methods.
+\label{tab:data_analysis_compact}}
+\centering
+\tiny
+\begin{tabular}{lllllllllllll}
+  \toprule
+      & \multicolumn{4}{c}{2009} & \multicolumn{4}{c}{2010} & \multicolumn{4}{c}{2011} \\ \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(lr){10-13}
+Method & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ & $\widehat{C}$ & $\widetilde{\lambda}$ & $\widehat{\alpha}$ & $\widehat{\delta}$ \\
+  \midrule
+{[0]} Unpenalized MLE & 73,404 & \textemdash & 0.00088 & 0.00180 & 47,631 & \textemdash & 0.00185 & 0.00253 & 57,686 & \textemdash & 0.00161 & 0.00140 \\
+  % {[1]} Minimum Variance & 584 & 700 & 0.28863 & 0.00326 & 572 & 500 & 0.62668 & 0.00724 & 718 & 500 & 0.40112 & 0.00355 \\
+  % {[2]} CV Likelihood & 25,930 & 220 & 0.00516 & 0.00142 & 24,799 & 0 & 0.00379 & 0.00270 & 118,547 & 0 & 0.00122 & 0.00193 \\
+  {[3]} Goodness of fit & 20,160 & 550 & 0.00323 & 0.00174 & 13,156 & 225 & 0.00685 & 0.00257 & 40,040 & 230 & 0.00231 & 0.00137 \\
+  % {[4]} CV G.O.F. & 36,893 & 85 & 0.00246 & 0.00251 & 47,098 & 0 & 0.00208 & 0.00277 & 36,395 & 5 & 0.00358 & 0.00178 \\
+   \bottomrule
+\end{tabular}
+\end{table}
+
+% \begin{figure}[t]
+% \caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) using our proposed methods.
+% \label{fig:data_analysis}}
+% \centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/data_analysis_expanded.pdf}}
+% \end{figure}
+
+% The richness estimates from each method are displayed in Figure \ref{fig:data_analysis}.
 We see that Methods 0 and 3 produce identical estimates for all samples, and these estimates are larger than estimates for Methods 1, 2 and 4.
@@
 
-Our estimates from each method are all at least 10 times as high as the observed richness, indicating that a significant fraction of species were unobserved.
+% Our estimates from each method are all at least 10 times as high as the observed richness, indicating that a significant fraction of species were unobserved.
 Based on our previous results, estimation error is lowest for Methods 0 and 3 when there are many rare species.
@@
 
-\begin{figure}[t]
-\caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) based on our proposed methods.
-%$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
+% \begin{figure}[t]
+% \caption{Estimates of strain-level microbial diversity in Lake Champlain in the summers of 2009 ($r = 8$), 2010 ($r = 6$) and 2011 ($r = 6$) based on our proposed methods.
+% %$\widehat{C}_{[\text{method}]}$ is the $C$ estimate obtained using the methods defined in Section \ref{sec:tuning_proposals}, shown for each sample.
 The number of replicates in each sample is given by $r$.
+% %
+% % A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.
+% \label{fig:data_analysis}}
+% \centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/data_analysis.pdf}}
+% \end{figure}
 %
-% A visual display of the results from Table \ref{tab:data_analysis}, $\widehat{C}$ estimates from each method.
-\label{fig:data_analysis}}
-\centering\makebox[\textwidth]{\includegraphics[width=\textwidth]{./images/data_analysis.pdf}}
-\end{figure}
-
 % \noindent \alex{Next: Modify the discussion to harmonious with what you've written here.}
@@
 Therefore, only Methods 0 and Method 3 can be applied.
-We found that $\widehat{C}_{[0]} = 420056$ with $(\hat{\alpha}, \hat{\delta}) = (0.00234, 0.00641).$ In contrast, $\widehat{C}_{[3]} = 163587$ with $\tilde{\lambda}_{[3]} = 905$ and $(\hat{\alpha}, \hat{\delta}) = (0.006045614, 0.006352191).$
+We found that $\widehat{C}_{[0]} = 420,056$ with $(\hat{\alpha}, \hat{\delta}) = (0.00234, 0.00641).$ In contrast, $\widehat{C}_{[3]} = 163,587$ with $\tilde{\lambda}_{[3]} = 905$ and  $(\hat{\alpha}, \hat{\delta}) = (0.00605, 0.00635).$
+% $(\hat{\alpha}, \hat{\delta}) = (0.006045614, 0.006352191).$
 % With $\lambdagrid = \{0, 0.25, \ldots, 1100\}$ we found that Method 3 selected $\lambda = 913.5$. However, $\widehat{C}_{[3]} = 98600$
 % We found that $\widehat{C}_{[3]} = 163587$ with $\tilde{\lambda}_{[3]} = 905$ and $(\hat{\alpha}, \hat{\delta}) = (0.006045614, 0.006352191).$ % GOF = 11196564679
+We therefore find Method 3 to produce an estimate approximately $2.5$ times lower than the estimate produced by Method 0.
+Similar to our analysis of the \cite{tromas_2017} dataset, we find comparable $\hat{\delta}$'s across the two methods but different $\hat{\alpha}$'s.
 
-\amy{TODO: Discussion}
+\section{Discussion}
 
-\section{Discussion and Conclusion}
 \label{sec:discussion}
-
-\subsection{Limitations and future work}
-
-The approach of \citet{wang_2005} is considerably more general than the gamma-Poisson model.
-The gamma-Poisson model is common for modeling microbiome data \citep{deseq2,mcmurdie2013phyloseq}, and for this reason we focused on it in our investigation.
-However, the best performing method, Method 3, is also amenable to other modeling choices provided that there is some way to calculate the fitted counts $\widehat{f_k}$.
-We leave investigation of tuning parameter selection under different models to future work.
-
-Our results from Section \ref{sec:fixed_lambda} hint at a possible positive correlation between $r$ and the optimal choice of $\lambda$.
-We investigated whether the optimal choice of $\lambda$ remained constant for penalties of the form $-h(r)\lambda \log p_{\eta}(0)$ (instead of $-\lambda \log p_{\eta}(0)$; see Equation \eqref{eq:objective}).
-We tested $h(r) = r$ and $h(r) = \sqrt{r}$, but found that the trend in optimal $\lambda$ is not so simple.
-We leave further investigation into how the optimal $\lambda$ varies with $r$ to future work.
-Understanding of $h(r)$ would allow us to consider unequally sized evaluation and training partitions for Methods 2 and 4.
-
-For our investigations, we intentionally chose a likelihood optimization algorithm which was stable and exhaustive.
-We also did not construct standard errors for our estimates, and the long computation times precluded the consideration of a bootstrapping approach.
-Refining the optimization algorithm would be a valuable extension, and a faster optimization algorithm would facilitate a resample-based variance estimation procedure.
 
@@
 
-In this paper we discussed an extension of the penalized maximum likelihood procedure of \citet{wang_2005} for species richness estimation to data with biological replicates, and proposed several methods for tuning the penalization parameter.
-We demonstrated that penalization can reduce estimation error (defined as square root mean squared error) when analyzing replicate data.
-We found that tuning the penalization parameter is challenging, but that a goodness of fit tuning method (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
-However, on a microbial abundance dataset with many rare species, it selected the same estimate as unregularized maximum likelihood estimation.
+In this paper we outlined an extension of the penalized maximum likelihood procedure of \citet{wang_2005} for species richness estimation to data with biological replicates, and proposed several methods for tuning the penalization parameter.
+We demonstrated that penalization can reduce estimation error when analyzing replicate data.
+We found that tuning the penalization parameter is challenging, but that a tuning method based on goodness of fit (Method 3) has better performance than the unpenalized MLE in 8 of the 12 parameter combinations we analyzed.
+On two datasets we found that it reduces the magnitude of species richness estimates.
+Since species richness estimates can be unstable, we find the reduction in estimates appealing.
+While we cannot conclude that our proposed goodness of fit tuning parameter selection method provides more reliable estimates than unregularized estimation, the performance of the goodness of fit approach on simulated data is encouraging.
 
@@
 with a parametric model for species abundance data we can use goodness of fit in conjunction with maximum likelihood to select $\lambda$.
-While simulation results suggest that regularisation can improve estimation, in practice we found no difference between regularised and unregularised estimates.
+We conjecture that the goodness of fit method performs well because it employs a combination of likelihood and goodness of fit metrics calculated on the full dataset to select the tuning parameter, unlike methods that only rely on likelihood (e.g., Method 2) or split the data (e.g., Method 4).
 
@@
 
+\subsection{Limitations and future work}
+
+The approach of \citet{wang_2005} is considerably more general than the
+-Poisson model.
+The gamma-Poisson model is common for modeling microbiome data \citep{deseq2,holmes2018modern}, and for this reason we focused on it in our investigation.
+However, our goodness of fit method is also amenable to other parametric models.
+We leave investigation of tuning parameter selection under different models to future work.
+
+Our results from Section \ref{sec:fixed_lambda} hint at a possible positive correlation between $r$ and the optimal choice of $\lambda$.
+We investigated whether the optimal choice of $\lambda$ remained constant for penalties of the form $-h(r)\lambda \log p_{\eta}(0)$ (instead of $-\lambda \log p_{\eta}(0)$; see Equation \eqref{eq:objective}).
+We tested $h(r) = r$ and $h(r) = \sqrt{r}$, but found that the trend in optimal $\lambda$ is not so simple.
+We leave further investigation into how the optimal $\lambda$ varies with $r$ to future work.
+Understanding of $h(r)$ would allow us to consider unequally sized evaluation and training partitions for Methods 2 and 4.
+
+For our investigations, we intentionally chose a likelihood optimization algorithm which was stable and exhaustive.
+We also did not construct standard errors for our estimates, and the long computation times precluded the consideration of a bootstrapping approach.
+Refining the optimization algorithm would be a valuable extension, and a faster optimization algorithm would facilitate a resample-based variance estimation procedure.
+
 \section{Code References}
-\label{sec:code}
-Our method implementations including our core objective optimization algorithm can be found at \texttt{github.com/statdivlab/rre}.
-The code to reproduce our figures and simulations can be found at \texttt{github.com/statdivlab/rre\_sims}.
-All code is written in \texttt{R} (\citet{r_project}).
+
+An \texttt{R} package implementing our methods is available at \texttt{github.com/statdivlab/rre}.
+Code to reproduce our figures and simulations can be found at \texttt{github.com/statdivlab/rre\_sims}.
+We are also grateful to the \cite{r_project} and authors of the packages \texttt{tidyverse} \citep{tidyverse}, \texttt{magrittr} \citep{magrittr}, \texttt{breakaway} \citep{breakawaypkg}, \texttt{foreach} \citep{foreach}, \texttt{Rcpp} \citep{Rcpp} and \texttt{data.table} \citep{datatable}, which were used for constructing the figures and running the analyses in this paper.
 
@@
 The authors of this manuscript are grateful to Jim Hughes for many helpful suggestions that improved content and exposition.
+This work was supported in part by the National Institute of General Medical Sciences (NIGMS) of the NIH under Grant R35 GM133420.
 
-{\it Conflict of Interest}: None to declare
+{\it Conflicts of Interest}: None to declare.
 
 \bibliographystyle{biorefs}
+% \bibliographystyle{natbib}
+% \bibliographystyle{Chicago}
 \bibliography{refs,amys-papers-library}

### 2019-09-03 data_analysis/pasolli/pasolli_investigation.R
add breakaway dependence

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(breakaway)
 pasolli_et_al <- openxlsx::read.xlsx("https://ars.els-cdn.com/content/image/1-s2.0-S0092867419300017-mmc4.xlsx", sheet = 2)
 ft <- pasolli_et_al$`#.Samples` %>% make_frequency_count_table
 ft %>% head
@@ -161,7 +162,7 @@ m3_n <- rre::gof_criterion(list(ft),
                            c_seq_len = 100,
                            multiplier = 60,
                            lambda_vec = 900:910)
-m3_n
+m3_n # lambda = 905
 163587/3930 # 41, ok
 
 m3_n$best
